### 2014-09-15 currentresults.tex
added currentresults tex file

\documentclass[11pt]{article}
\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS
\topmargin -1.5cm        % read Lamport p.163
\oddsidemargin -0.04cm   % read Lamport p.163
\evensidemargin -0.04cm  % same as oddsidemargin but for left-hand pages
\textwidth 16.59cm
\textheight 21.94cm 
%\pagestyle{empty}       % Uncomment if don't want page numbers
\parskip 7.2pt           % sets spacing between paragraphs
%\renewcommand{\baselinestretch}{1.5} % Uncomment for 1.5 spacing between lines
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{verbatim}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\parindent 0pt		 % sets leading space for paragraphs
\author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
\title{18.821 - Poker Chips}

\begin{document}         
\maketitle

\begin{theorem}
Here is a theorem
\end{theorem}
\begin{proof}
it is true
\end{proof}
\end{document}

### 2014-09-15 currentresults.tex
modified the results doc

@@
 \newtheorem{lemma}{Lemma}
+\newtheorem{definition}{Definition}
 \parindent 0pt		 % sets leading space for paragraphs
@@
 
+\begin{definition}
+Let a \textbf{valid} earthquake be an earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
+\end{definition}
 \begin{theorem}
-Here is a theorem
+For any board with finitely many chips that stabilizes after finitely many earthquakes, the order in which the earthquakes occur does not matter.
 \end{theorem}
 \begin{proof}
-it is true
+The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final result.
+This technique will then uniquely (up to a point) specify an order in which the earthquakes will occur, which completes the proof.
+
+First, we need a lemma in order to do the rearrangements.
+
+\begin{lemma}
+Let $E_1,E_2,E_3,\dots, E_k$ be a sequence of earthquakes, where each $E_i$ specifies the location of a square hit.
+Then the sequence given by moving some earthquake $E_j$ forward in the sequence to a new spot where $E_j$ is still a valid earthquake gives the same final configuration.
+\end{lemma}
+\begin{proof}
+Will need to formalize this, but this is true because addition and subtraction are commutative, and that moving $E_j$ earlier up in the sequence can't hurt another other squares' ability to have an earthquake.
+\end{proof}
+
+Now, given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
+
 \end{proof}

### 2014-09-15 FiniteBoards.tex
adding proof for finite boards, and images from simulations

\documentclass[11pt]{article}
\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS
\topmargin -1.5cm        % read Lamport p.163
\oddsidemargin -0.04cm   % read Lamport p.163
\evensidemargin -0.04cm  % same as oddsidemargin but for left-hand pages
\textwidth 16.59cm
\textheight 21.94cm 
%\pagestyle{empty}       % Uncomment if don't want page numbers
\parskip 7.2pt           % sets spacing between paragraphs
%\renewcommand{\baselinestretch}{1.5} % Uncomment for 1.5 spacing between lines
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{verbatim}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\parindent 0pt		 % sets leading space for paragraphs
\author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
\title{18.821 - Poker Chips Finite board configurations}

\begin{document}         
\maketitle

In this file, we analyze the case where the initial number of chips is finite. That is, the sum of all the chips on the board is finite.

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ such that after $N$ earthquakes, the system will be in a stable state (any subsequence earthquake does not cause any chips to move). 
\end{theorem}

The proof will be as follows: first, we will show that finite boards cannot return to a previous state. Let $B$ be the set of all possible board configurations, that is, $B = \{ f: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^{\geq 0}\}$. Each board configuration is given by a function of the integer lattice to the nonnegative integers. So $f(x,y)$ is the number of chips in the grid location $x,y$.

The earthquake is defined as $E: B \rightarrow B$ acts in the following manner:
\[ E(f(x,y)) = f(x,y) + \sum_{l \in n(x,y)} [[ f(l) \geq 4 ]] - 4[[f(x,y) \geq 4]] \]
Where $n(x,y)$ is the set of neighbors of $x,y$, and $[[ X ]] = 1$ if $X$ happened and $0$ otherwise. Let $B_f \subset B$ with all boards with a finite number of chips. 
\[ B_f = \bigcup_{n \geq 0} \{ f \in B | \sum_{x,y} f(x,y) = n \} \]

So a restatement of the theorem is to say that for all $f \in B$ where $\sum_{x,y} f(x,y) = S$, there exists an $N$ such that for all $m \in \mathbb{Z}$, $E^N(f) = E^{N+m}(f)$.

\begin{lemma}
If $E(f) \neq f$, then $E^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: B_f \rightarrow \mathbb{Z}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. I will show that 
\[ E(f) \neq f \Rightarrow \Phi(E(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(E(f)) - \Phi(f)$.
\[ \Phi(E(f)) = \sum_{x,y} E(f(x,y)) (x+y)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (x+y)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in of location $(x,y)$ of $4$, there is an increase in two neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(E(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
\label{finiteextension}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$. Then $E^n(f)$ only contains chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
\end{lemma}

\begin{proof}
For each chip that moves in one direction, there is another chip that moves in the opposite direction. Therefore, at each time, the number of chips that extend the range can be at most half of what extended before. This means that there are at most $\log S$ extensions on each side. 
\end{proof}

So now we are ready to prove the theorem that all finite boards will reach a stable state.

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in B$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$. Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.

There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in B$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \geq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}

\end{document}

### 2014-09-15 simulate.py

@@ -43,6 +43,10 @@ def getBoundary(board):
     highestY = max(allY)
     return (lowestX, highestX, lowestY, highestY)
 
+def getStableBoundry(board):
+    actUntilStable(board)
+    return getBoundary(board)
+
 def printBoundary(board):
     (lowestX, highestX, lowestY, highestY) = getBoundary(board)
     print "Boundary: X:[%d, %d] \tY:[%d, %d]" % (lowestX, highestX, lowestY, lowestY)
@@ -62,7 +66,7 @@ def actUntilStable(board):
     count = 0
     while act(board):
         count += 1
-    print "Done: took %d actions" % count
+    return count
 
 def actAndSee(board):
     while 1:

### 2014-09-15 InfiniteBoards.tex
added infinite boards file

--- 2014-09-15 FiniteBoards.tex
@@
 \usepackage{verbatim}
+\usepackage{graphicx}
 \newtheorem{theorem}{Theorem}
@@
 \newtheorem{definition}{Definition}
+\newtheorem{observation}{Observation}
+
 \parindent 0pt		 % sets leading space for paragraphs
 \author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
-\title{18.821 - Poker Chips Finite board configurations}
+\title{18.821 - Poker Chips infinite board configurations}
 
@@
 \maketitle
-
-In this file, we analyze the case where the initial number of chips is finite. That is, the sum of all the chips on the board is finite.
-
-\begin{theorem}
-\label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ such that after $N$ earthquakes, the system will be in a stable state (any subsequence earthquake does not cause any chips to move).
-\end{theorem}
 
@@
 \[ E(f(x,y)) = f(x,y) + \sum_{l \in n(x,y)} [[ f(l) \geq 4 ]] - 4[[f(x,y) \geq 4]] \]
-Where $n(x,y)$ is the set of neighbors of $x,y$, and $[[ X ]] = 1$ if $X$ happened and $0$ otherwise. Let $B_f \subset B$ with all boards with a finite number of chips. 
-\[ B_f = \bigcup_{n \geq 0} \{ f \in B | \sum_{x,y} f(x,y) = n \} \]
+Where $n(x,y)$ is the set of neighbors of $x,y$, and $[[ X ]] = 1$ if $X$ happened and $0$ otherwise.
 
-So a restatement of the theorem is to say that for all $f \in B$ where $\sum_{x,y} f(x,y) = S$, there exists an $N$ such that for all $m \in \mathbb{Z}$, $E^N(f) = E^{N+m}(f)$.
-
-\begin{lemma}
-If $E(f) \neq f$, then $E^n(f) \neq f$ for all $n > 0$.
-\end{lemma}
-
-\begin{proof}
-Let $\Phi: B_f \rightarrow \mathbb{Z}$ be the following function
-\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
-Note that this is well-defined since the boards contain a finite number of chips. I will show that 
-\[ E(f) \neq f \Rightarrow \Phi(E(f)) > \Phi(f) \]
-This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
-
-We can evaluate $\Phi(E(f)) - \Phi(f)$.
-\[ \Phi(E(f)) = \sum_{x,y} E(f(x,y)) (x+y)^2 \]
-\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (x+y)^2 \]
-Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$.
-For each decrease in of location $(x,y)$ of $4$, there is an increase in two neighbors by $1$.
-
-Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
-So
-\[ \Phi(E(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
-\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
-									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
-									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
-\[ \geq 4 > 0 \] 
-\end{proof}
-
-\begin{lemma}
-\label{finiteextension}
-Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $E^n(f)$ only contains chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
-\end{lemma}
-
-\begin{proof}
-For each chip that moves in one direction, there is another chip that moves in the opposite direction.
-Therefore, at each time, the number of chips that extend the range can be at most half of what extended before.
-This means that there are at most $\log S$ extensions on each side.
-\end{proof}
-
-So now we are ready to prove the theorem that all finite boards will reach a stable state.
-
-\begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $f \in B$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
-Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.
-
-There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in B$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \geq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
-\end{proof}
+\begin{observation}
+There exists boards $f \in B$ with $E(f) \neq f$, but there exists some $n > 0$, $E^n(f) = f$.
+\end{observation}
+Take
+\[ f = \begin{array}{cccccc} \ddots & \vdots & \vdots & \vdots & \vdots & \reflectbox{$\ddots$} \\
+				      \dots & 0 & 4 & 0 & 4 & \dots \\
+				      \dots & 4 & 0 & 4 & 0 & \dots \\
+				      \dots & 0 & 4 & 0 & 4 & \dots \\
+				      \reflectbox{$\ddots$} & \vdots & \vdots & \vdots & \vdots & \ddots \end{array} \]
+Note that $E(f) \neq f$ but $E^2(f) = f$.
 

### 2014-09-15 FiniteBoards.tex

@@
 In this file, we analyze the case where the initial number of chips is finite. That is, the sum of all the chips on the board is finite. 
+In this file, we analyze the case where the initial number of chips is finite.
+ That is, the sum of all the chips on the board is finite.
+
+\begin{definition}
+We say that a board is stable if its values are strictly less than 4. This is equivalent to earthquakes having no effect on the board. 
+\end{definition}
 
@@
 \label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ such that after $N$ earthquakes, the system will be in a stable state (any subsequence earthquake does not cause any chips to move).
+If the initial board has a finite number of chips, there exists an $N$ such that after $N$ earthquakes, the resultant board will be in a stable.
 \end{theorem}
@@
 \[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
-Note that this is well-defined since the boards contain a finite number of chips. I will show that 
+Note that this is well-defined since the boards contain a finite number of chips. We will show that 
 \[ E(f) \neq f \Rightarrow \Phi(E(f)) > \Phi(f) \]
@@
 Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$.
-For each decrease in of location $(x,y)$ of $4$, there is an increase in two neighbors by $1$.
+For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.
 
@@
 There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in B$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \geq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
+We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
 \end{proof}

### 2014-09-19 FiniteBoards.tex
Added small earthquake
Also attempted to provide more structure/ordering of definitions.

@@
 
-In this file, we analyze the case where the initial number of chips is finite. That is, the sum of all the chips on the board is finite. 
 In this file, we analyze the case where the initial number of chips is finite.
@@
 
+\section{Preliminaries}
+\begin{definition} A board is a map from the integer lattice to the non-negative integers.
+That is a Board $f$ is an element of $\mathcal{B} := \text{Hom}(\mathbb{Z} \times \mathbb{Z}, \mathbb{Z}_{\geq 0})$.
+The input values of a board represent the coordinates on the integer lattice that the lower left hand corner of square occupies and the output value refers to the number of chips on that square. 
+A square is an element of $\mathbb{Z} \times \mathbb{Z}$. 
+\end{definition}
+
+\begin{definition} We define the big earthquake operator as a map from the set of all boards to the set of all boards and write $BE: \mathcal{B} \rightarrow \mathcal{B}$. 
+It transforms all of the squares on the board with the following rule.
+\begin{equation}
+BE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
+\end{equation}
+where $[[A]]$ is defined as 
+\begin{equation}
+[[A]] = 
+\begin{cases} 
+1 \text{ if } A \text{ is true} \\ 
+0 \text{ if } A  \text{ is false}
+ \end{cases}.
+\end{equation}
+\end{definition}
+
 \begin{definition}
-We say that a board is stable if its values are strictly less than 4. This is equivalent to earthquakes having no effect on the board. 
+ We define the small earthquake operator as a map from the set of all boards to the set of all boards and write $SE: \mathcal{B} \rightarrow \mathcal{B}$.
+It selects one square $(x,y)$ with $f(x,y) \geq 4$ and transforms it and all of its neighbors with the following rule: 
+\begin{align*}
+SE( f(x,y)) = f(x,y)-4 \\
+SE( f(x+1,y)) = f(x+1,y)+1 \\
+SE( f(x-1,y)) = f(x-1,y)+1 \\
+SE( f(x,y+1)) = f(x,y+1)+1 \\
+SE( f(x,y-1)) = f(x,y-1)+1,
+\end{align*}
+and leaves all other squares unchanged.
+\end{definition}
+
+\begin{definition}
+We say that a board is stable if its values are strictly less than 4. This is equivalent to both of the earthquake maps having no effect on the board. 
+\end{definition}
+
+\begin{definition} 
+We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
+\begin{equation}
+S= \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y) 
+\end{equation}
+is finite. 
+We denote the set of finite boards as $\mathcal{B}_f$.
+\end{definition}
+
+\begin{definition}
+We define the function $n: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 $ as 
+\begin{equation}
+n(x,y) = \{ (x+1, y), (x-1, y), (x, y-1), (x, y+1) \}
+\end{equation}
+This function takes in a square and returns its neighbors. 
 \end{definition}
@@
 \label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ such that after $N$ earthquakes, the resultant board will be in a stable.
+If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
 \end{theorem}
 
-The proof will be as follows: first, we will show that finite boards cannot return to a previous state.
-Let $B$ be the set of all possible board configurations, that is, $B = \{ f: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^{\geq 0}\}$.
-Each board configuration is given by a function of the integer lattice to the nonnegative integers.
-So $f(x,y)$ is the number of chips in the grid location $x,y$.
+We will first show that finite boards cannot return to a previous state.
+We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy.
+This shows that all finite boards
 
-The earthquake is defined as $E: B \rightarrow B$ acts in the following manner:
-\[ E(f(x,y)) = f(x,y) + \sum_{l \in n(x,y)} [[ f(l) \geq 4 ]] - 4[[f(x,y) \geq 4]] \]
-Where $n(x,y)$ is the set of neighbors of $x,y$, and $[[ X ]] = 1$ if $X$ happened and $0$ otherwise. Let $B_f \subset B$ with all boards with a finite number of chips. 
-\[ B_f = \bigcup_{n \geq 0} \{ f \in B | \sum_{x,y} f(x,y) = n \} \]
-
-So a restatement of the theorem is to say that for all $f \in B$ where $\sum_{x,y} f(x,y) = S$, there exists an $N$ such that for all $m \in \mathbb{Z}$, $E^N(f) = E^{N+m}(f)$.
+So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}$, $SE^{M}(f) = SE^{M+k}(f)$
 
 \begin{lemma}
-If $E(f) \neq f$, then $E^n(f) \neq f$ for all $n > 0$.
+If $BE(f) \neq f$, then $BE^n(f) \neq f$ for all $n > 0$.
 \end{lemma}
@@
 \begin{proof}
-Let $\Phi: B_f \rightarrow \mathbb{Z}$ be the following function
+Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
 \[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
 Note that this is well-defined since the boards contain a finite number of chips. We will show that 
-\[ E(f) \neq f \Rightarrow \Phi(E(f)) > \Phi(f) \]
+\[ E(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
 This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
 
-We can evaluate $\Phi(E(f)) - \Phi(f)$.
-\[ \Phi(E(f)) = \sum_{x,y} E(f(x,y)) (x+y)^2 \]
+We can evaluate $\Phi(BE(f)) - \Phi(f)$.
+\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (x+y)^2 \]
 \[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (x+y)^2 \]
@@
 So
-\[ \Phi(E(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
+\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
 \[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
@@
 \begin{lemma}
+If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
+\end{lemma}
+Our proof of this lemma is essentiall the same as the preceeding one.
+We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.
+
+\begin{proof}
+We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
+\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2, \]
+and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
+\begin{align}
+\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
+= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
+= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
+\end{align}
+And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
+\begin{align}
+4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
+\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
+\end{align}
+\end{proof}
+
+\begin{lemma}
 \label{finiteextension}
 Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $E^n(f)$ only contains chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
+Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
 \end{lemma}
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $f \in B$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
+(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
 Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.
 
-There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in B$ that satisfy this.
+There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in \mathcal{B}$ that satisfy this.
 We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.

### 2014-09-20 FiniteBoards.tex
Added in some needed absolute value signs to \phi
Also fixed some other minor errors in notation

@@
 Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
-\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
+\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
 Note that this is well-defined since the boards contain a finite number of chips. We will show that 
-\[ E(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
+\[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
 This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
@@
 We can evaluate $\Phi(BE(f)) - \Phi(f)$.
-\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (x+y)^2 \]
-\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (x+y)^2 \]
+\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (|x|+|y|)^2 \]
+\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
 Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$.
@@
 So
-\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
+\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
 \[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
@@
 We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
-\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2, \]
+\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
 and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then

### 2014-09-20 finitetrees.tex
Proves that all infinite n-trees stabilize
Uses a proof similar to finite boards to show that all double ended and infinite n trees eventually stabilize under large and small earthquakes. In need of proof reading and notation checking.

\documentclass[11pt]{article}
\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS
\topmargin -1.5cm        % read Lamport p.163
\oddsidemargin -0.04cm   % read Lamport p.163
\evensidemargin -0.04cm  % same as oddsidemargin but for left-hand pages
\textwidth 16.59cm
\textheight 21.94cm 
%\pagestyle{empty}       % Uncomment if don't want page numbers
\parskip 7.2pt           % sets spacing between paragraphs
%\renewcommand{\baselinestretch}{1.5} % Uncomment for 1.5 spacing between lines
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{verbatim}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\parindent 0pt		 % sets leading space for paragraphs
\author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
\title{18.821 - Poker Chips Finite board configurations}

\begin{document}

\maketitle

\section{$n$-Trees}
Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.

We can join two semi infinite $n$-trees by an edge connecting their parent nodes. This forms a graph where every vertex has $n+1$ neighbors. We would like to adapt our description of chips and earthquakes to this graph.

\begin{definition} We define a coordinate system for our graph, with the following rules:
\begin{enumerate}
	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq n^{|a|}$
	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq n^{|a-1|}$
\end{enumerate}
We define the set of all valid coordinates $G_n$ as 
\begin{equation}
G_n := \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
\end{equation}
\end{definition}

\begin{definition}
We define a tree to be a map from the set of valid coordinates to the nonnegative integers, that is 
\begin{equation}
f:  G\rightarrow \mathbb{Z}_{\geq 0}
\end{equation}
We write $\mathcal{T}_n$ to refer to the set of all joined $n$-tree's 
\end{definition}

\begin{definition}
\label{bigearthquakedeftree}
We define the big earthquake operator as  as a map from the set of all boards to the set of all boards and write $BE: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
It transforms all of the nodes on the tree with the following rule.
\begin{equation}
BE(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
\end{equation} 
where $k$ is defined as 
\begin{equation}
k = \begin{cases}
1 \text{ if } x\leq 0 \\
-1 \text{ if } x \geq 1
\end{cases}
\end{equation}
and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$ and $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the small earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $SE: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
It selects one valid node $(x,y)$ with $f(x,y) \geq n+1$ and transforms it and all of its neighbors with the following rule: 
\begin{align*}
SE( f( x, y ) ) = f( x , y )-n-1 \\
SE( f( x + k, y)) = f( x + k, y )+1 \\
SE( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
\end{align*}
where $k$ is as defined above and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$
and leaves all other squares unchanged.
\end{definition}

\begin{definition}
We say that a $n$-Tree is stable if its values are strictly less than $n+1$. This is equivalent to both of the earthquake maps having no effect on the board. 
\end{definition}

\begin{definition} 
We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
\begin{equation}
S= \sum_{(x,y) \in G_n} f(x,y) 
\end{equation}
is finite. 
We denote the set of finite boards as $\mathcal{T}_n^f$.
\end{definition}

As it turns out we can use a  method of proof nearly identical to the standard board case to show that the state of a tree under this process can never cycle.

We once again define a potential function 
\begin{definition} Let $\Phi: \mathcal{T}_n^f \rightarrow \mathbb{Z}_{\geq 0}$ be the  function
\begin{equation}
\Phi(f) = \sum_{(x,y) \in G_n} f(x,y)(|x|)^2
\end{equation}
Note that this is well defined for trees which contain a finite number of chips. 
\end{definition}

\begin{lemma}
If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that  if $SE$ acts on a board nontrivially then $\Phi(SE(f))$ will be strictly larger. That is  
\begin{equation}
SE(f) \neq f \Rightarrow \Phi(SE(f)) > \Phi(f)
\end{equation}

We can evaluate $\Phi(SE(f)) - \Phi(f)$.
\begin{align}
\Phi(SE(f))-\Phi(f) = \sum_{(x,y) \in G_n} SE(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
\end{align}
Only one node call it $(x_0,y_0)$ and its neighbors are affected. Therefore the above expression simplifies to
\begin{align*}
\Phi(SE(f)-\Phi(f) = SE( f(x_0,y_0) )* (x_0)^2 + SE( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
+ \sum_{i} SE( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
= -(n+1)*(|x_0|)^2 + (|x_0+k)^2 + n (|x_0-k|)^2 \\
\geq-(n+1)*(|x_0|)^2 + (|x_0|-1)^2 + n (|x_0|+1)^2
= -nx^2-x^2+x^2-2|x|+1+n(x^2+2|x|+1) \\
= 2x(n-1)+n+1 > 0
\end{align*}
Where $(x_0-k,y_i)$ refers to the $j$th child of $(x_0,y_0)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
\end{proof}

\begin{lemma}
If $BE(f) \neq f$, then $BE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that 
\begin{equation}
BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f)
\end{equation}
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE(f)) - \Phi(f)$.
\begin{align}
\Phi(BE(f))-\Phi(f) = \sum_{(x,y) \in G_n} BE(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
\end{align}
Let $A$ be the set of active locations, or locations with $f(x,y) \geq (n+1)$, which is non-empty by assumption. 
\begin{align*}
\Phi(BE(f)) - \Phi(f) =  \sum_{(x,y) \in A}  BE(f(x,y))x^2 + BE(f(x+k,y))(|x+k|)^2 + \\
\sum_{i} \left( BE(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
=\sum_{(x,y) \in A} -(n+1)x^2 + (|x+k|)^2 + n(|x-k|)^2 
\end{align*}
Where $(x-k,y_i)$ refers to the $j$th child of $(x,y)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}. We note that each term in this sum is positive by our proof of the lemma for small earthquakes and so the sum itself must be positive. That is 
\begin{equation}
\Phi(BE(f)) - \Phi(f) > 0
\end{equation}

\end{proof}

\begin{lemma}
\label{finiteextensiontree}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $BE^m(f)$ and $SE^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
\end{lemma}

\begin{proof}
It is obvious that a board with $S$ chips on level $l<0$ will have a lower $X_{min}$ than a board with $S$ chips on level 0, and the analogous statement for $X_{max}$ holds as well.

Now as a very crude lower bound we know that $S$ chips on a single node at level $l<0$ will have an $X_{min}$ no lower than $l-2S$ as there must be at least one chip on every other level, as chips only move one level at a time.

If we take any other arrangement of $S$ chips, with $X_{min}=l$ but not all $S$ chips are on a single node. We know that $X_{min}$ for this board is bounded below by $l-2S$. In order to have an $X_{min}$ below this at some point in time it would be possible to have $S$ chips on the same node on level $l$. But this cannot occur because in order to move a chip down a level, one chip must move up a level. Therefore if $X_{min}=x_l$, both $BE^m(f)$ and $SE^m(f)$ do not contain chips below level $x_l -2S$.

An analogous argument shows that if $X_{max}=x_h$, both $BE^m(f)$ and $SE^m(f)$ do not contain chips above level $x_h+2S$.
\end{proof}

So now we are ready to prove the theorem that all finite boards will reach a stable state.

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie between levels  $[x_l, x_h]$. Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $BE^m(f)$ and $SE^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$

We have restricted the size of the board to be finite, and the number of chips that can be placed on the board to be finite. 
Thus the total number of ways that the chips can be arranged on the board is finite. 
In particular there are fewer than $(x_h - x_l+1 + 4 S)$ levels and there are at most 
\begin{equation}
n^d,
\end{equation}
nodes at each level, where $d= \max(x_h, |x_l|)$. So there are at most 
\begin{equation}
N:=(x_h - x_l+1 + 4 S)n^d
\end{equation}
nodes that are potentially occupied. Therefore there are at most 
\begin{equation}
\binom{N+S-1}{N-1}
\end{equation}
possible boards $b \in \mathcal{T}_n^f$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < \binom{N+S-1}{N-1}$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}

\end{document}

### 2014-09-22 FiniteBoards.tex
modified stuff

--- 2014-09-15 FiniteBoards.tex
@@
 Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $E^n(f)$ only contains chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
+Then $E^n(f)$ only contains chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
 \end{lemma}
@@
 \begin{proof}
-For each chip that moves in one direction, there is another chip that moves in the opposite direction.
-Therefore, at each time, the number of chips that extend the range can be at most half of what extended before.
-This means that there are at most $\log S$ extensions on each side.
+Let $A$ be an initial region of the board
 \end{proof}

### 2014-09-22 simulate.py

@@ -47,6 +47,10 @@ def getStableBoundry(board):
     actUntilStable(board)
     return getBoundary(board)
 
+def getStableBoundryDistance(n):
+    board = {(0,0):n}
+    return getStableBoundry(board)[1]
+
 def printBoundary(board):
     (lowestX, highestX, lowestY, highestY) = getBoundary(board)
     print "Boundary: X:[%d, %d] \tY:[%d, %d]" % (lowestX, highestX, lowestY, lowestY)
@@ -68,9 +72,83 @@ def actUntilStable(board):
         count += 1
     return count
 
+def actUntilStableCountingCenter(n):
+    count = 0
+    board = {(0,0):n}
+    while act(board):
+        if board[(0,0)] >= 4:
+            count += 1
+    return count
+
+def actUntilStableCountingPosition(n, positions):
+    counts = {}
+    for p in positions:
+        counts[p] = 0
+    board = {(0,0): n}
+    while act(board):
+        for pos in positions:
+            try:
+                if board[pos] >= 4:
+                    counts[pos] += 1
+            except KeyError:
+                print 'badaccess'
+    return counts
+
+import matplotlib as mpl
+import pylab
+import math
+
+def seeRelationshipBetweenSquares(n):
+    x = range(2, int(math.log(n)/math.log(4)))
+    positions = [(i, 0) for i in x]
+    counts = actUntilStableCountingPosition(n, positions)
+    y = [counts[(i, 0)]*i/math.log(i) for i in x if i != 0]
+    mpl.pyplot.scatter(x,y)
+    mpl.pyplot.show()
+
 def actAndSee(board):
     while 1:
         act(board)
         printBoard(board)
         input = raw_input()
 
+import Image,ImageDraw
+from random import randint as rint
+import pylab as pl
+
+def drawBoard(board, xrange, yrange, maxValue, recomputeMax=False):
+    if recomputeMax:
+        maxValue = max(board.values())
+    img = Image.new("RGB", (xrange, yrange), "#FFFFFF")
+    draw = ImageDraw.Draw(img)
+    dr = 255./(maxValue/3)
+    dg = 255./(maxValue/3)
+    db = 255./(maxValue/3)
+    for i in range(xrange):
+        for j in range(yrange):
+            try:
+                point = board[(i - xrange/2, j - yrange/2)]
+            except KeyError:
+                point = 0
+            r, g, b = point*dr, 0, 0
+            if r > 255:
+                r = 255
+                g = point*dg
[…]

### 2014-09-22 FiniteBoards.tex

--- 2014-09-20 FiniteBoards.tex
@@
 Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
+Then $E^n(f)$ only contains chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
 \end{lemma}

### 2014-09-22 currentresults.tex

@@
 \begin{proof}
-The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final result.
-This technique will then uniquely (up to a point) specify an order in which the earthquakes will occur, which completes the proof.
+The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
+This technique will then specify an order in which the earthquakes will occur, which completes the proof.
 
@@
 Let $E_1,E_2,E_3,\dots, E_k$ be a sequence of earthquakes, where each $E_i$ specifies the location of a square hit.
-Then the sequence given by moving some earthquake $E_j$ forward in the sequence to a new spot where $E_j$ is still a valid earthquake gives the same final configuration.
+Then the sequence given by moving some earthquake $E_j$ forward in the sequence to a new spot where $E_j$ is still a valid earthquake gives the same final configuration, and the sequence of earthquakes is still entirely valid.
 \end{lemma}
@@
 
-Now, given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
+Given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
+We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes or else it will not eventually stabilize.
+Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquake events that correspond to square $i$.
+All these earthquakes can be moved to the front of the sequence.
+We do this for all squares $i$.
+By the lemma, the final positioning of the board does not change.
+
+The order of the earthquakes that have been moved to the front of the sequence can be in any order.
+This holds true because all these earthquakes could have happened with the initial values of the squares (maybe need to explain this better?
+but i think it's clear...).
+
+Now, find the point in the sequence that begins right after this point.
+Consider the board at this current point in time, and note that we have new values for all the squares.
+Then repeat this argument, moving a new set of earthquakes to the front.
+We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
 

### 2014-09-22 draft0.tex

\documentclass[11pt]{article}
\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS
\topmargin -1.5cm        % read Lamport p.163
\oddsidemargin -0.04cm   % read Lamport p.163
\evensidemargin -0.04cm  % same as oddsidemargin but for left-hand pages
\textwidth 16.59cm
\textheight 21.94cm 
%\pagestyle{empty}       % Uncomment if don't want page numbers
\parskip 7.2pt           % sets spacing between paragraphs
%\renewcommand{\baselinestretch}{1.5} % Uncomment for 1.5 spacing between lines
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{verbatim}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\parindent 0pt		 % sets leading space for paragraphs
\author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
\title{18.821 - Poker Chips Finite board configurations}

\begin{document}         
\maketitle

In this file, we analyze the case where the initial number of chips is finite.
That is, the sum of all the chips on the board is finite.

\section{Standard Board}

\begin{definition} A board is a map from the integer lattice to the non-negative integers.
That is a Board $f$ is an element of $\mathcal{B} := \text{Hom}(\mathbb{Z} \times \mathbb{Z}, \mathbb{Z}_{\geq 0})$.
The input values of a board represent the coordinates on the integer lattice that the lower left hand corner of square occupies and the output value refers to the number of chips on that square. 
A square is an element of $\mathbb{Z} \times \mathbb{Z}$. 
\end{definition}

\begin{definition} We define the big earthquake operator as a map from the set of all boards to the set of all boards and write $BE: \mathcal{B} \rightarrow \mathcal{B}$. 
It transforms all of the squares on the board with the following rule.
\begin{equation}
BE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
\end{equation}
where $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the small earthquake operator as a map from the set of all boards to the set of all boards and write $SE: \mathcal{B} \rightarrow \mathcal{B}$.
It selects one square $(x,y)$ with $f(x,y) \geq 4$ and transforms it and all of its neighbors with the following rule: 
\begin{align*}
SE( f(x,y)) = f(x,y)-4 \\
SE( f(x+1,y)) = f(x+1,y)+1 \\
SE( f(x-1,y)) = f(x-1,y)+1 \\
SE( f(x,y+1)) = f(x,y+1)+1 \\
SE( f(x,y-1)) = f(x,y-1)+1,
\end{align*}
and leaves all other squares unchanged.
\end{definition}

\begin{definition}
We say that a board is stable if its values are strictly less than 4. This is equivalent to both of the earthquake maps having no effect on the board. 
\end{definition}

\begin{definition} 
We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
\begin{equation}
S= \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y) 
\end{equation}
is finite. 
We denote the set of finite boards as $\mathcal{B}_f$.
\end{definition}

\begin{definition}
We define the function $n: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 $ as 
\begin{equation}
n(x,y) = \{ (x+1, y), (x-1, y), (x, y-1), (x, y+1) \}
\end{equation}
This function takes in a square and returns its neighbors. 
\end{definition}

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
\end{theorem}

We will first show that finite boards cannot return to a previous state. We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy. This shows that all finite boards

So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}$, $SE^{M}(f) = SE^{M+k}(f)$

\begin{lemma}
If $BE(f) \neq f$, then $BE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE(f)) - \Phi(f)$.
\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (|x|+|y|)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.

\begin{proof}
We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
\begin{align}
\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
\end{align}
And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
\begin{align}
4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
\end{align}
\end{proof}

\begin{lemma}
\label{finiteextension}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$. Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
\end{lemma}

\begin{proof}
For each chip that moves in one direction, there is another chip that moves in the opposite direction. Therefore, at each time, the number of chips that extend the range can be at most half of what extended before. This means that there are at most $\log S$ extensions on each side. 
\end{proof}

So now we are ready to prove the theorem that all finite boards will reach a stable state.

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$. Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.

There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in \mathcal{B}$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}

\section{$n$-Trees}
Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.

We can join two semi infinite $n$-trees by an edge connecting their parent nodes. This forms a graph where every vertex has $n+1$ neighbors. We would like to adapt our description of chips and earthquakes to this graph.

\begin{definition} We define a coordinate system for our graph, with the following rules:
\begin{enumerate}
	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq n^{|a|}$
	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq n^{|a-1|}$
\end{enumerate}
We define the set of all valid coordinates $G_n$ as 
\begin{equation}
G_n := \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
\end{equation}
\end{definition}

\begin{definition}
We define a tree to be a map from the set of valid coordinates to the nonnegative integers, that is 
\begin{equation}
f:  G\rightarrow \mathbb{Z}_{\geq 0}
\end{equation}
We write $\mathcal{T}_n$ to refer to the set of all joined $n$-tree's 
\end{definition}

\begin{definition}
\label{bigearthquakedeftree}
We define the big earthquake operator as  as a map from the set of all $n$ trees to the set of all $n$ trees and write $BE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
It transforms all of the nodes on the tree with the following rule.
\begin{equation}
BE_T(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
\end{equation} 
where $k$ is defined as 
\begin{equation}
k = \begin{cases}
1 \text{ if } x\leq 0 \\
-1 \text{ if } x \geq 1
\end{cases}
\end{equation}
and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$ and $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the small earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $SE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
It selects one valid node $(x,y)$ with $f(x,y) \geq n+1$ and transforms it and all of its neighbors with the following rule: 
\begin{align*}
SE_T( f( x, y ) ) = f( x , y )-n-1 \\
SE_T( f( x + k, y)) = f( x + k, y )+1 \\
SE_T( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
\end{align*}
where $k$ is as defined above and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$
and leaves all other squares unchanged.
\end{definition}

\begin{definition}
We say that a $n$-Tree is stable if its values are strictly less than $n+1$. This is equivalent to both of the earthquake maps having no effect on the board. 
\end{definition}

\begin{definition} 
We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
\begin{equation}
S= \sum_{(x,y) \in G_n} f(x,y) 
\end{equation}
is finite. 
We denote the set of finite boards as $\mathcal{T}_n^f$.
\end{definition}

As it turns out we can use a  method of proof nearly identical to the standard board case to show that the state of a tree under this process can never cycle.

We once again define a potential function 
\begin{definition} Let $\Phi: \mathcal{T}_n^f \rightarrow \mathbb{Z}_{\geq 0}$ be the  function
\begin{equation}
\Phi(f) = \sum_{(x,y) \in G_n} f(x,y)(|x|)^2
\end{equation}
Note that this is well defined for trees which contain a finite number of chips. 
\end{definition}

\begin{lemma}
If $SE_T(f) \neq f$, then $SE_T^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that  if $SE$ acts on a board nontrivially then $\Phi(SE(f))$ will be strictly larger. That is  
\begin{equation}
SE(f) \neq f \Rightarrow \Phi(SE(f)) > \Phi(f)
\end{equation}

We can evaluate $\Phi(SE(f)) - \Phi(f)$.
\begin{align}
\Phi(SE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} SE_T(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
\end{align}
Only one node call it $(x_0,y_0)$ and its neighbors are affected. Therefore the above expression simplifies to
\begin{align*}
\Phi(SE_T(f)-\Phi(f) = SE_T( f(x_0,y_0) )* (x_0)^2 + SE_T( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
+ \sum_{i} SE_T( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
= -(n+1)*(|x_0|)^2 + (|x_0+k)^2 + n (|x_0-k|)^2 \\
\geq-(n+1)*(|x_0|)^2 + (|x_0|-1)^2 + n (|x_0|+1)^2
= -nx^2-x^2+x^2-2|x|+1+n(x^2+2|x|+1) \\
= 2x(n-1)+n+1 > 0
\end{align*}
Where $(x_0-k,y_i)$ refers to the $j$th child of $(x_0,y_0)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
\end{proof}

\begin{lemma}
If $BE_T(f) \neq f$, then $BE_T^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that 
\begin{equation}
BE_T(f) \neq f \Rightarrow \Phi(BE_T(f)) > \Phi(f)
\end{equation}
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE_T(f)) - \Phi(f)$.
\begin{align}
\Phi(BE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} BE_T(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
\end{align}
Let $A$ be the set of active locations, or locations with $f(x,y) \geq (n+1)$, which is non-empty by assumption. 
\begin{align*}
\Phi(BE_T(f)) - \Phi(f) =  \sum_{(x,y) \in A}  BE_T(f(x,y))x^2 + BE_T(f(x+k,y))(|x+k|)^2 + \\
\sum_{i} \left( BE_T(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
=\sum_{(x,y) \in A} -(n+1)x^2 + (|x+k|)^2 + n(|x-k|)^2 
\end{align*}
Where $(x-k,y_i)$ refers to the $j$th child of $(x,y)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}. We note that each term in this sum is positive by our proof of the lemma for small earthquakes and so the sum itself must be positive. That is 
\begin{equation}
\Phi(BE_T(f)) - \Phi(f) > 0
\end{equation}

\end{proof}

\begin{lemma}
\label{finiteextensiontree}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $BE_T^m(f)$ and $SE_T^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
\end{lemma}

\begin{proof}
It is obvious that a board with $S$ chips on level $l<0$ will have a lower $X_{min}$ than a board with $S$ chips on level 0, and the analogous statement for $X_{max}$ holds as well.

Now as a very crude lower bound we know that $S$ chips on a single node at level $l<0$ will have an $X_{min}$ no lower than $l-2S$ as there must be at least one chip on every other level, as chips only move one level at a time.

If we take any other arrangement of $S$ chips, with $X_{min}=l$ but not all $S$ chips are on a single node. We know that $X_{min}$ for this board is bounded below by $l-2S$. In order to have an $X_{min}$ below this at some point in time it would be possible to have $S$ chips on the same node on level $l$. But this cannot occur because in order to move a chip down a level, one chip must move up a level. Therefore if $X_{min}=x_l$, both $BE_T^m(f)$ and $SE_T^m(f)$ do not contain chips below level $x_l -2S$.

An analogous argument shows that if $X_{max}=x_h$, both $BE_T^m(f)$ and $SE_T^m(f)$ do not contain chips above level $x_h+2S$.
\end{proof}

So now we are ready to prove the theorem that all finite boards will reach a stable state.

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie between levels  $[x_l, x_h]$. Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $BE_T^m(f)$ and $SE_T^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$

We have restricted the size of the board to be finite, and the number of chips that can be placed on the board to be finite. 
Thus the total number of ways that the chips can be arranged on the board is finite. 
In particular there are fewer than $(x_h - x_l+1 + 4 S)$ levels and there are at most 
\begin{equation}
n^d,
\end{equation}
nodes at each level, where $d= \max(x_h, |x_l|)$. So there are at most 
\begin{equation}
N:=(x_h - x_l+1 + 4 S)n^d
\end{equation}
nodes that are potentially occupied. Therefore there are at most 
\begin{equation}
\binom{N+S-1}{N-1}
\end{equation}
possible boards $b \in \mathcal{T}_n^f$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < \binom{N+S-1}{N-1}$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}

\section{Order of Redistribution}
\begin{definition}
We define a valid earthquake to be a small earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.

We say that a sequence of small earthquakes $SE_1, SE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
\end{definition}

\begin{theorem}
Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ small earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
\end{theorem}

\begin{proof}
The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board. This technique will then specify an order in which the earthquakes will occur, which completes the proof.

First, we need a lemma in order to do the rearrangements.

\begin{lemma} Let $SE_1$ and $SE_2$ be small earthquakes. If $SE_1, SE_2$ is a valid sequence and $SE_2$ is a valid earthquake then when before $SE_1$ then $SE_2, SE_1$ is also a valid sequence of earthquakes and $SE_2( SE_1 ( B ) ) = SE_1( SE_2 ( B ) )$
\end{lemma}
\begin{proof}
Suppose $SE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $SE_2$ affects the square $(x_2, y_2)$ and its neighbors. Suppose  $SE_2, SE_1$ was an invalid sequence, that is $SE_2(B)(x_1, y_1)<4$, but since $SE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$. Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$. But then $SE_1(B)(x_1, y_1)=3$ which means $SE_1, SE_2$ would not be a valid sequence of earthquakes. Thus $SE_2, SE_1$ is a valid sequence.

Now consider $SE_2( SE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the small earthquake we see that 
\begin{align*}
SE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
SE_2 ( SE_1 (B)) (u,v) = SE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
= B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]]  \\
= B(u,v) + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
= SE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
= SE_1 (SE_2 (B))(u,v) 
\end{align*}
The critical step in this proof is recognizing that because $SE_1$ and $SE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

\begin{lemma}
Let $SE_1,SE_2,SE_3,\dots, SE_k$ be a sequence of earthquakes, where $SE_i$ hits the square $(x_i, y_i)$. Consider  the sequence given by moving some earthquake $SE_j$ to position $m<j$ in the sequence, such that $SE_j$ is a valid earthquake for all positions between $m$ and $j$. 
This sequence is valid and produces the same final board configuration.
\end{lemma}
\begin{proof}
It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes. In fact we can also ignore the first $m-1$ terms, because their order is not changed they will all be valid and the board state will be identical after $m-1$ terms for both sequences. Let us call this board state $B_{m-1}$

Therefore we must show that the sequence $SE_j, SE_m, SE_{m+1}, \ldots, SE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $SE_m, SE_{m+1}, \ldots, SE_{j-1}, SE_j$ applied to $B_{m-1}$.

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in the above lemma. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
SE_m, SE_{m+1}, \ldots, SE_{j-1}, SE_j
\end{equation*}
where $p+1=j-m+1$. Then if we set $B_{m} = SE_{m} (B_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
SE_{m+1}, \ldots, SE_{j-1}, SE_{j} \\ 
SE_{j}, SE_{m+1}, \ldots,  SE_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $B_{m-1}$ by our above Lemma we know that the two sequences 
\begin{align*}
SE_{m}, SE_{j} \\
SE_{j}, SE_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction as above  we have  that $SE_j, SE_m, SE_{m+1}, \ldots, SE_{j-1}$  and $SE_m, SE_{m+1}, \ldots, SE_{j-1}, SE_j$ both applied to $B_{m-1}$ are valid and produce the same board. 
\end{proof}

Given the original configuration of the board, let $A_i$ be the number of chips on square $i$. We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes or else it will not eventually stabilize. Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquake events that correspond to square $i$. All these earthquakes can be moved to the front of the sequence. We do this for all squares $i$. By the lemma, the final positioning of the board does not change.

The order of the earthquakes that have been moved to the front of the sequence can be in any order. This holds true because all these earthquakes could have happened with the initial values of the squares (maybe need to explain this better? but i think it's clear...).

Now, find the point in the sequence that begins right after this point. Consider the board at this current point in time, and note that we have new values for all the squares. Then repeat this argument, moving a new set of earthquakes to the front. We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).

\end{proof}

\end{document}

### 2014-09-22 draft0.tex
changed notation a bit, corrected finite boards section

@@
 \begin{definition} A board is a map from the integer lattice to the non-negative integers.
-That is a Board $f$ is an element of $\mathcal{B} := \text{Hom}(\mathbb{Z} \times \mathbb{Z}, \mathbb{Z}_{\geq 0})$.
+That is a Board $f$ is an element of $\mathcal{B} := \{ f: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}_{\geq 0} | f \text{ is a function }\}$.
 The input values of a board represent the coordinates on the integer lattice that the lower left hand corner of square occupies and the output value refers to the number of chips on that square. 
@@
 \begin{definition}
- We define the small earthquake operator as a map from the set of all boards to the set of all boards and write $SE: \mathcal{B} \rightarrow \mathcal{B}$.
-It selects one square $(x,y)$ with $f(x,y) \geq 4$ and transforms it and all of its neighbors with the following rule: 
+ We define the small earthquake operator as a map from the set of all boards and locations to the set of all boards and write $SE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
+If $f(x,y) \leq 4$, then $SE(f, (x,y)) = f$, otherwise,  
 \begin{align*}
-SE( f(x,y)) = f(x,y)-4 \\
-SE( f(x+1,y)) = f(x+1,y)+1 \\
-SE( f(x-1,y)) = f(x-1,y)+1 \\
-SE( f(x,y+1)) = f(x,y+1)+1 \\
-SE( f(x,y-1)) = f(x,y-1)+1,
+SE( f(x,y), (x,y)) = f(x,y)-4 \\
+SE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
+SE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
+SE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
+SE( f(x,y-1), (x,y)) = f(x,y-1)+1,
 \end{align*}
@@
 
-So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}$, $SE^{M}(f) = SE^{M+k}(f)$
+So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $SE^{M}(f) = SE^{M+k}(f)$
 
@@
 Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
+Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
 \end{lemma}
@@
 \begin{proof}
-For each chip that moves in one direction, there is another chip that moves in the opposite direction.
-Therefore, at each time, the number of chips that extend the range can be at most half of what extended before.
-This means that there are at most $\log S$ extensions on each side.
+The most compact stable state of a board with $S$ chips is to have $3$ chips on each square adjacent to each other.
+This happens in a square $\sqrt{S/3} \times \sqrt{S/3}$.
+Similarly, a connected component of chips can only extend by a certain amount, where there are no two 0s adjacent to each other.
+This means that the most spread board which started with a connected component of $S$ chips can be $2\sqrt{S} \times 2\sqrt{S}$.
+
+If we do this for each initial connected component and take the total area of the board, we get the desired bound.
 \end{proof}
@@
 (of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
-Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.
+Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
 
-There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in \mathcal{B}$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
+There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}$ that satisfy this.
+We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
 \end{proof}

### 2014-09-24 draft0.tex
drafted computation stuff

@@
 
+\section{Computing with Chips}
+
+In this section, we show how we can compute different boolean functions with chips.
+We will show how we can build circuits and prove some impossibility results for certain circuits.
+This might seem similar to computation with Game of Life; however, we note that these two are inherently different since Chips will always terminate with a finite board, whereas Game of Life might not.
+
+\subsection{Model of Computation}
+
+Our model of computation is as follows: we assume a finite board with finitely many chips.
+The inputs are designated by each two squares, one apart one is the ``clock" square and other the bit square.
+Likewise, the outputs each have a ``clock" square and a bit square.
+The ``clock" squares are set the value 4 and the bit squres are set to 4 if they are 1 and 0 otherwise.
+A computation is done by a series of earthquakes, when an output has the clock square active, then the bit square is read, and that output bit is considered read.
+
+\subsection{Example}
+
+A simple example is given by a board that computes the identity.
+
+\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
+				     4_c & 3 & 3 & 3 & 3 & 3_e \\
+				     0 & 0 & 0 & 0 & 0 & 0 \\
+				     i  & 3 & 3 & 3 & 3 & 3_o \\
+				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
+Where the subscript $c$ indicates the clock, the subscript $e$ indicates the output clock square. $i$ indicates the bit square, and $o$ indicates the output bit.
+
+One can see that $e$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $e$.
+Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active.
+Therefore, we can compute the identity.
+
+This circuit also shows how to make a ``wire" in the circuit. We will represent a wire as two parralel arrows, one containing the clock path and the other carrying the input.
+
+\subsection{Turner}
+
+Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
+this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $e$ is the output clock and $o$ is the output bit.
+
+\[ \begin{array}{cccccc} 0 & 0 & e & 0 & o \\
+				     0 & 3 & 2 & 0 & 3 \\
+			            0 & 2 & 3 & 0 & 3 \\
+				    0 & 3 & 2 & 0 & 3 \\
+				    0 & 2 & 3 & 0 & 3 \\
+				    c & 3 & 3 & 0 & 3 \\
+				    0 & 0 & 0 & 0 & 3 \\
+				    i & 3 & 3 & 3 & 3 \end{array} \]	
+Since the clock snakes along the path while $i$ takes the shorter path. 
+Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
+
+\subsection{Synchronize}
+
+We can synchronize the inputs so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
+\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
+					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
+					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
+					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
+					 0 & 0 & 0 & 0 & o & 0 & e & 3 \end{array} \]
+Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
+$i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.
+
+\subsection{Boolean circuits}
+
+Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
+The following can compute the AND of two inputs.
+
+\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
+					   &    & 3   & 2 &  3    &    &  \\
+					   &    &    & 3 &    & 3 & c_2 \\
+					o & 3 & 3 & 2 & 3 & 2 & 	    \\
+					   &    &    &    &    & 3 & i_2 \end{array} \]
+Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
+So in this case, $o = i_1$ AND $i_2$.
+
+Similarly, we can compute the OR of two circuits with a very simple modification:
+\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
+					   &    & 3   & 2 &  3    &    &  \\
+					   &    &    & 3 &    & 3 & c_2 \\
+					o & 3 & 3 & 3 & 3 & 2 & 	    \\
+					   &    &    &    &    & 3 & i_2 \end{array} \]
+And we can do a similar set up to synchronize the clocks.
+
+So currently, we can compute any boolean circuit which can be drawn in a planar graph with only ANDs and ORs. So the question is whether we can compute the NOT? We argue that this is not possible.
+
+\begin{theorem}
+There does not exists any board configuration which computes the NOT operator in the above model of computation.
+\end{theorem}
+
+\begin{proof}
+Suppose we have a board configuration $B$ with input squares $c$, $i$ and output squares $e, o$ that computes the NOT function.
+Then we can make another board $B'$ that synchronizes the final clock with the input so that we do not need to worry about timing of the clock.
+That is, we can guarantee that if $o$ is active, then $o$ is active at the same time $e$ is active.
+
+Now lets make $B'$ an infinite board by putting 0s on the sides.
+So we have a finite board with finitely many chips.
+Let $S_{c_0i_0}$ be the set of squares which were active (had an earthquake affect them) during the execution of $B$ with the input value $i = i_0$ and $c = c_0$.
+
+The first claim to make is that $S_{c_0i_0} \subset S_{c_1i_1}$ if $c_0 \leq c_1$ and $i_0 \leq i_1$.
+This is clearly true since the order of redistribution does not matter.
+In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with small earthquakes.
+
+However, this arises in a contradiction, since $o \in S_{10}$ but $o \notin S_{11}$. Therefore, computing NOT with a board is not possible.
+\end{proof}
+
 \end{document}

### 2014-09-25 poker chips format.tex
added an alternative format

--- 2014-09-24 draft0.tex
@@
-\documentclass[11pt]{article}
-\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS
-\topmargin -1.5cm        % read Lamport p.163
-\oddsidemargin -0.04cm   % read Lamport p.163
-\evensidemargin -0.04cm  % same as oddsidemargin but for left-hand pages
-\textwidth 16.59cm
-\textheight 21.94cm 
-%\pagestyle{empty}       % Uncomment if don't want page numbers
-\parskip 7.2pt           % sets spacing between paragraphs
-%\renewcommand{\baselinestretch}{1.5} % Uncomment for 1.5 spacing between lines
+\documentclass[runningheads,a4paper]{llncs}
+
+\usepackage{amssymb}
+\setcounter{tocdepth}{3}
+\usepackage{graphicx}
 \usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\usepackage{verbatim}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
-\parindent 0pt		 % sets leading space for paragraphs
-\author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
-\title{18.821 - Poker Chips Finite board configurations}
+\usepackage[margin=0.9in]{geometry}
+%\usepackage{amsfonts}
+%\usepackage{amsthm}
+\usepackage{subfigure}
+%\usepackage{caption}
+%\usepackage{subcaption}
+%\usepackage{cite}
+\usepackage{hyperref}
+\usepackage{url}
+\urlstyle{same}
+\newcommand{\keywords}[1]{\par\addvspace\baselineskip
+\noindent\keywordname\enspace\ignorespaces#1}
+
+\makeatletter
+\let\c@lemma=\c@theorem
+\let\c@corollary=\c@theorem
+\let\c@fact=\c@theorem
+\makeatother
+
+\let\realendproof=\endproof
+\def\endproof{\hspace*{\fill}$\Box$\realendproof}
 
 \begin{document}
+
+\title{Poker Chips and Earthquakes}
+\titlerunning{Poker Chips and Earthquakes}
+
+\author{Perry Kleinhenz \and Fermi Ma \and Erik Waingarten}
+%
+\authorrunning{Perry Kleinhenz \and Fermi Ma \and Erik Waingarten}
+% (feature abused for this document to repeat the title also on left hand pages)
+
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{
+\protect\url{{pkleinhe, fermima,eaw}@mit.edu}}
+
 \maketitle
 
-In this file, we analyze the case where the initial number of chips is finite.
-That is, the sum of all the chips on the board is finite.
+\section{Introduction}
 
+Written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.
+
+In this paper, we investigate the following problem set up:
+
+\vbox{
+\noindent
+\begin{quote}
+Each square of an infinite chess board contains some nonnegative number of poker chips.
+An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions.
+\end{quote}
+}
+We also consider extensions of this problem set up where the earthquake hits all squares on the board at once.
+We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
 \section{Standard Board}
+
+Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz
 
@@
 \section{Order of Redistribution}
+
+written by Fermi Ma and Perry Kleinhenz, edited by Erik Waingarten.
+
 \begin{definition}
@@
 \section{Computing with Chips}
+
+written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz
 

### 2014-09-25 draft0.tex
added editing and writing tags
changed format of draft0

@@
-\documentclass[11pt]{article}
-\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS
-\topmargin -1.5cm        % read Lamport p.163
-\oddsidemargin -0.04cm   % read Lamport p.163
-\evensidemargin -0.04cm  % same as oddsidemargin but for left-hand pages
-\textwidth 16.59cm
-\textheight 21.94cm 
-%\pagestyle{empty}       % Uncomment if don't want page numbers
-\parskip 7.2pt           % sets spacing between paragraphs
-%\renewcommand{\baselinestretch}{1.5} % Uncomment for 1.5 spacing between lines
+\documentclass[runningheads,a4paper]{llncs}
+
+\usepackage{amssymb}
+\setcounter{tocdepth}{3}
+\usepackage{graphicx}
 \usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\usepackage{verbatim}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
-\parindent 0pt		 % sets leading space for paragraphs
-\author{Perry Kleinhenz, Fermi Ma, and Erik Waingarten}
-\title{18.821 - Poker Chips Finite board configurations}
+\usepackage[margin=0.9in]{geometry}
+%\usepackage{amsfonts}
+%\usepackage{amsthm}
+\usepackage{subfigure}
+%\usepackage{caption}
+%\usepackage{subcaption}
+%\usepackage{cite}
+\usepackage{hyperref}
+\usepackage{url}
+\urlstyle{same}
+\newcommand{\keywords}[1]{\par\addvspace\baselineskip
+\noindent\keywordname\enspace\ignorespaces#1}
+
+\makeatletter
+\let\c@lemma=\c@theorem
+\let\c@corollary=\c@theorem
+\let\c@fact=\c@theorem
+\makeatother
+
+\let\realendproof=\endproof
+\def\endproof{\hspace*{\fill}$\Box$\realendproof}
 
 \begin{document}
+
+\title{Computing with an Angry Poker Player}
+\titlerunning{Computing with an Angry Poker Player}
+
+\author{Perry Kleinhenz \and Fermi Ma \and Erik Waingarten}
+%
+\authorrunning{Perry Kleinhenz \and Fermi Ma \and Erik Waingarten}
+% (feature abused for this document to repeat the title also on left hand pages)
+
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{
+\protect\url{{pkleinhe, fermima,eaw}@mit.edu}}
+
 \maketitle
 
-In this file, we analyze the case where the initial number of chips is finite.
-That is, the sum of all the chips on the board is finite.
+\section{Introduction}
 
+Written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.
+
+In this paper, we investigate the following problem set up:
+
+\vbox{
+\noindent
+\begin{quote}
+Each square of an infinite chess board contains some nonnegative number of poker chips.
+An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions.
+\end{quote}
+}
+We also consider extensions of this problem set up where the earthquake hits all squares on the board at once.
+We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
 \section{Standard Board}
+
+Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz
 

### 2014-09-25 draft0.tex
Edited Section 2
Fixed grammar and notation errors in Section 2.

@@
 We also consider extensions of this problem set up where the earthquake hits all squares on the board at once.
-We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
+We refer to the former type of earthquake as $\textbf{small earthquakes}$ and earthquakes that hit the entire board as $\textbf{big earthquakes}$.
 \section{Standard Board}
 
-Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz
+Written by Erik Waingarten and Perry Kleinhenz, edited by Fermi Ma
 
@@
 and leaves all other squares unchanged.
+
+We use the notation $SE^n(f, \{(x_i,y_i)\})$ to refer to the application of $SE$ to the board $f$, $n$ times to the squares $(x_1, y_1), \ldots (x_{n}, y_{n})$ respectively. 
 \end{definition}
@@
 \label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
+If the initial board has a finite number of chips, there exists an $N$ such that after $N$ big earthquakes or $N$ small earthquakes the resultant board will be stable.
 \end{theorem}
@@
 We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy.
-This shows that all finite boards
+This shows that all finite boards eventually become stable.
 
-So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $SE^{M}(f) = SE^{M+k}(f)$
+A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and $SE^{M}(f) = SE^{M+k}(f)$
 
@@
 \[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
-This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
+That is if big earthquake changes the state of the board, then $\Phi(f)$ increases, therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words that you never return to your original state.
 
@@
 \[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
-Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$.
-For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.
+
+Suppose $f(x_0,y_0) \geq 4$ for some location $(x_0, y_0)$, then the number of chips in location $(x_0,y_0)$ will have decreased by $4$ after applying $BE$. 
+For each decrease in the number of chips at location $(x_0,y_0)$ of $4$, the number of chips at the location $(y,z)$ increases by $1$, for each $(y,z) \in n(x_0,y_0)$.
 
@@
 So
-\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+\[ \Phi(BE(f)) - \Phi(f) = \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
+\[\geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+
 \[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
@@
 \[ \geq 4 > 0 \] 
+where the first inequality holds because at most two of the neighbors of $(x,y)$ are one unit closer to the origin then $(x,y)$ and at least two are one unit farther away and $(|x|+|y|+1)^2 > 2(|x|+|y|-1)^2$
 \end{proof}
@@
 \begin{lemma}
-If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
+If $SE(f, (x_1, y_1)) \neq f$, then $SE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
 \end{lemma}
@@
 \[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
-and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
+and evaluate $\Phi(SE(f, (x_1,y_1))) - \Phi(f)$:
 \begin{align}
-\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
-= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
-= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
+\Phi(SE(f), (x_1, y_1)) - \Phi(f) =\sum_{x,y} SE(f(x,y), (x_1, y_1))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
+= SE(f(x_1,y_1), (x_1, y_1))(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1,y_1)} SE(f(x,y), (x_1, y_1))(|x|+|y|)^2 \\- f(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1, y_1)} f(x,y)(|x|+|y|)^2 \\
+= 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2
 \end{align}
-And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
+And because at least two of the neighbors of $(x_1, y_1)$ are one unit farther from the origin than $(x_1, y_1)$ and at most two of the neighbors are one unit closer, and $(|x_1|+|y_1| +1)^2 > (|x_1|+|y_1| - 1)^2$, we can see,
 \begin{align}
-4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
-\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
+4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2  \\ 
+\geq 4(|x_1| + |y_1|)^2 + 2 (|x_1| + |y_1|+1)^2 + 2 (|x_1| + |y_1|-1)^2 = 4 >0
 \end{align}
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
-Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
+(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
+Then by Lemma~\ref{finiteextension}, we have that for all $n$ all chips on $SE^n(f)$ and $BE^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
 
-There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
+There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}_f$ that satisfy this.
+We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $SE^N(f) = SE^{N+m}(f)$ and $BE^N(f)=BE^{N+m}$ for all $m \geq 0$.
 \end{proof}
@@
 \section{$n$-Trees}
+Written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten
+
 Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.

### 2014-09-25 draft0.tex
added more sections, subsections, added written tags, started new section on acyclicity of infinite boards

--- 2014-09-25 draft0.tex
@@
 \usepackage{url}
+\setlength{\parindent}{0cm}
 \urlstyle{same}
@@
 
-Written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.
+\emph{Written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\
 
@@
 An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions.
+Does this process reach a stable state (with fewer than four chips on every square)?
+Is it independent of the order of redistribution?
+What can be said about the steady state?
 \end{quote}
@@
 We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
-\section{Standard Board}
+\section{Definitions and Notation}
 
-Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz
+\emph{Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz.}\\
 
@@
 
-\begin{definition} We define the big earthquake operator as a map from the set of all boards to the set of all boards and write $BE: \mathcal{B} \rightarrow \mathcal{B}$. 
+\begin{definition} We define the global earthquake operator as a map from the set of all boards to the set of all boards and write $GE: \mathcal{B} \rightarrow \mathcal{B}$. 
 It transforms all of the squares on the board with the following rule.
 \begin{equation}
-BE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
+GE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
 \end{equation}
@@
 \begin{definition}
- We define the small earthquake operator as a map from the set of all boards and locations to the set of all boards and write $SE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
-If $f(x,y) \leq 4$, then $SE(f, (x,y)) = f$, otherwise,  
+ We define the local earthquake operator as a map from the set of all boards and locations to the set of all boards and write $LE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
+If $f(x,y) \leq 4$, then $LE(f, (x,y)) = f$, otherwise,  
 \begin{align*}
-SE( f(x,y), (x,y)) = f(x,y)-4 \\
-SE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
-SE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
-SE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
-SE( f(x,y-1), (x,y)) = f(x,y-1)+1,
+LE( f(x,y), (x,y)) = f(x,y)-4 \\
+LE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
+LE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
+LE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
+LE( f(x,y-1), (x,y)) = f(x,y-1)+1,
 \end{align*}
@@
 
+\section{Stability of Finite Boards}
+\emph{Written by Erik Waingarten, Edited by Fermi Ma and Perry Kleinhenz.}
+
 \begin{theorem}
 \label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
+If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ global earthquakes or $M$ local earthquakes the resultant board will be stable.
 \end{theorem}
@@
 
-So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $SE^{M}(f) = SE^{M+k}(f)$
+So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $LE^{M}(f) = LE^{M+k}(f)$
 
+\subsection{A Potential Function}
 \begin{lemma}
-If $BE(f) \neq f$, then $BE^n(f) \neq f$ for all $n > 0$.
+If $GE(f) \neq f$, then $GE^n(f) \neq f$ for all $n > 0$.
 \end{lemma}
@@
 Note that this is well-defined since the boards contain a finite number of chips. We will show that 
-\[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
+\[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
 This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
 
-We can evaluate $\Phi(BE(f)) - \Phi(f)$.
-\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (|x|+|y|)^2 \]
+We can evaluate $\Phi(GE(f)) - \Phi(f)$.
+\[ \Phi(GE(f)) = \sum_{x,y} GE(f(x,y)) (|x|+|y|)^2 \]
 \[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
@@
 So
-\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+\[ \Phi(GE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
 \[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
@@
 \begin{lemma}
-If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
+If $LE(f) \neq f$, then $LE^n(f) \neq f$ for all $n > 0$.
 \end{lemma}
 Our proof of this lemma is essentiall the same as the preceeding one.
-We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.
+We construct a potential function and show that if a local earthquake acts nontrivially on a board then the value of the potential function increases.
 
@@
 \[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
-and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
+and evaluate $\Phi(LE(f)) - \Phi(f)$. Suppose $LE$ effected square $(x_0,y_0)$ and its neighbors, then
 \begin{align}
-\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
-= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
+\Phi(LE(f)) - \Phi(f) =\sum_{x,y} LE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
+= LE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} LE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
 = 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
@@
 
+\subsection{A Size Bound}
 \begin{lemma}
@@
 Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
+Then $GE^n(f)$ and $LE^n(f)$ only contain chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
 \end{lemma}
@@
 
+\subsection{Stability Result}
 So now we are ready to prove the theorem that all finite boards will reach a stable state.
@@
 
-\section{$n$-Trees}
+\section{Acyclicity of Infinite Boards}
+
+\emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\
+
+It turns out that a potential function argument can also show that infinite boards with a bounded number of chips on each square can never cycle.
+
+\section{Board Variants: $n$-Trees}
+\emph{written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten.}\\
+
 Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.
@@
 \label{bigearthquakedeftree}
-We define the big earthquake operator as  as a map from the set of all $n$ trees to the set of all $n$ trees and write $BE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
+We define the global earthquake operator as  as a map from the set of all $n$ trees to the set of all $n$ trees and write $GE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
 It transforms all of the nodes on the tree with the following rule.
 \begin{equation}
-BE_T(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
+GE_T(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
 \end{equation} 
@@
 \begin{definition}
- We define the small earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $SE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
+ We define the local earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $LE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
 It selects one valid node $(x,y)$ with $f(x,y) \geq n+1$ and transforms it and all of its neighbors with the following rule: 
 \begin{align*}
-SE_T( f( x, y ) ) = f( x , y )-n-1 \\
-SE_T( f( x + k, y)) = f( x + k, y )+1 \\
-SE_T( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
+LE_T( f( x, y ) ) = f( x , y )-n-1 \\
+LE_T( f( x + k, y)) = f( x + k, y )+1 \\
+LE_T( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
 \end{align*}
@@
 \begin{lemma}
-If $SE_T(f) \neq f$, then $SE_T^n(f) \neq f$ for all $n > 0$.
+If $LE_T(f) \neq f$, then $LE_T^n(f) \neq f$ for all $n > 0$.
 \end{lemma}
@@
 \begin{proof}
-We will show that  if $SE$ acts on a board nontrivially then $\Phi(SE(f))$ will be strictly larger. That is  
+We will show that  if $LE$ acts on a board nontrivially then $\Phi(LE(f))$ will be strictly larger. That is  
 \begin{equation}
-SE(f) \neq f \Rightarrow \Phi(SE(f)) > \Phi(f)
+LE(f) \neq f \Rightarrow \Phi(LE(f)) > \Phi(f)
 \end{equation}
 
-We can evaluate $\Phi(SE(f)) - \Phi(f)$.
+We can evaluate $\Phi(LE(f)) - \Phi(f)$.
 \begin{align}
-\Phi(SE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} SE_T(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
+\Phi(LE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} LE_T(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
 \end{align}
@@
 \begin{align*}
-\Phi(SE_T(f)-\Phi(f) = SE_T( f(x_0,y_0) )* (x_0)^2 + SE_T( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
-+ \sum_{i} SE_T( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
+\Phi(LE_T(f)-\Phi(f) = LE_T( f(x_0,y_0) )* (x_0)^2 + LE_T( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
++ \sum_{i} LE_T( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
 = -(n+1)*(|x_0|)^2 + (|x_0+k)^2 + n (|x_0-k|)^2 \\
@@
 \begin{lemma}
-If $BE_T(f) \neq f$, then $BE_T^n(f) \neq f$ for all $n > 0$.
+If $GE_T(f) \neq f$, then $GE_T^n(f) \neq f$ for all $n > 0$.
 \end{lemma}
@@
 \begin{equation}
-BE_T(f) \neq f \Rightarrow \Phi(BE_T(f)) > \Phi(f)
+GE_T(f) \neq f \Rightarrow \Phi(GE_T(f)) > \Phi(f)
 \end{equation}
@@
 
-We can evaluate $\Phi(BE_T(f)) - \Phi(f)$.
+We can evaluate $\Phi(GE_T(f)) - \Phi(f)$.
 \begin{align}
-\Phi(BE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} BE_T(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
+\Phi(GE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} GE_T(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
 \end{align}
@@
 \begin{align*}
-\Phi(BE_T(f)) - \Phi(f) =  \sum_{(x,y) \in A}  BE_T(f(x,y))x^2 + BE_T(f(x+k,y))(|x+k|)^2 + \\
-\sum_{i} \left( BE_T(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
+\Phi(GE_T(f)) - \Phi(f) =  \sum_{(x,y) \in A}  GE_T(f(x,y))x^2 + GE_T(f(x+k,y))(|x+k|)^2 + \\
+\sum_{i} \left( GE_T(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
 =\sum_{(x,y) \in A} -(n+1)x^2 + (|x+k|)^2 + n(|x-k|)^2 
@@
 Where $(x-k,y_i)$ refers to the $j$th child of $(x,y)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
-We note that each term in this sum is positive by our proof of the lemma for small earthquakes and so the sum itself must be positive.
+We note that each term in this sum is positive by our proof of the lemma for local earthquakes and so the sum itself must be positive.
 That is
 \begin{equation}
-\Phi(BE_T(f)) - \Phi(f) > 0
+\Phi(GE_T(f)) - \Phi(f) > 0
 \end{equation}
@@
 \label{finiteextensiontree}
-Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $BE_T^m(f)$ and $SE_T^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
+Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $GE_T^m(f)$ and $LE_T^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
 \end{lemma}
@@
 But this cannot occur because in order to move a chip down a level, one chip must move up a level.
-Therefore if $X_{min}=x_l$, both $BE_T^m(f)$ and $SE_T^m(f)$ do not contain chips below level $x_l -2S$.
+Therefore if $X_{min}=x_l$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not contain chips below level $x_l -2S$.
 
-An analogous argument shows that if $X_{max}=x_h$, both $BE_T^m(f)$ and $SE_T^m(f)$ do not contain chips above level $x_h+2S$.
+An analogous argument shows that if $X_{max}=x_h$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not contain chips above level $x_h+2S$.
 \end{proof}
@@
 (of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie between levels  $[x_l, x_h]$.
-Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $BE_T^m(f)$ and $SE_T^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$
+Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $GE_T^m(f)$ and $LE_T^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$
 
@@
 \begin{definition}
-We define a valid earthquake to be a small earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
+We define a valid earthquake to be a local earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
 
-We say that a sequence of small earthquakes $SE_1, SE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
+We say that a sequence of local earthquakes $LE_1, LE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
 \end{definition}
@@
 \begin{theorem}
-Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ small earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
+Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ local earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
 \end{theorem}
@@
 
-\begin{lemma} Let $SE_1$ and $SE_2$ be small earthquakes.
-If $SE_1, SE_2$ is a valid sequence and $SE_2$ is a valid earthquake then when before $SE_1$ then $SE_2, SE_1$ is also a valid sequence of earthquakes and $SE_2( SE_1 ( B ) ) = SE_1( SE_2 ( B ) )$
+\begin{lemma} Let $LE_1$ and $LE_2$ be local earthquakes.
+If $LE_1, LE_2$ is a valid sequence and $LE_2$ is a valid earthquake then when before $LE_1$ then $LE_2, LE_1$ is also a valid sequence of earthquakes and $LE_2( LE_1 ( B ) ) = LE_1( LE_2 ( B ) )$
 \end{lemma}
 \begin{proof}
-Suppose $SE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $SE_2$ affects the square $(x_2, y_2)$ and its neighbors.
-Suppose  $SE_2, SE_1$ was an invalid sequence, that is $SE_2(B)(x_1, y_1)<4$, but since $SE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$.
+Suppose $LE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $LE_2$ affects the square $(x_2, y_2)$ and its neighbors.
+Suppose  $LE_2, LE_1$ was an invalid sequence, that is $LE_2(B)(x_1, y_1)<4$, but since $LE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$.
 Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$.
-But then $SE_1(B)(x_1, y_1)=3$ which means $SE_1, SE_2$ would not be a valid sequence of earthquakes.
-Thus $SE_2, SE_1$ is a valid sequence.
+But then $LE_1(B)(x_1, y_1)=3$ which means $LE_1, LE_2$ would not be a valid sequence of earthquakes.
+Thus $LE_2, LE_1$ is a valid sequence.
 
-Now consider $SE_2( SE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the small earthquake we see that 
+Now consider $LE_2( LE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the local earthquake we see that 
 \begin{align*}
-SE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
-SE_2 ( SE_1 (B)) (u,v) = SE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
+LE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
+LE_2 ( LE_1 (B)) (u,v) = LE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
 = B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]]  \\
 = B(u,v) + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
-= SE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
-= SE_1 (SE_2 (B))(u,v) 
+= LE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
+= LE_1 (LE_2 (B))(u,v) 
 \end{align*}
-The critical step in this proof is recognizing that because $SE_1$ and $SE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
+The critical step in this proof is recognizing that because $LE_1$ and $LE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
 \end{proof}
@@
 \begin{lemma}
-Let $SE_1,SE_2,SE_3,\dots, SE_k$ be a sequence of earthquakes, where $SE_i$ hits the square $(x_i, y_i)$.
-Consider  the sequence given by moving some earthquake $SE_j$ to position $m<j$ in the sequence, such that $SE_j$ is a valid earthquake for all positions between $m$ and $j$.
+Let $LE_1,LE_2,LE_3,\dots, LE_k$ be a sequence of earthquakes, where $LE_i$ hits the square $(x_i, y_i)$.
+Consider  the sequence given by moving some earthquake $LE_j$ to position $m<j$ in the sequence, such that $LE_j$ is a valid earthquake for all positions between $m$ and $j$.
 This sequence is valid and produces the same final board configuration.
@@
 
-Therefore we must show that the sequence $SE_j, SE_m, SE_{m+1}, \ldots, SE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $SE_m, SE_{m+1}, \ldots, SE_{j-1}, SE_j$ applied to $B_{m-1}$.
+Therefore we must show that the sequence $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ applied to $B_{m-1}$.
 
@@
 \begin{equation*}
-SE_m, SE_{m+1}, \ldots, SE_{j-1}, SE_j
+LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j
 \end{equation*}
-where $p+1=j-m+1$. Then if we set $B_{m} = SE_{m} (B_{m-1})$ by applying our inductive step we have that the two sequences 
+where $p+1=j-m+1$. Then if we set $B_{m} = LE_{m} (B_{m-1})$ by applying our inductive step we have that the two sequences 
 \begin{align*}
-SE_{m+1}, \ldots, SE_{j-1}, SE_{j} \\ 
-SE_{j}, SE_{m+1}, \ldots,  SE_{j-1} 
+LE_{m+1}, \ldots, LE_{j-1}, LE_{j} \\ 
+LE_{j}, LE_{m+1}, \ldots,  LE_{j-1} 
 \end{align*}
@@
 \begin{align*}
-SE_{m}, SE_{j} \\
-SE_{j}, SE_{m}
+LE_{m}, LE_{j} \\
+LE_{j}, LE_{m}
 \end{align*}
 are both valid and produce the same final board configuration.
-If we apply the same reduction as above  we have  that $SE_j, SE_m, SE_{m+1}, \ldots, SE_{j-1}$  and $SE_m, SE_{m+1}, \ldots, SE_{j-1}, SE_j$ both applied to $B_{m-1}$ are valid and produce the same board.
+If we apply the same reduction as above  we have  that $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$  and $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ both applied to $B_{m-1}$ are valid and produce the same board.
 \end{proof}
@@
 
-written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz
+\emph{Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz.}
 
@@
 This is clearly true since the order of redistribution does not matter.
-In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with small earthquakes.
+In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with local earthquakes.
 

### 2014-09-25 draft0.tex
fixed more conflicts

@@
 }
+
 We also consider extensions of this problem set up where the earthquake hits all squares on the board at once.
@@
 and leaves all other squares unchanged.
+
+We use the notation $LE^n(f, \{(x_i,y_i)\})$ to refer to the application of $LE$ to the board $f$, $n$ times to the squares $(x_1, y_1), \ldots (x_{n}, y_{n})$ respectively. 
 \end{definition}
@@
 \label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ global earthquakes or $M$ local earthquakes the resultant board will be stable.
+If the initial board has a finite number of chips, there exists an $N$ such that after $N$ global earthquakes or $N$ local earthquakes the resultant board will be stable.
 \end{theorem}
@@
 We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy.
-This shows that all finite boards
+This shows that all finite boards eventually become stable.
 
-So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $LE^{M}(f) = LE^{M+k}(f)$
+A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and $LE^{M}(f) = LE^{M+k}(f)$
 
@@
 Note that this is well-defined since the boards contain a finite number of chips. We will show that
+
 \[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
-This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
+That is if global earthquake changes the state of the board, then $\Phi(f)$ increases, therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words that you never return to your original state.
 
@@
 \[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
-Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$.
-For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.
+
+Suppose $f(x_0,y_0) \geq 4$ for some location $(x_0, y_0)$, then the number of chips in location $(x_0,y_0)$ will have decreased by $4$ after applying $GE$. 
+For each decrease in the number of chips at location $(x_0,y_0)$ of $4$, the number of chips at the location $(y,z)$ increases by $1$, for each $(y,z) \in n(x_0,y_0)$.
 
@@
 So
-\[ \Phi(GE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+\[ \Phi(GE(f)) - \Phi(f) = \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
+\[\geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+
 \[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
@@
 \begin{lemma}
-If $LE(f) \neq f$, then $LE^n(f) \neq f$ for all $n > 0$.
+If $LE(f, (x_1, y_1)) \neq f$, then $LE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
 \end{lemma}
@@
 \[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
-and evaluate $\Phi(LE(f)) - \Phi(f)$. Suppose $LE$ effected square $(x_0,y_0)$ and its neighbors, then
+and evaluate $\Phi(LE(f, (x_1,y_1))) - \Phi(f)$:
 \begin{align}
-\Phi(LE(f)) - \Phi(f) =\sum_{x,y} LE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
-= LE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} LE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
-= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
+\Phi(LE(f), (x_1, y_1)) - \Phi(f) =\sum_{x,y} LE(f(x,y), (x_1, y_1))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
+= LE(f(x_1,y_1), (x_1, y_1))(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1,y_1)} LE(f(x,y), (x_1, y_1))(|x|+|y|)^2 \\- f(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1, y_1)} f(x,y)(|x|+|y|)^2 \\
+= 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2
 \end{align}
-And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
+And because at least two of the neighbors of $(x_1, y_1)$ are one unit farther from the origin than $(x_1, y_1)$ and at most two of the neighbors are one unit closer, and $(|x_1|+|y_1| +1)^2 > (|x_1|+|y_1| - 1)^2$, we can see,
 \begin{align}
-4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
-\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
+4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2  \\ 
+\geq 4(|x_1| + |y_1|)^2 + 2 (|x_1| + |y_1|+1)^2 + 2 (|x_1| + |y_1|-1)^2 = 4 >0
 \end{align}
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
-Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
+(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
+Then by Lemma~\ref{finiteextension}, we have that for all $n$ all chips on $LE^n(f)$ and $GE^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
 
-There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
+There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}_f$ that satisfy this.
+We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $LE^N(f) = LE^{N+m}(f)$ and $GE^N(f)=GE^{N+m}$ for all $m \geq 0$.
 \end{proof}

### 2014-09-25 draft0.tex
added intro paragraph for order of redistribution section

@@
 \section{Stability of Finite Boards}
+\label{Stability of Finite Boards}
 \emph{Written by Erik Waingarten, Edited by Fermi Ma and Perry Kleinhenz.}
@@
 
-written by Fermi Ma and Perry Kleinhenz, edited by Erik Waingarten.
+\emph{written by Fermi Ma and Perry Kleinhenz, edited by Erik Waingarten.}\\
+
+Here, we address the following question posed in the original problem statement: does the order of redistribution matter?
+As there is no order associated with global earthquakes, this question is naturally restricted to the case of local earthquakes.
+More specifically, we consider some finite starting board and some sequence of local earthquakes that hit the board until it finally stabilizes (note that our results from Section~\ref{Stability of Finite Boards} imply that such a sequence of earthquakes exist).
+We will find that the exact sequence of local earthquakes does not matter, and that if we keep hitting the board with local earthquakes anywhere they are allowed to happen, the board will always stabilize to the same final configuration.\\
+
+First, we formalize the notion of earthquakes that are ``allowed to happen".
 
@@
 We define a valid earthquake to be a local earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
-
 We say that a sequence of local earthquakes $LE_1, LE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.

### 2014-09-25 draft0.tex
added labels for every section
added how paper is organized in the introduction

@@
 
-In this paper, we investigate the following problem set up:
+In this paper, we investigate the following problem:
 
@@
 We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
+\\
+
+The paper is organized as follows.
+In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the game.
+In Section~\ref{Stability of Finite Boards}, we introduce the idea of a potential function defined on boards to prove that all boards with a finite number of starting chips eventually stabilize.
+We then modify the potential function to work for infinite boards and we explore its consequences in Section~\ref{Acyclicity of Infinite Boards}.
+In Section~\ref{Order of Redistribution}, we show that the order in which local earthquakes happens does not matter, thus answering a question in the original problem statement.
+We then shift our approach in Section ADDSECTIONNAME and present results from our simulations of repeated earthquakes.
+In particular, we look at the starting arrangement where $n$ chips all start on the same square, and we conjecture based on experimental evidence that such arragnements stabilize after $\Theta(n \log n)$ global earthquakes.
+In Section~\ref{Board Variants: n-Trees}, we consider what happens when alternative board setups are considered.
+Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the problem set up, and we show how to construct simple Boolean circuits with AND and OR gates.
 \section{Definitions and Notation}
+\label{Definitions and Notation}
 
@@
 \section{Acyclicity of Infinite Boards}
+\label{Acyclicity of Infinite Boards}
 
@@
 \section{Board Variants: $n$-Trees}
+\label{Board Variants: n-Trees}
 \emph{written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten.}\\
@@
 \section{Order of Redistribution}
+\label{Order of Redistribution}
 
@@
 \section{Computing with Chips}
+\label{Computing with Chips}
 

### 2014-09-25 draft0.tex
adding bounds until stable section

--- 2014-09-25 draft0.tex
@@
 \usepackage{amsmath}
+\usepackage{verbatim}
 \usepackage[margin=0.9in]{geometry}
@@
 
+\section{Time until Stability}
+
+We want to be able to bound the number of global earthquakes that it takes for a finite board to reach a stable state.
+We will first focus on the case of the board with only one stack containing $n$ chips.
+Without loss of generality, lets assume that the board contains $n$ chips at location $(0,0)$.
+We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes.
+We also let $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
+Its clear that $T(n) \geq T'(n)$.
+We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$.
+
+One can get another trivial upper bound for $T(n)$ by looking at the maximum value of the potential function.
+Remember that the board will only contain chips in the areas denotes by locations in $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
+So we know that when the board is stable at some $f'$,
+\[ \Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq 4n*3*4n = O(n^2) \]
+And since $\Phi(f) = 0$ and we know that $\Phi$ increases by at least $4$ in each global earthquake, 
+\[ T(n) = O(n^2) \]
+We can get tighter bounds by looking at the number of earthquakes more carefully.
+Lets let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize.
+Suppose $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (taking floors ofcourse).
+In particular,
+\[ T(x,y) = \left\{ \begin{array}{cc} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) \neq (0,0) \\
+ \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) = (0,0) \end{array} \right.
+\]
+So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
+\begin{figure}[!ht]
+\centering
+\subfigure[Global earthquakes until stable.
+Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
+\subfigure[Global earthquakes until stable.
+Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
+\caption{Number of global earthquakes until stability}
+\label{fig:growthofT}
+\end{figure}
+By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$.
+Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence, we believe that $T'(n) = \Theta(n\log n)$.
+
+\begin{figure}[!ht]
+\centering
+\subfigure[Local earthquakes of center until stable.
+Horizontal axis is $n$ and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
+\subfigure[Local earthquakes of center until stable.
+Horizontal axis is $n$ and vertical axis is the number of local earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
+\caption{Number of local earthquakes in center until stability}
+\label{fig:growthofT}
+\end{figure}
+
 \end{document}

### 2014-09-25 draft0.tex
fixed figure reference

@@
 By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$.
-Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence, we believe that $T'(n) = \Theta(n\log n)$.
+Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.
 
@@
 \caption{Number of local earthquakes in center until stability}
-\label{fig:growthofT}
+\label{fig:growthofT'}
 \end{figure}

### 2014-09-25 draft0.tex
adding small insight to model of computation

@@
 
+The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
+
 \subsection{Boolean circuits}
@@
 \end{proof}
+
+What does this mean?
+Well, in particular, it means that we cannot do much.
+In particular, we can't add, since $x + 1$ will act as the NOT operator on the lowest order bit.
+We can only compute boolean circuits in planar graphs.
 

### 2014-09-25 simulate.py
added simulations of gates

@@ -116,14 +116,14 @@ import Image,ImageDraw
 from random import randint as rint
 import pylab as pl
 
-def drawBoard(board, xrange, yrange, maxValue, recomputeMax=False):
+def drawBoard(board, xrange, yrange, maxValue, recomputeMax=False, showActive=False):
     if recomputeMax:
         maxValue = max(board.values())
     img = Image.new("RGB", (xrange, yrange), "#FFFFFF")
     draw = ImageDraw.Draw(img)
-    dr = 255./(maxValue/3)
-    dg = 255./(maxValue/3)
-    db = 255./(maxValue/3)
+    dr = 255./(maxValue/3. + 0.001)
+    dg = 255./(maxValue/3. + 0.001)
+    db = 255./(maxValue/3. + 0.001)
     for i in range(xrange):
         for j in range(yrange):
             try:
@@ -131,6 +131,8 @@ def drawBoard(board, xrange, yrange, maxValue, recomputeMax=False):
             except KeyError:
                 point = 0
             r, g, b = point*dr, 0, 0
+            if point >= 4 and showActive:
+                r, g, b = 0, 255, 0
             if r > 255:
                 r = 255
                 g = point*dg
@@ -140,15 +142,91 @@ def drawBoard(board, xrange, yrange, maxValue, recomputeMax=False):
             draw.point((i,j), fill=(int(r),int(g),int(b)))
     return img
 
-def showImage(board, xrange, yrange):
+def showImage(board, xrange, yrange, wait=False, showActive=False):
     maxValue = max(board.values())
-    img = drawBoard(board, xrange, yrange, maxValue)
+    img = drawBoard(board, xrange, yrange, maxValue, showActive=showActive)
     img = pl.imshow(img)
+    pl.pause(0.02)
+    pl.draw()
     changed = True
+    raw_input()
     while changed:
         changed = act(board)
-        img = drawBoard(board, xrange, yrange, maxValue, True)
+        img = drawBoard(board, xrange, yrange, maxValue, True, showActive=showActive)
         pl.imshow(img)
         pl.pause(0.02)
         pl.draw()
+        if wait:
+            raw_input()
     print "Done"
+
+def wire(x,y):
[…]

### 2014-09-26 draft0.tex
adding one more image

@@
 
-\title{Computing with an Angry Poker Player}
-\titlerunning{Computing with an Angry Poker Player}
+\title{Poker Chips, Earthquakes, and Computation}
+\titlerunning{Poker Chips, Earthquakes, and Computation}
 
@@
 In Section~\ref{Order of Redistribution}, we show that the order in which local earthquakes happens does not matter, thus answering a question in the original problem statement.
-We then shift our approach in Section ADDSECTIONNAME and present results from our simulations of repeated earthquakes.
+We then shift our approach in Section~\ref{Time until Stability} and present results from our simulations of repeated earthquakes.
 In particular, we look at the starting arrangement where $n$ chips all start on the same square, and we conjecture based on experimental evidence that such arragnements stabilize after $\Theta(n \log n)$ global earthquakes.
 In Section~\ref{Board Variants: n-Trees}, we consider what happens when alternative board setups are considered.
-Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the problem set up, and we show how to construct simple Boolean circuits with AND and OR gates.
+Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the problem set up, and show how to construct simple Boolean circuits and prove some impossibility results.
 \section{Definitions and Notation}
@@
 \section{Time until Stability}
+\label{Time until Stability}
 

### 2014-09-26 draft0.tex
rearranged section order
heavily revised the introduction

@@
 \usepackage{url}
-\setlength{\parindent}{0cm}
 \urlstyle{same}
@@
 
-In this paper, we investigate the following problem:
+\noindent In this paper, we investigate the following problem:
 
@@
 
-We also consider extensions of this problem set up where the earthquake hits all squares on the board at once.
-We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
-\\
+We also consider extensions of this problem where the earthquake hits all squares on the board at once.
+We will refer to earthquakes that hit only one square as $\emph{local earthquakes}$ and earthquakes that hit the entire board as $\emph{global earthquakes}$.
 
-The paper is organized as follows.
-In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the game.
-In Section~\ref{Stability of Finite Boards}, we introduce the idea of a potential function defined on boards to prove that all boards with a finite number of starting chips eventually stabilize.
-We then modify the potential function to work for infinite boards and we explore its consequences in Section~\ref{Acyclicity of Infinite Boards}.
+The paper is organized so that purely mathematical results addressing the original problem statement come first.
+In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the problem.
+In particular, we formalize the effects of local and global earthquakes, as well as what it means for a board of chips to be finite or infinite.
+In Section~\ref{Stability of Finite Boards}, we introduce a potential function that maps finite board configurations to $\mathbb{R}$.
+We then use the fact that local earthquakes strictly increase the value of this potential function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards stabilize (reach a steady state).
+In Section~\ref{Acyclicity of Infinite Boards}, we modify the potential function so that it converges to a real number even for infinite boards.
+From this, we conclude that infinite boards with bounded chip stack size never cycle back to previous configurations.
 In Section~\ref{Order of Redistribution}, we show that the order in which local earthquakes happens does not matter, thus answering a question in the original problem statement.
-We then shift our approach in Section~\ref{Time until Stability} and present results from our simulations of repeated earthquakes.
-In particular, we look at the starting arrangement where $n$ chips all start on the same square, and we conjecture based on experimental evidence that such arragnements stabilize after $\Theta(n \log n)$ global earthquakes.
-In Section~\ref{Board Variants: n-Trees}, we consider what happens when alternative board setups are considered.
-Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the problem set up, and show how to construct simple Boolean circuits and prove some impossibility results.
+
+For the remainder of the paper, we consider other approaches and extensions of the problem.
+In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes.
+In particular, we look at the starting arrangement where all of the board's $n$ chips are initially on one central square.
+We conjecture based on experimental evidence that such arragnements stabilize after $\Theta(n \log n)$ global earthquakes.
+In Section~\ref{Board Variants: n-Trees}, we generalize away from the chessboard set up, and consider what happens when a similar process happens on tree graphs of infinite size.
+Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits.
+We will also give impossibility results that show the limits of our model of computation.
 \section{Definitions and Notation}
@@
 It turns out that a potential function argument can also show that infinite boards with a bounded number of chips on each square can never cycle.
+
+\section{Order of Redistribution}
+\label{Order of Redistribution}
+
+\emph{written by Fermi Ma and Perry Kleinhenz, edited by Erik Waingarten.}\\
+
+Here, we address the following question posed in the original problem statement: does the order of redistribution matter?
+As there is no order associated with global earthquakes, this question is naturally restricted to the case of local earthquakes.
+More specifically, we consider some finite starting board and some sequence of local earthquakes that hit the board until it finally stabilizes (note that our results from Section~\ref{Stability of Finite Boards} imply that such a sequence of earthquakes exist).
+We will find that the exact sequence of local earthquakes does not matter, and that if we keep hitting the board with local earthquakes anywhere they are allowed to happen, the board will always stabilize to the same final configuration.\\
+
+First, we formalize the notion of earthquakes that are ``allowed to happen".
+
+\begin{definition}
+We define a valid earthquake to be a local earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
+We say that a sequence of local earthquakes $LE_1, LE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
+\end{definition}
+
+\begin{theorem}
+Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ local earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
+\end{theorem}
+
+\begin{proof}
+The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
+This technique will then specify an order in which the earthquakes will occur, which completes the proof.
+
+First, we need a lemma in order to do the rearrangements.
+
+\begin{lemma} Let $LE_1$ and $LE_2$ be local earthquakes.
+If $LE_1, LE_2$ is a valid sequence and $LE_2$ is a valid earthquake then when before $LE_1$ then $LE_2, LE_1$ is also a valid sequence of earthquakes and $LE_2( LE_1 ( B ) ) = LE_1( LE_2 ( B ) )$
+\end{lemma}
+\begin{proof}
+Suppose $LE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $LE_2$ affects the square $(x_2, y_2)$ and its neighbors.
+Suppose  $LE_2, LE_1$ was an invalid sequence, that is $LE_2(B)(x_1, y_1)<4$, but since $LE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$.
+Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$.
+But then $LE_1(B)(x_1, y_1)=3$ which means $LE_1, LE_2$ would not be a valid sequence of earthquakes.
+Thus $LE_2, LE_1$ is a valid sequence.
+
+Now consider $LE_2( LE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the local earthquake we see that 
+\begin{align*}
+LE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
+LE_2 ( LE_1 (B)) (u,v) = LE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
+= B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]]  \\
+= B(u,v) + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
+= LE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
+= LE_1 (LE_2 (B))(u,v) 
+\end{align*}
+The critical step in this proof is recognizing that because $LE_1$ and $LE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
+\end{proof}
+
+\begin{lemma}
+Let $LE_1,LE_2,LE_3,\dots, LE_k$ be a sequence of earthquakes, where $LE_i$ hits the square $(x_i, y_i)$.
+Consider  the sequence given by moving some earthquake $LE_j$ to position $m<j$ in the sequence, such that $LE_j$ is a valid earthquake for all positions between $m$ and $j$.
+This sequence is valid and produces the same final board configuration.
+\end{lemma}
+\begin{proof}
+It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes.
+In fact we can also ignore the first $m-1$ terms, because their order is not changed they will all be valid and the board state will be identical after $m-1$ terms for both sequences.
+Let us call this board state $B_{m-1}$
+
+Therefore we must show that the sequence $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ applied to $B_{m-1}$.
+
+We can show this by inducting on the length of the sequence $l=j-m+1$.
+The base case of $l=2$ is shown in the above lemma.
+Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$.
+So our sequence is
+\begin{equation*}
+LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j
+\end{equation*}
+where $p+1=j-m+1$. Then if we set $B_{m} = LE_{m} (B_{m-1})$ by applying our inductive step we have that the two sequences 
+\begin{align*}
+LE_{m+1}, \ldots, LE_{j-1}, LE_{j} \\ 
+LE_{j}, LE_{m+1}, \ldots,  LE_{j-1} 
+\end{align*}
+are both valid and produce the same final board configuration. Now if we consider the board state $B_{m-1}$ by our above Lemma we know that the two sequences 
+\begin{align*}
+LE_{m}, LE_{j} \\
+LE_{j}, LE_{m}
+\end{align*}
+are both valid and produce the same final board configuration.
+If we apply the same reduction as above  we have  that $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$  and $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ both applied to $B_{m-1}$ are valid and produce the same board.
+\end{proof}
+
+Given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
+We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes or else it will not eventually stabilize.
+Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquake events that correspond to square $i$.
+All these earthquakes can be moved to the front of the sequence.
+We do this for all squares $i$.
+By the lemma, the final positioning of the board does not change.
+
+The order of the earthquakes that have been moved to the front of the sequence can be in any order.
+This holds true because all these earthquakes could have happened with the initial values of the squares (maybe need to explain this better?
+but i think it's clear...).
+
+Now, find the point in the sequence that begins right after this point.
+Consider the board at this current point in time, and note that we have new values for all the squares.
+Then repeat this argument, moving a new set of earthquakes to the front.
+We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
+
+\end{proof}
+
+\section{Time until Stability}
+\label{Time until Stability}
+
+We want to be able to bound the number of global earthquakes that it takes for a finite board to reach a stable state.
+We will first focus on the case of the board with only one stack containing $n$ chips.
+Without loss of generality, lets assume that the board contains $n$ chips at location $(0,0)$.
+We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes.
+We also let $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
+Its clear that $T(n) \geq T'(n)$.
+We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$.
+
+One can get another trivial upper bound for $T(n)$ by looking at the maximum value of the potential function.
+Remember that the board will only contain chips in the areas denotes by locations in $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
+So we know that when the board is stable at some $f'$,
+\[ \Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq 4n*3*4n = O(n^2) \]
+And since $\Phi(f) = 0$ and we know that $\Phi$ increases by at least $4$ in each global earthquake, 
+\[ T(n) = O(n^2) \]
+We can get tighter bounds by looking at the number of earthquakes more carefully.
+Lets let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize.
+Suppose $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (taking floors ofcourse).
+In particular,
+\[ T(x,y) = \left\{ \begin{array}{cc} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) \neq (0,0) \\
+ \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) = (0,0) \end{array} \right.
+\]
+So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
+\begin{figure}[!ht]
+\centering
+\subfigure[Global earthquakes until stable.
+Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
+\subfigure[Global earthquakes until stable.
+Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
+\caption{Number of global earthquakes until stability}
+\label{fig:growthofT}
+\end{figure}
+By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$.
+Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.
+
+\begin{figure}[!ht]
+\centering
+\subfigure[Local earthquakes of center until stable.
+Horizontal axis is $n$ and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
+\subfigure[Local earthquakes of center until stable.
+Horizontal axis is $n$ and vertical axis is the number of local earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
+\caption{Number of local earthquakes in center until stability}
+\label{fig:growthofT'}
+\end{figure}
 
@@
 
-\section{Order of Redistribution}
-\label{Order of Redistribution}
-
-\emph{written by Fermi Ma and Perry Kleinhenz, edited by Erik Waingarten.}\\
-
-Here, we address the following question posed in the original problem statement: does the order of redistribution matter?
-As there is no order associated with global earthquakes, this question is naturally restricted to the case of local earthquakes.
-More specifically, we consider some finite starting board and some sequence of local earthquakes that hit the board until it finally stabilizes (note that our results from Section~\ref{Stability of Finite Boards} imply that such a sequence of earthquakes exist).
-We will find that the exact sequence of local earthquakes does not matter, and that if we keep hitting the board with local earthquakes anywhere they are allowed to happen, the board will always stabilize to the same final configuration.\\
-
-First, we formalize the notion of earthquakes that are ``allowed to happen".
-
-\begin{definition}
-We define a valid earthquake to be a local earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
-We say that a sequence of local earthquakes $LE_1, LE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
-\end{definition}
-
-\begin{theorem}
-Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ local earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
-\end{theorem}
-
-\begin{proof}
-The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
-This technique will then specify an order in which the earthquakes will occur, which completes the proof.
-
-First, we need a lemma in order to do the rearrangements.
-
-\begin{lemma} Let $LE_1$ and $LE_2$ be local earthquakes.
-If $LE_1, LE_2$ is a valid sequence and $LE_2$ is a valid earthquake then when before $LE_1$ then $LE_2, LE_1$ is also a valid sequence of earthquakes and $LE_2( LE_1 ( B ) ) = LE_1( LE_2 ( B ) )$
-\end{lemma}
-\begin{proof}
-Suppose $LE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $LE_2$ affects the square $(x_2, y_2)$ and its neighbors.
-Suppose  $LE_2, LE_1$ was an invalid sequence, that is $LE_2(B)(x_1, y_1)<4$, but since $LE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$.
-Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$.
-But then $LE_1(B)(x_1, y_1)=3$ which means $LE_1, LE_2$ would not be a valid sequence of earthquakes.
-Thus $LE_2, LE_1$ is a valid sequence.
-
-Now consider $LE_2( LE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the local earthquake we see that 
-\begin{align*}
-LE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
-LE_2 ( LE_1 (B)) (u,v) = LE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
-= B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]]  \\
-= B(u,v) + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
-= LE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
-= LE_1 (LE_2 (B))(u,v) 
-\end{align*}
-The critical step in this proof is recognizing that because $LE_1$ and $LE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
-\end{proof}
-
-\begin{lemma}
-Let $LE_1,LE_2,LE_3,\dots, LE_k$ be a sequence of earthquakes, where $LE_i$ hits the square $(x_i, y_i)$.
-Consider  the sequence given by moving some earthquake $LE_j$ to position $m<j$ in the sequence, such that $LE_j$ is a valid earthquake for all positions between $m$ and $j$.
-This sequence is valid and produces the same final board configuration.
-\end{lemma}
-\begin{proof}
-It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes.
-In fact we can also ignore the first $m-1$ terms, because their order is not changed they will all be valid and the board state will be identical after $m-1$ terms for both sequences.
-Let us call this board state $B_{m-1}$
-
-Therefore we must show that the sequence $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ applied to $B_{m-1}$.
-
-We can show this by inducting on the length of the sequence $l=j-m+1$.
-The base case of $l=2$ is shown in the above lemma.
-Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$.
-So our sequence is
-\begin{equation*}
-LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j
-\end{equation*}
-where $p+1=j-m+1$. Then if we set $B_{m} = LE_{m} (B_{m-1})$ by applying our inductive step we have that the two sequences 
-\begin{align*}
-LE_{m+1}, \ldots, LE_{j-1}, LE_{j} \\ 
-LE_{j}, LE_{m+1}, \ldots,  LE_{j-1} 
-\end{align*}
-are both valid and produce the same final board configuration. Now if we consider the board state $B_{m-1}$ by our above Lemma we know that the two sequences 
-\begin{align*}
-LE_{m}, LE_{j} \\
-LE_{j}, LE_{m}
-\end{align*}
-are both valid and produce the same final board configuration.
-If we apply the same reduction as above  we have  that $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$  and $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ both applied to $B_{m-1}$ are valid and produce the same board.
-\end{proof}
-
-Given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
-We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes or else it will not eventually stabilize.
-Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquake events that correspond to square $i$.
-All these earthquakes can be moved to the front of the sequence.
-We do this for all squares $i$.
-By the lemma, the final positioning of the board does not change.
-
-The order of the earthquakes that have been moved to the front of the sequence can be in any order.
-This holds true because all these earthquakes could have happened with the initial values of the squares (maybe need to explain this better?
-but i think it's clear...).
-
-Now, find the point in the sequence that begins right after this point.
-Consider the board at this current point in time, and note that we have new values for all the squares.
-Then repeat this argument, moving a new set of earthquakes to the front.
-We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
-
-\end{proof}
-
 \section{Computing with Chips}
@@
 
-\section{Time until Stability}
-\label{Time until Stability}
-
-We want to be able to bound the number of global earthquakes that it takes for a finite board to reach a stable state.
-We will first focus on the case of the board with only one stack containing $n$ chips.
-Without loss of generality, lets assume that the board contains $n$ chips at location $(0,0)$.
-We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes.
-We also let $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
-Its clear that $T(n) \geq T'(n)$.
-We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$.
-
-One can get another trivial upper bound for $T(n)$ by looking at the maximum value of the potential function.
-Remember that the board will only contain chips in the areas denotes by locations in $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
-So we know that when the board is stable at some $f'$,
-\[ \Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq 4n*3*4n = O(n^2) \]
-And since $\Phi(f) = 0$ and we know that $\Phi$ increases by at least $4$ in each global earthquake, 
-\[ T(n) = O(n^2) \]
-We can get tighter bounds by looking at the number of earthquakes more carefully.
-Lets let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize.
-Suppose $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (taking floors ofcourse).
-In particular,
-\[ T(x,y) = \left\{ \begin{array}{cc} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) \neq (0,0) \\
- \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) = (0,0) \end{array} \right.
-\]
-So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
-\begin{figure}[!ht]
-\centering
-\subfigure[Global earthquakes until stable.
-Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
-\subfigure[Global earthquakes until stable.
-Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
-\caption{Number of global earthquakes until stability}
-\label{fig:growthofT}
-\end{figure}
-By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$.
-Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.
-
-\begin{figure}[!ht]
-\centering
-\subfigure[Local earthquakes of center until stable.
-Horizontal axis is $n$ and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
-\subfigure[Local earthquakes of center until stable.
-Horizontal axis is $n$ and vertical axis is the number of local earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
-\caption{Number of local earthquakes in center until stability}
-\label{fig:growthofT'}
-\end{figure}
-
 \end{document}

### 2014-09-26 draft0.tex
edited some of the definitions and notation section

@@
 
-\begin{definition} A board is a map from the integer lattice to the non-negative integers.
-That is a Board $f$ is an element of $\mathcal{B} := \{ f: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}_{\geq 0} | f \text{ is a function }\}$.
-The input values of a board represent the coordinates on the integer lattice that the lower left hand corner of square occupies and the output value refers to the number of chips on that square. 
-A square is an element of $\mathbb{Z} \times \mathbb{Z}$. 
+First, we formalize the board of poker chips.
+We can overlay the grid on the $xy$ plane so that each grid square is a $1$ by $1$ square, and each grid intersection point is at a coordinate $(a,b)$ where $a,b \in \mathbb{Z}$.
+Then, the coordinate of a given grid square is just the coordinate of the grid intersection point of its lower left corner.
+
+\begin{definition} A board is a function $f: \mathbb{Z} \times \mathbb{Z} \to \mathbb{Z}_{\geq 0}$.
+The inputs to this function correspond to coordinates of grid squares and the output is the number of chips on that square. Let $\mathcal{B}$ denote the set of all boards.
 \end{definition}
+
+Recall from the introduction that local earthquakes (abbreviated as $LE$) are operations that hit one square with $\geq 4$ chips and redistributes 1 chip to each adjacent square.
+We define this operation in terms of $f$.
+
+\begin{definition}
+A local earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $LE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
+If $f(x,y) \geq 4$,
+\begin{align*}
+LE( f(x,y), (x,y)) = f(x,y)-4 \\
+LE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
+LE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
+LE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
+LE( f(x,y-1), (x,y)) = f(x,y-1)+1,
+\end{align*}
+Otherwise if $f(x,y) \leq 4$, then $LE(f(x,y)) = f$, as the earthquake does not change the board.
+\end{definition}
+We use the notation $LE^n(f, \{(x_i,y_i)\})$ to denote the application of $LE$ to the board $f$, $n$ times to the squares $(x_1, y_1), \ldots (x_{n}, y_{n})$ respectively.
+
+Recall that a global earthquake (abbreviated as $GE$) is equivalent to a local earthquake hitting every square on the board.
 
@@
 It transforms all of the squares on the board with the following rule.
-\begin{equation}
-GE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
-\end{equation}
+\begin{align*}
+GE(f(x,y)) = &f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + \\
+&[[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
+\end{align*}
 where $[[A]]$ is defined as 
@@
 \end{equation}
-\end{definition}
-
-\begin{definition}
- We define the local earthquake operator as a map from the set of all boards and locations to the set of all boards and write $LE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
-If $f(x,y) \leq 4$, then $LE(f, (x,y)) = f$, otherwise,  
-\begin{align*}
-LE( f(x,y), (x,y)) = f(x,y)-4 \\
-LE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
-LE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
-LE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
-LE( f(x,y-1), (x,y)) = f(x,y-1)+1,
-\end{align*}
-and leaves all other squares unchanged.
-
-We use the notation $LE^n(f, \{(x_i,y_i)\})$ to refer to the application of $LE$ to the board $f$, $n$ times to the squares $(x_1, y_1), \ldots (x_{n}, y_{n})$ respectively. 
 \end{definition}

### 2014-09-26 draft0.tex
added acyclicity of infinite board content

@@
 
+%%This section needs to be fixed up
+
 It turns out that a potential function argument can also show that infinite boards with a bounded number of chips on each square can never cycle.
+
+Consider the following potential function
+\[ \Phi(f) = \sum_{x,y}\frac{f(x,y)}{4(|x|+|y|)\cdot4^{(|x|+|y|)}} \]
+First, we note that for $n$, there are exactly $4n$ squares $(x,y)$ such that $|x| + |y| = n$. If we assume that the maximum chip stack is $M$, then the potential function can be bounded:
+\begin{align*}
+\sum_{x,y}\frac{f(x,y)}{4(|x|+|y|)\cdot4^{(|x|+|y|)}} \leq \sum_{n=1}^\infty \frac{M}{4^n} = \frac{M}{3}
+\end{align*}
+If any square at distance $d$ is hit by a local earthquake, the value of the potential at that square decreases by $\frac{4}{4d\cdot 4^d}$.
+The potential increases on the four adjacent squares.
+In particular we know that one of the adjacent squares is at distance $d-1$, and has a potential increase of $\frac{1}{4(d-1)\cdot 4^{d-1}} = \frac{4}{4(d-1)\cdot 4^d}$ which is larger in magnitude than $\frac{4}{4d\cdot 4^d}$.
+So the potential strictly increases after any local earthquake.
+
+It follows that infinite boards hit with local earthquakes can never cycle back to a previous arrangement
 

### 2014-09-26 draft0.tex
Added guiding Text
In section 2, in between definitions and at the start of section 3.

@@
 
+Now that we have formalized the earthquake operators we would also like to make a precise definition of stability of a board.
 \begin{definition}
@@
 
+In order to keep our notation compact we define a function that takes in the coordinate of a square and returns the coordinates of its neighbors in a set. 
 \begin{definition}
@@
 \end{equation}
-This function takes in a square and returns its neighbors. 
 \end{definition}
@@
 
+In this section we show that any board with a finite number of chips on it will reach a stable state after applying finitely many local or global earthquakes. 
+We do so by showing first that if a local or global earthquake changes the state of a board then no sequence of local or global earthquakes can transform the output board back into the original board.
+We will then show that for a given finite board there is a maximum number of states that local or global earthquakes can transform it into. 
+Together these show that all finite boards eventually become stable. 
 \begin{theorem}
@@
 
-We will first show that finite boards cannot return to a previous state.
-We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy.
-This shows that all finite boards eventually become stable.
-
-A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and $LE^{M}(f) = LE^{M+k}(f)$
+A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and $LE^{M}(f) = LE^{M+m}(f)$
 
 \subsection{A Potential Function}
+In order to show that if the state of a finite board cannot cycle under local or global earthquakes we define a potential function that maps board states to the positive Reals and show that applying either kind of earthquake to a board increases the value of the function.
 \begin{lemma}

### 2014-09-26 draft0.tex
Edited Potential Function Lemma for GE
Tried to increase the Clarity of the first lemma of section 3.1

@@
 \subsection{A Potential Function}
-In order to show that if the state of a finite board cannot cycle under local or global earthquakes we define a potential function that maps board states to the positive Reals and show that applying either kind of earthquake to a board increases the value of the function.
+In order to show that if the state of a finite board cannot cycle under local or global earthquakes we define a potential function that maps board states to the nonnegative integers and show that applying either kind of earthquake to a board increases the value of the function.
 \begin{lemma}
@@
 \[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
+we are taking a sum of the number of chips on each square with a weight of the taxi cab distance from the square to the origin.
 Note that this is well-defined since the boards contain a finite number of chips. We will show that
@@
 \[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
-That is if global earthquake changes the state of the board, then $\Phi(f)$ increases, therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words that you never return to your original state.
+That is if global earthquake changes the state of the board, then $\Phi(f)$ increases.
+Therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words applying a series of earthquakes to a board can never produce a previous board state.
 
-We can evaluate $\Phi(GE(f)) - \Phi(f)$.
-\[ \Phi(GE(f)) = \sum_{x,y} GE(f(x,y)) (|x|+|y|)^2 \]
-\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
+We can evaluate $\Phi(GE(f)) - \Phi(f),$ 
+\[ \Phi(GE(f)) - \Phi(f) = \sum_{ (x,y)} (GE(f(x,y)) -f(x,y))(|x|+|y|)^2 \]
 
-Suppose $f(x_0,y_0) \geq 4$ for some location $(x_0, y_0)$, then the number of chips in location $(x_0,y_0)$ will have decreased by $4$ after applying $GE$. 
-For each decrease in the number of chips at location $(x_0,y_0)$ of $4$, the number of chips at the location $(y,z)$ increases by $1$, for each $(y,z) \in n(x_0,y_0)$.
+Let $A$ be the set of locations $(x,y)$ with $f(x,y) \geq 4$, which is non-empty by assumption.
+We call this the set of active squares.
+Then we know that $GE(f(x,y)) -f(x,y) \neq 0$ if $(x,y) \in A \cup \bigcup_{(x,y) \in A} n(x,y)$.
+That is the contribution from square $(x,y)$ to the sum is nonzero only if $(x,y)$ is an active square or it neighbors an active square.
 
-Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
-So
-\[ \Phi(GE(f)) - \Phi(f) = \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
-\[\geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+For each $(x,y) \in A$ we know that four chips are removed from the square and one is placed on each of its four neighbors. Thus we have that 
+\[ \Phi(GE(f)) - \Phi(f)= \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
+That is
 
-\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
+Now since at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that
+\[\sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
+ \[ = \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
 									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
-									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
-\[ \geq 4 > 0 \] 
-where the first inequality holds because at most two of the neighbors of $(x,y)$ are one unit closer to the origin then $(x,y)$ and at least two are one unit farther away and $(|x|+|y|+1)^2 > 2(|x|+|y|-1)^2$
+									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} = \sum_{(x,y) \in A} 4> 0 \] 
 \end{proof}

### 2014-09-26 draft0.tex
edited section 2

@@
 We can overlay the grid on the $xy$ plane so that each grid square is a $1$ by $1$ square, and each grid intersection point is at a coordinate $(a,b)$ where $a,b \in \mathbb{Z}$.
-Then, the coordinate of a given grid square is just the coordinate of the grid intersection point of its lower left corner.
+The coordinate of a given grid square is the coordinate of the grid intersection point of its lower left corner.
 
@@
 
-Recall from the introduction that local earthquakes (abbreviated as $LE$) are operations that hit one square with $\geq 4$ chips and redistributes 1 chip to each adjacent square.
+Recall from the introduction that local earthquakes (abbreviated as $LE$) are operations that hit one square with more than $4$ chips and redistributes 1 chip to each cardinal neighbor.
 We define this operation in terms of $f$.
@@
 \begin{align*}
-LE( f(x,y), (x,y)) = f(x,y)-4 \\
-LE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
-LE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
-LE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
-LE( f(x,y-1), (x,y)) = f(x,y-1)+1,
+LE(f, (x,y))(x,y) = f(x,y)-4 \\
+LE( f, (x,y))(x+1, y) = f(x+1,y)+1 \\
+LE( f, (x,y))(x-1, y) = f(x-1,y)+1 \\
+LE( f, (x,y))(x, y+1) = f(x,y+1)+1 \\
+LE( f, (x,y))(x, y-1) = f(x,y-1)+1,
 \end{align*}
-Otherwise if $f(x,y) \leq 4$, then $LE(f(x,y)) = f$, as the earthquake does not change the board.
+Otherwise if $f(x,y) \leq 4$, then $LE(f) = f$, as the local earthquake does not move any chips.
 \end{definition}
@@
 
-Recall that a global earthquake (abbreviated as $GE$) is equivalent to a local earthquake hitting every square on the board.
+A global earthquake (abbreviated as $GE$) is similar to a local earthquake applied to every square on the board simultaneously.
 
@@
 \begin{align*}
-GE(f(x,y)) = &f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + \\
-&[[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
+GE(f) = \sum_{i, j \in \mathbb{Z}\times \mathbb{Z}} LE(f, (i,j))
 \end{align*}
-where $[[A]]$ is defined as 
-\begin{equation}
-[[A]] = 
-\begin{cases} 
-1 \text{ if } A \text{ is true} \\ 
-0 \text{ if } A  \text{ is false}
- \end{cases}.
-\end{equation}
 \end{definition}
@@
 \begin{definition}
-We say that a board is stable if its values are strictly less than 4. This is equivalent to both of the earthquake maps having no effect on the board. 
+We say that a board is stable if all squares contain values strictly less than 4. This is equivalent to saying both earthquakes (local and global) have no effect on the board. 
 \end{definition}

### 2014-09-26 draft0.tex
edited section 3.0

@@
 is finite. 
-We denote the set of finite boards as $\mathcal{B}_f$.
+We denote the set of finite boards as $\mathcal{B}_F$.
 \end{definition}
@@
 
-In this section we show that any board with a finite number of chips on it will reach a stable state after applying finitely many local or global earthquakes. 
-We do so by showing first that if a local or global earthquake changes the state of a board then no sequence of local or global earthquakes can transform the output board back into the original board.
-We will then show that for a given finite board there is a maximum number of states that local or global earthquakes can transform it into. 
-Together these show that all finite boards eventually become stable. 
+In this section we show that any finite board will reach a stable state after finitely many local or global earthquakes. 
+We do so by showing that if a local or global earthquake changes the state of a board then no sequence of local or global earthquakes can transform the output board back into the original board. 
+We will then bound the maximum number of configurations that local or global earthquakes can transform it into. 
+Together these show that all finite boards stabilize. 
 \begin{theorem}
 \label{finitestability}
-If the initial board has a finite number of chips, there exists an $N$ such that after $N$ global earthquakes or $N$ local earthquakes the resultant board will be stable.
+If $f \in B_F$, there exists $N, M > 0$ such that after $N$ global earthquakes or $M$ local earthquakes the resultant board will be stable, or for all $m > 0$, $GE^{N+m}(f) = GE^N(f)$, and $LE^{M+m}(f, X_1 \cup X_2) = LE^M(f, X_1)$.
 \end{theorem}

### 2014-09-26 draft0.tex
forgot to delete one thing
Edited Section 3 Lemma 3
Fixed notation and made more similar to Section 3 Lemma 2

@@
 
-A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and $LE^{M}(f) = LE^{M+m}(f)$
-
 \subsection{A Potential Function}
@@
 \[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
-and evaluate $\Phi(LE(f, (x_1,y_1))) - \Phi(f)$:
-\begin{align}
-\Phi(LE(f), (x_1, y_1)) - \Phi(f) =\sum_{x,y} LE(f(x,y), (x_1, y_1))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
-= LE(f(x_1,y_1), (x_1, y_1))(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1,y_1)} LE(f(x,y), (x_1, y_1))(|x|+|y|)^2 \\- f(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1, y_1)} f(x,y)(|x|+|y|)^2 \\
-= 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2
-\end{align}
-And because at least two of the neighbors of $(x_1, y_1)$ are one unit farther from the origin than $(x_1, y_1)$ and at most two of the neighbors are one unit closer, and $(|x_1|+|y_1| +1)^2 > (|x_1|+|y_1| - 1)^2$, we can see,
-\begin{align}
+and evaluate $\Phi(LE(f, (x_1,y_1))) - \Phi(f)$, recalling that $LE(f, (x_1, y_1))(x,y)$ refers to the value of  $f$ at $(x,y)$ after a Local earthquake effected $(x_1, y_1)$.
+\begin{align*}
+\Phi(LE(f), (x_1, y_1)) - \Phi(f) =\sum_{x,y} LE(f, (x_1, y_1))(x,y)(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
+= LE(f, (x_1, y_1))(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1,y_1)} LE(f, (x_1, y_1)) (x,y) (|x|+|y|)^2 \\- f(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1, y_1)} f(x,y)(|x|+|y|)^2 \\
+= 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2,
+\end{align*}
+Because at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2$ we have that
+\begin{align*}
 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2  \\ 
-\geq 4(|x_1| + |y_1|)^2 + 2 (|x_1| + |y_1|+1)^2 + 2 (|x_1| + |y_1|-1)^2 = 4 >0
-\end{align}
+\geq 4(|x_1| + |y_1|)^2 + 2 (|x_1| + |y_1|+1)^2 + 2 (|x_1| + |y_1|-1)^2 = 4 >0.
+\end{align*}
 \end{proof}

### 2014-09-26 draft0.tex
changed up to lemma 3

--- 2014-09-26 draft0.tex
@@
 
-A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and $LE^{M}(f) = LE^{M+m}(f)$
-
 \subsection{A Potential Function}
-In order to show that if the state of a finite board cannot cycle under local or global earthquakes we define a potential function that maps board states to the nonnegative integers and show that applying either kind of earthquake to a board increases the value of the function.
+In order to show finite boards do not cycle, we define a potential function mapping boards to nonnegative integers.
+We show that earthquakes increase the value of the function.
+We use the potential function to define a measure of how ``spread out" a board is.
+So we claim that spread always increases.
 \begin{lemma}
@@
 \begin{proof}
-Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
+Let $\Phi: \mathcal{B}_F \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
 \[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
-we are taking a sum of the number of chips on each square with a weight of the taxi cab distance from the square to the origin.
-Note that this is well-defined since the boards contain a finite number of chips. We will show that
-
+We can think of $\Phi(f)$ as the sum of the values weighted by the ``taxi-cab" distance from the origin squared. This is well-defined since boards are finite. We claim
 \[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
-That is if global earthquake changes the state of the board, then $\Phi(f)$ increases.
-Therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words applying a series of earthquakes to a board can never produce a previous board state.
+If a global earthquake changes the board, $\Phi$ increases.
+This implies if $GE(f) \neq f$, then for all $n > 0$, $GE^n(f) \neq f$.
+In other words, a series of earthquakes to a finite board can never produce a previous board.
 
 We can evaluate $\Phi(GE(f)) - \Phi(f),$ 
-\[ \Phi(GE(f)) - \Phi(f) = \sum_{ (x,y)} (GE(f(x,y)) -f(x,y))(|x|+|y|)^2 \]
+\[ \Phi(GE(f)) - \Phi(f) = \sum_{ (x,y)} (GE(f)(x,y) -f(x,y))(|x|+|y|)^2 \]
 
-Let $A$ be the set of locations $(x,y)$ with $f(x,y) \geq 4$, which is non-empty by assumption.
-We call this the set of active squares.
-Then we know that $GE(f(x,y)) -f(x,y) \neq 0$ if $(x,y) \in A \cup \bigcup_{(x,y) \in A} n(x,y)$.
-That is the contribution from square $(x,y)$ to the sum is nonzero only if $(x,y)$ is an active square or it neighbors an active square.
+Let $A$ be the set of locations $(x,y)$ with $f(x,y) \geq 4$.
+$A$ is non-empty by assumption.
+We call $A$ the set of active squares.
+We know $GE(f)(x,y) -f(x,y) \neq 0$ if $(x,y) \in A \cup \bigcup_{(x,y) \in A} n(x,y)$.
 
-For each $(x,y) \in A$ we know that four chips are removed from the square and one is placed on each of its four neighbors. Thus we have that 
+For each $(x,y) \in A$, four chips are removed from the square and one is placed on each of its four neighbors. Thus we have that 
 \[ \Phi(GE(f)) - \Phi(f)= \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
-That is
 
-Now since at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that
+Since at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that
 \[\sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
@@
 \begin{lemma}
-If $LE(f, (x_1, y_1)) \neq f$, then $LE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
+Suppose $f \in B_F$. If $LE(f, (x_1, y_1)) \neq f$, then $LE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
 \end{lemma}

### 2014-09-27 draft0.tex
Edited Section 3
Edited Subsection "A Size Bound" and "Stability Result"

@@
 \label{finiteextension}
-Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $GE^n(f)$ and $LE^n(f)$ only contain chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
+Suppose $\sum_{x,y} f(x,y) = S$, and $f$ is nonzero only in the range $[x_l, x_h] \times [y_l, y_h]$.
+Then $GE^n(f)$ and $LE^n(f)$ are nonzero only in the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
 \end{lemma}
@@
 \subsection{Stability Result}
-So now we are ready to prove the theorem that all finite boards will reach a stable state.
+We are now ready to prove the theorem that all finite boards will reach a stable state.
 
 \begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$.
-Then by Lemma~\ref{finiteextension}, we have that for all $n$ all chips on $LE^n(f)$ and $GE^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
+(of Theorem~\ref{finitestability}) Recall that $f \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = S$.
+Because $f$ is a finite board we know the locations with a nonzero number of chips are bounded.
+Without loss of generality we can say that $f$ is nonzero only in the range $[x_l, x_h] \times [y_l, y_h]$.
+Then by Lemma~\ref{finiteextension}, we have that for all $n$ $LE^n(f)$ and $GE^n(f)$ are nonzero only in the range $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
 
-There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}_f$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $LE^N(f) = LE^{N+m}(f)$ and $GE^N(f)=GE^{N+m}$ for all $m \geq 0$.
+There exists less than $\binom{R+S-1}{R-1}$ possible boards $b \in \mathcal{B}_f$ that satisfy this.
+We know that the state of the board cannot cycle, so there must be an $0 \leq N <\binom{R+S-1}{R-1}$ such that $LE^N(f) = LE^{N+m}(f)$ and $GE^N(f)=GE^{N+m}$ for all $m \geq 0$.
 \end{proof}

### 2014-09-27 draft0.tex
Edited Section "Acyclitiy"
Changed the denominator from 4(|x|+|y|) to 4(|x|+|y|+1) so that the function does not blow up for (0,0). Also added structure to make more similar to previous sections

@@
 
-%%This section needs to be fixed up
+We can in fact show that an infinite board with a bounded number of chips on each square cannot cycle under local earthquakes.
+To show this we will mirror the previous section by defining a potential function on boards and then showing that the functions value only increases after a local earthquake has occurred.
 
-It turns out that a potential function argument can also show that infinite boards with a bounded number of chips on each square can never cycle.
+We make the quick definition of the set of infinite boards
+\begin{definition}
+We say that a board is infinite if the total number of chips on it is finite. That is a board $f$ is infinite if the sum 
+\begin{equation*}
+ \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y)
+\end{equation*}
+does not converge.
+We denote the set of infinite boards as $\mathcal{B}_I$.
+We say that an infinite board $f \in \mathcal{B}_I$ is bounded if $\exists M \in \mathbb{Z}$ such that $f(x,y) \leq M$ for all $(x,y) \in \mathbb{Z} \times \mathbb{Z}$.
+We denote the set of bounded infinite boards as $\mathcal{B}_B$
+\end{definition}
 
-Consider the following potential function
-\[ \Phi(f) = \sum_{x,y}\frac{f(x,y)}{4(|x|+|y|)\cdot4^{(|x|+|y|)}} \]
-First, we note that for $n$, there are exactly $4n$ squares $(x,y)$ such that $|x| + |y| = n$. If we assume that the maximum chip stack is $M$, then the potential function can be bounded:
+We once again define a potential function, although this time it is from bounded infinite boards to the positive reals. \begin{definition} Let $\Phi: \mathcal{B}_B \rightarrow \mathbb{R}_{\geq 0}$
+\[ \Phi(f) = \sum_{(x,y)}\frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \]
+\end{definition}
+%include the +1 in the denominator so there is not a singularity at (0,0)
+Before it was obvious that our potential function converged and was well defined. However we are now considering boards for which the sum of all chips does not converge. Therefore we show 
+\begin{lemma} $\Phi(f)$ converges for all $f \in \mathcal{B}_B$.
+\end{lemma}
+\begin{proof}
+We note that for $n > 0$, there are exactly $4n$ squares $(x,y)$ such that $|x| + |y| = n$. If we assume that the maximum chip stack is $M$, then we have the bound:
 \begin{align*}
-\sum_{x,y}\frac{f(x,y)}{4(|x|+|y|)\cdot4^{(|x|+|y|)}} \leq \sum_{n=1}^\infty \frac{M}{4^n} = \frac{M}{3}
+\Phi(f) = \sum_{(x,y)}\frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \leq \sum_{n=1}^\infty \frac{M}{4^n} = \frac{M}{3}
 \end{align*}
-If any square at distance $d$ is hit by a local earthquake, the value of the potential at that square decreases by $\frac{4}{4d\cdot 4^d}$.
-The potential increases on the four adjacent squares.
-In particular we know that one of the adjacent squares is at distance $d-1$, and has a potential increase of $\frac{1}{4(d-1)\cdot 4^{d-1}} = \frac{4}{4(d-1)\cdot 4^d}$ which is larger in magnitude than $\frac{4}{4d\cdot 4^d}$.
-So the potential strictly increases after any local earthquake.
+Thus the sum in the potential function converges and the function is well defined. 
+\end{proof}
 
-It follows that infinite boards hit with local earthquakes can never cycle back to a previous arrangement
+\begin{theorem} 
+If $f \in \mathcal{B}_B$ and $LE(f, (x_0,y_0)) \neq f$, then $LE^{n}(f) \neq f$ for all $n>0$.
+\end{theorem}
+\begin{proof}
+We once again compute 
+\begin{align*}
+\Phi(LE(f))-\Phi(f) = \sum_{(x,y)} \frac{LE(f, (x_0,y_0))(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} - \frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \\
+=\frac{LE(f, (x_0,y_0))(x_0,y_0)-f(x_0,y_0)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} + \sum_{(x,y) \in n(x_0,y_0)} \frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} 
+\end{align*}
+Suppose $|x|+|y|=d$, then 
+\begin{equation*}
+\frac{LE(f,(x,y))(x,y)}{4(d+1)*4^d} -\frac{f(x,y)}{4(d+1)*4^d} =-\frac{4}{4(d+1)\cdot 4^d}
+\end{equation*}
+and since $LE(f, (x,y))(x_0,y_0)>f(x,y)$ for $(x,y) \in n(x_0,y_0)$ we have
+\begin{equation*}
+\frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} >0
+\end{equation*}
+Furthermore we know that one of the adjacent squares is at distance $d-1$, and that 
+\begin{equation*}
+\frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}}  = \frac{1}{4(d)\cdot 4^{d-1}} = \frac{4}{4(d)\cdot 4^d}
+\end{equation*}which is larger in magnitude than $\frac{4}{4(d+1)\cdot 4^d}$. Therefore the value of $\Phi$ strictly increases after any local earthquake.
+It follows that infinite boards hit with local earthquakes can never cycle back to a previous arrangement.
+\end{proof}
 

### 2014-09-27 draft0.tex
Edited Section "Time Until Stability"

@@
 \usepackage{subfigure}
+\usepackage{mathtools}
 %\usepackage{caption}
@@
 
-We want to be able to bound the number of global earthquakes that it takes for a finite board to reach a stable state.
+Another question of interest is how many global earthquakes it would take for a finite board to reach a stable state.
 We will first focus on the case of the board with only one stack containing $n$ chips.
-Without loss of generality, lets assume that the board contains $n$ chips at location $(0,0)$.
+
+Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
+\begin{definition}
 We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes.
-We also let $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
-Its clear that $T(n) \geq T'(n)$.
-We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$.
+We also define $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
+\end{definition}
+Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$, since that many earthquakes will reduce the number of chips on $(0,0)$ to less than $4$.
 
-One can get another trivial upper bound for $T(n)$ by looking at the maximum value of the potential function.
-Remember that the board will only contain chips in the areas denotes by locations in $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
+We can get another upper bound for $T(n)$ by looking at the maximum value of the potential function.
+Recall that the board will be nonzero only in locations in the range $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
 So we know that when the board is stable at some $f'$, 
-\[ \Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq 4n*3*4n = O(n^2) \]
-And since $\Phi(f) = 0$ and we know that $\Phi$ increases by at least $4$ in each global earthquake, 
+\begin{equation*}
+\Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n = 3*4n*4n= O(n^2)
+\end{equation*}
+We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
 \[ T(n) = O(n^2) \]
 We can get tighter bounds by looking at the number of earthquakes more carefully.
-Lets let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize.
-Suppose $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (taking floors ofcourse).
-In particular,
-\[ T(x,y) = \left\{ \begin{array}{cc} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) \neq (0,0) \\
- \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) = (0,0) \end{array} \right.
-\]
+\begin{definition}
+ Let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
+ \end{definition}
+ If $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (after taking floors).
+For $(0,0)$ the number of earthquakes that must hit the square equals the number of chips entering $(x,y)$ divided by 4 (after taking floors) plus the number of earthquakes needed to clear the original $n$ chips off the square.
+ Since a chip only enters a square when an earthquake hits an adjacent square, we have the following formula
+\begin{equation*}
+T(x,y) = \begin{dcases} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) \neq (0,0) \\
+ \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) = (0,0) .
+ \end{dcases}
+ \end{equation*}
 So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
@@
 \end{figure}
-By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$.
+By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$.
 Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.

### 2014-09-27 draft0.tex
revised a lot of small things, edited the circuits section to explain what boolean circuits are

@@
 
-Since at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that
-\[\sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
- \[ = \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
-									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
-									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} = \sum_{(x,y) \in A} 4> 0 \] 
+Since at most two of the neighbors of $(x,y)$ are a taxicab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxicab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that
+\begin{align*}
+& \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \\
+&\geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2+ 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \\
+&= \sum_{(x,y) \in A} -4x^2 -4y^2 -8xy +2x^2 +2y^2 +4xy +2x + \\
+&\hspace{1.7cm} 2y + 2+2x^2 +2y^2 +4xy -2x -2y +2\\
+&= \sum_{(x,y) \in A} 4\\
+&> 0
+\end{align*}
 \end{proof}
@@
 \end{align*}
-Because at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2$ we have that
+Because at most two of the neighbors of $(x,y)$ are a taxicab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxicab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2$ we have that
 \begin{align*}
@@
 
-In this section, we show how we can compute different boolean functions with chips.
-We will show how we can build circuits and prove some impossibility results for certain circuits.
-This might seem similar to computation with Game of Life; however, we note that these two are inherently different since Chips will always terminate with a finite board, whereas Game of Life might not.
+In this section, we show how we can compute different Boolean functions using the poker chip board and the earthquake operator.
+
+A Boolean input is a number in $\{0,1\}$, and a Boolean function is a function that takes $k$ Boolean inputs and outputs a number in $\{0,1\}$.
+A simple Boolean function is the AND function that takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are set to 1.
+Another Boolean function is the OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
+If 0 is associated with ``false" and 1 is associated with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
+More complicated Boolean functions can be made with more than 2 inputs and compositions of these inputs.
+Denoting an AND with $\vee$ an OR with $\wedge$, we can construct composite functions such as $b_1 \wedge (b_2 \vee b_3)$.
+
+Boolean circuits are formed by using wires that transmit 0's and 1's, and feeding them through various \emph{gates}.
+For example, a wire carrying a 0 and a wire carrying a 1 can be fed through an OR gate, which will ``output" a wire carrying a 1.
+In this section, we construct circuits using the poker chips board that use AND an OR gates.
 

### 2014-09-30 draft0.tex
Edited Intro and Start of Section 2
Still need to move global earthquake stuff around though

@@
 Is it independent of the order of redistribution?
-What can be said about the steady state?
+What can be said about the question when the underlying graph is different?
 \end{quote}
@@
 
-The paper is organized so that purely mathematical results addressing the original problem statement come first.
 In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the problem.
-In particular, we formalize the effects of local and global earthquakes, as well as what it means for a board of chips to be finite or infinite.
+In particular, we formalize the effects of local earthquakes, as well as what it means for a board of chips to be finite or infinite.
 In Section~\ref{Stability of Finite Boards}, we introduce a potential function that maps finite board configurations to $\mathbb{R}$. 
-We then use the fact that local earthquakes strictly increase the value of this potential function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards stabilize (reach a steady state).
-In Section~\ref{Acyclicity of Infinite Boards}, we modify the potential function so that it converges to a real number even for infinite boards.
-From this, we conclude that infinite boards with bounded chip stack size never cycle back to previous configurations.
-In Section~\ref{Order of Redistribution}, we show that the order in which local earthquakes happens does not matter, thus answering a question in the original problem statement.
+We then use the fact that local earthquakes strictly increase the value of this potential function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
+In Section~\ref{Acyclicity of Infinite Boards}, we modify the potential function so that it converges to a real number even for infinite boards, with bounded chip stack size. 
+From this, we conclude that such boards never cycle back to previous configurations. 
+In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a local earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
 
@@
 
-First, we formalize the board of poker chips.
-We can overlay the grid on the $xy$ plane so that each grid square is a $1$ by $1$ square, and each grid intersection point is at a coordinate $(a,b)$ where $a,b \in \mathbb{Z}$.
-The coordinate of a given grid square is the coordinate of the grid intersection point of its lower left corner.
+First, we formalize the board of poker chips. We consider the integer lattice $\mathbb{Z} \times \mathbb{Z}$ where each coordinate refers to an individual square.
 
-\begin{definition} A board is a function $f: \mathbb{Z} \times \mathbb{Z} \to \mathbb{Z}_{\geq 0}$.
+\begin{definition} A board is a function $b: \mathbb{Z} \times \mathbb{Z} \to \mathbb{Z}_{\geq 0}$.
 The inputs to this function correspond to coordinates of grid squares and the output is the number of chips on that square. Let $\mathcal{B}$ denote the set of all boards.
@@
 
-Recall from the introduction that local earthquakes (abbreviated as $LE$) are operations that hit one square with more than $4$ chips and redistributes 1 chip to each cardinal neighbor.
-We define this operation in terms of $f$.
+In order to keep our notation compact we define a function that takes in the coordinate of a square and returns the coordinates of its neighbors in a set. 
+\begin{definition}
+We define the function $n: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 $ as 
+\begin{equation}
+n(x,y) = \{ (x+1, y), (x-1, y), (x, y-1), (x, y+1) \}
+\end{equation}
+\end{definition}
+
+Recall from the introduction that local earthquakes (abbreviated as $E$) are operations that hit one square with more than $4$ chips and redistributes 1 chip to each neighbor.
+We define this operation in terms of $b$.
 
 \begin{definition}
-A local earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $LE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
-If $f(x,y) \geq 4$,
+A local earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
+If $b(x) \geq 4$, and $E$ effects square $x$ then
 \begin{align*}
-LE(f, (x,y))(x,y) = f(x,y)-4 \\
-LE( f, (x,y))(x+1, y) = f(x+1,y)+1 \\
-LE( f, (x,y))(x-1, y) = f(x-1,y)+1 \\
-LE( f, (x,y))(x, y+1) = f(x,y+1)+1 \\
-LE( f, (x,y))(x, y-1) = f(x,y-1)+1,
+LE[ b, x](x) = b(x)-4 \\
+LE[ b, x](y) = b(y)+1 \\
 \end{align*}
-Otherwise if $f(x,y) \leq 4$, then $LE(f) = f$, as the local earthquake does not move any chips.
+for $ y \in n(x)$. Otherwise  $b(x,y) < 4$, and $E[b] = b$, as the local earthquake does not move any chips.
 \end{definition}
-We use the notation $LE^n(f, \{(x_i,y_i)\})$ to denote the application of $LE$ to the board $f$, $n$ times to the squares $(x_1, y_1), \ldots (x_{n}, y_{n})$ respectively.
+We use the notation $E^n(f, \{x_i\})$ to denote the application of $E$ to the board $b$, $n$ times to the squares $x_1, \ldots x_{n}$ respectively.
 
+%move stuff below somewhere else
 A global earthquake (abbreviated as $GE$) is similar to a local earthquake applied to every square on the board simultaneously.
@@
 \end{definition}
+%move stuff above somewhere else
 
@@
 \begin{definition}
-We say that a board is stable if all squares contain values strictly less than 4. This is equivalent to saying both earthquakes (local and global) have no effect on the board. 
+We say that a board is stable if all squares contain values strictly less than 4. This is equivalent to saying a local earthquake has no effect on the board. 
 \end{definition}
@@
 \end{equation}
-is finite. 
+is finite. Otherwise we say a board is infinite. 
 We denote the set of finite boards as $\mathcal{B}_F$.
-\end{definition}
-
-In order to keep our notation compact we define a function that takes in the coordinate of a square and returns the coordinates of its neighbors in a set. 
-\begin{definition}
-We define the function $n: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 $ as 
-\begin{equation}
-n(x,y) = \{ (x+1, y), (x-1, y), (x, y-1), (x, y+1) \}
-\end{equation}
 \end{definition}

### 2014-10-06 draft0.tex
Edited Section 2 and 3
Still more work to be done on them. Also probably need to put global earthquake stuff in its own section.

@@
 In particular, we formalize the effects of local earthquakes, as well as what it means for a board of chips to be finite or infinite.
-In Section~\ref{Stability of Finite Boards}, we introduce a potential function that maps finite board configurations to $\mathbb{R}$. 
-We then use the fact that local earthquakes strictly increase the value of this potential function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
-In Section~\ref{Acyclicity of Infinite Boards}, we modify the potential function so that it converges to a real number even for infinite boards, with bounded chip stack size. 
+In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to $\mathbb{R}$. 
+We then use the fact that local earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
+In Section~\ref{Acyclicity of Infinite Boards}, we modify the spread function so that it converges to a real number even for infinite boards, with bounded chip stack size. 
 From this, we conclude that such boards never cycle back to previous configurations. 
@@
 In particular, we look at the starting arrangement where all of the board's $n$ chips are initially on one central square.
-We conjecture based on experimental evidence that such arragnements stabilize after $\Theta(n \log n)$ global earthquakes. 
+We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
 In Section~\ref{Board Variants: n-Trees}, we generalize away from the chessboard set up, and consider what happens when a similar process happens on tree graphs of infinite size. 
@@
 
-Recall from the introduction that local earthquakes (abbreviated as $E$) are operations that hit one square with more than $4$ chips and redistributes 1 chip to each neighbor.
+Recall from the introduction that an earthquake (abbreviated as $E$) is an operation that hits one square with more than $4$ chips and redistributes 1 chip to each neighbor.
 We define this operation in terms of $b$.
@@
 \begin{definition}
-A local earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
-If $b(x) \geq 4$, and $E$ effects square $x$ then
+An earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
+If $b(x) \geq 4$, then
 \begin{align*}
-LE[ b, x](x) = b(x)-4 \\
-LE[ b, x](y) = b(y)+1 \\
+E[ b, x](x) = b(x)-4 \\
+E[ b, x](y) = b(y)+1 \\
 \end{align*}
-for $ y \in n(x)$. Otherwise  $b(x,y) < 4$, and $E[b] = b$, as the local earthquake does not move any chips.
+for $ y \in n(x)$. Otherwise  $b(x,y) < 4$, and $E[b] = b$, as the earthquake does not move any chips.
+We use the notation $E^n(f, \{x_i\})$ to denote the application of $E$ to the board $b$, $n$ times to the squares $x_1, \ldots x_{n}$ respectively. 
 \end{definition}
-We use the notation $E^n(f, \{x_i\})$ to denote the application of $E$ to the board $b$, $n$ times to the squares $x_1, \ldots x_{n}$ respectively.
 
 %move stuff below somewhere else
-A global earthquake (abbreviated as $GE$) is similar to a local earthquake applied to every square on the board simultaneously.
+A global earthquake (abbreviated as $GE$) is similar to a standard earthquake applied to every square on the board simultaneously.
 
@@
 \begin{align*}
-GE(f) = \sum_{i, j \in \mathbb{Z}\times \mathbb{Z}} LE(f, (i,j))
+GE(f) = \sum_{i, j \in \mathbb{Z}\times \mathbb{Z}} E(f, (i,j))
 \end{align*}
@@
 
-Now that we have formalized the earthquake operators we would also like to make a precise definition of stability of a board.
+Now that we have formalized the earthquake operator we would like to make a precise definition of stability of a board.
 \begin{definition}
-We say that a board is stable if all squares contain values strictly less than 4. This is equivalent to saying a local earthquake has no effect on the board. 
+We say that a board is stable if all squares contain values strictly less than 4. 
 \end{definition}
+This is equivalent to saying an earthquake has no effect on the board. We would also like to formalize the definition of a finite board.
 
@@
 \begin{equation}
-S= \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y) 
+C= \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y) 
 \end{equation}
@@
 
-In this section we show that any finite board will reach a stable state after finitely many local or global earthquakes. 
-We do so by showing that if a local or global earthquake changes the state of a board then no sequence of local or global earthquakes can transform the output board back into the original board. 
-We will then bound the maximum number of configurations that local or global earthquakes can transform it into. 
+In this section we show that any finite board will reach a stable state after finitely many earthquakes.
+We do so by showing that if a local earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
+We will then give a bound for the maximum number of configurations that earthquakes can transform a given board into. 
 Together these show that all finite boards stabilize. 
@@
 \label{finitestability}
-If $f \in B_F$, there exists $N, M > 0$ such that after $N$ global earthquakes or $M$ local earthquakes the resultant board will be stable, or for all $m > 0$, $GE^{N+m}(f) = GE^N(f)$, and $LE^{M+m}(f, X_1 \cup X_2) = LE^M(f, X_1)$.
+If $f \in B_F$, there exists $N> 0$ such that after $N$ earthquakes the resultant board will be stable, or for all $m > 0$, $E^{N+m}[b] = E^N[b]$.
 \end{theorem}
 
-\subsection{A Potential Function}
-In order to show finite boards do not cycle, we define a potential function mapping boards to nonnegative integers.
-We show that earthquakes increase the value of the function.
-We use the potential function to define a measure of how ``spread out" a board is.
-So we claim that spread always increases.
+In order to show finite boards do not cycle, we define a spread function that maps boards to nonnegative integers.
+The spread function should give some measure of how ``spread out" the chips on a board are.
+Earthquakes seem to ``spread out" the chips on a board and so we would like earthquakes to increase the value of the function.
+
+\begin{lemma}
+Suppose $b \in B_F$. If $E[b, x] \neq b$, then $E^n[b, \{x_n\}] \neq b$ for all $n > 0$ and all sequences $\{x_n\}$.
+\end{lemma}
+
+\begin{proof}
+For ease of notation we use $x=(x_1, x_2)$. We define our spread function as $S: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
+\begin{equation*}
+S(f) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
+\end{equation*}
+We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared.
+This is well-defined since the total number of chips on the board is finite.
+We claim
+\begin{equation*}
+E[f, x] \neq f \Rightarrow S(E[f, x]) > S(f)
+\end{equation*}
+That is if an earthquake changes the board, $S$ increases.
+This implies if $E[b, x] \neq b$, then $E^n[b, \{x_n\}] \neq b$ for all $n > 0$ and all sequences $\{x_n\}$ .
+In other words, a series of earthquakes applied to a finite board can never produce a previous board.
+
+We evaluate $S(E[b, x]) - S(f)$, recalling that $E[b, x](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $x$.
+\begin{align*}
+&S(E[b, x]) - S(f) =\sum_{i,j} E[b, x](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
+&= E[b, x](x_1,x_2)(|x_1|+|x_2|)^2 + \sum_{(i,j) \in n(x_1,x_2)} E[b, x] (i,j) (|i|+|j|)^2 \\
+&- b(x_1, x_2)(|x_1|+|x_2|)^2 + \qquad \;  \; \sum_{(i,j) \in n(x_1, x_2)} b(i,j)(|i|+|j|)^2 \\
+&= 4(|x_1| + |x_2|)^2 +\qquad \qquad \quad \; \; \sum_{(i,j) \in n(x_1, x_2)} (|i| + |j|)^2,
+\end{align*}
+At most two of the neighbors of $(x_1, x_2))$ are a taxicab distance $(|x_1|+|x_2|-1)^2$ from the origin and at least two of the neighbors are a taxicab distance $(|x_1|+|x_2|+1)^2$ from the origin and $(|x_1|+|x_2|+1)^2 > (|x_1|+|x_2|-1)^2$.
+Therefore
+\begin{align*}
+&4(|x_1| + |x_2|)^2 +\sum_{(i,j) \in n(x_1, x_2)} (|i| + |j|)^2 \\
+& \geq 4(|x_1| + |x_2|)^2 + 2 (|x_1| + |x_2|+1)^2 + 2 (|x_1| + |x_2|-1)^2 = 4 >0.
+\end{align*}
+\end{proof}
+
+%Move stuff below somewhere else
 \begin{lemma}
@@
 \end{proof}
+%Move stuff above somewhere else
 
-\begin{lemma}
-Suppose $f \in B_F$. If $LE(f, (x_1, y_1)) \neq f$, then $LE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
-\end{lemma}
-Our proof of this lemma is essentiall the same as the preceeding one.
-We construct a potential function and show that if a local earthquake acts nontrivially on a board then the value of the potential function increases.
-
-\begin{proof}
-We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
-\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
-and evaluate $\Phi(LE(f, (x_1,y_1))) - \Phi(f)$, recalling that $LE(f, (x_1, y_1))(x,y)$ refers to the value of  $f$ at $(x,y)$ after a Local earthquake effected $(x_1, y_1)$.
-\begin{align*}
-\Phi(LE(f), (x_1, y_1)) - \Phi(f) =\sum_{x,y} LE(f, (x_1, y_1))(x,y)(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
-= LE(f, (x_1, y_1))(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1,y_1)} LE(f, (x_1, y_1)) (x,y) (|x|+|y|)^2 \\- f(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1, y_1)} f(x,y)(|x|+|y|)^2 \\
-= 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2,
-\end{align*}
-Because at most two of the neighbors of $(x,y)$ are a taxicab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxicab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2$ we have that
-\begin{align*}
-4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2  \\ 
-\geq 4(|x_1| + |y_1|)^2 + 2 (|x_1| + |y_1|+1)^2 + 2 (|x_1| + |y_1|-1)^2 = 4 >0.
-\end{align*}
-\end{proof}
-
-\subsection{A Size Bound}
+Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that any number of earthquakes can transform a given finite board into.
 \begin{lemma}
 \label{finiteextension}
-Suppose $\sum_{x,y} f(x,y) = S$, and $f$ is nonzero only in the range $[x_l, x_h] \times [y_l, y_h]$.
-Then $GE^n(f)$ and $LE^n(f)$ are nonzero only in the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
+Suppose $\sum_{x,y} b(x,y) = C$, and $f$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
+Then   $E^n[b, \{x_n\}$ is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$for all $n$ and for all sequences $\{x_n\}$ .
 \end{lemma}
@@
 
-\subsection{Stability Result}
 We are now ready to prove the theorem that all finite boards will reach a stable state.
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Recall that $f \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = S$.
-Because $f$ is a finite board we know the locations with a nonzero number of chips are bounded.
-Without loss of generality we can say that $f$ is nonzero only in the range $[x_l, x_h] \times [y_l, y_h]$.
-Then by Lemma~\ref{finiteextension}, we have that for all $n$ $LE^n(f)$ and $GE^n(f)$ are nonzero only in the range $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.
-
-There exists less than $\binom{R+S-1}{R-1}$ possible boards $b \in \mathcal{B}_f$ that satisfy this.
-We know that the state of the board cannot cycle, so there must be an $0 \leq N <\binom{R+S-1}{R-1}$ such that $LE^N(f) = LE^{N+m}(f)$ and $GE^N(f)=GE^{N+m}$ for all $m \geq 0$.
+(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = C$.
+Because $b$ is a finite board we know the locations with a nonzero number of chips are bounded.
+Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
+Then by Lemma~\ref{finiteextension}, we have that $E^n[f, \{x_n\}]$ is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ x_n \}$.
+If we set 
+\begin{equation*}
+D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
+\end{equation*}
+then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,.
+We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{x_n\}$ such that $E^N[ f, \{x_n\}] = LE^{N+m}[f, \{x_n\}]$.
 \end{proof}

### 2014-10-09 draft0.tex
Edited first 7 sections
Moved Global earthquakes to its own section and provided an alternative proof (which I think is better) for why global earthquakes stabilize a finite board. 

I've made the change to notation that Susan suggested for earthquakes so they now look like E[b, s](i,j)

@@
 \begin{quote}
-Each square of an infinite chess board contains some nonnegative number of poker chips.
+Each node of an infinite chess board contains some nonnegative number of poker chips.
 An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions.
@@
 \begin{definition} A board is a function $b: \mathbb{Z} \times \mathbb{Z} \to \mathbb{Z}_{\geq 0}$.
-The inputs to this function correspond to coordinates of grid squares and the output is the number of chips on that square. Let $\mathcal{B}$ denote the set of all boards.
+The input to this function correspond to the coordinates of a square and the output is the number of chips on that square. Let $\mathcal{B}$ denote the set of all boards.
 \end{definition}
@@
 \begin{equation}
-n(x,y) = \{ (x+1, y), (x-1, y), (x, y-1), (x, y+1) \}
+n(i,j) = \{ (i+1, j), (i-1, j), (i, j-1), (i, j+1) \}.
 \end{equation}
@@
 An earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
-If $b(x) \geq 4$, then
+If $b(s) \geq 4$, then
 \begin{align*}
-E[ b, x](x) = b(x)-4 \\
-E[ b, x](y) = b(y)+1 \\
+E[b, s](s) &= b(s)-4 \\
+E[b, s](t) &= b(t)+1, \\
 \end{align*}
-for $ y \in n(x)$. Otherwise  $b(x,y) < 4$, and $E[b] = b$, as the earthquake does not move any chips.
-We use the notation $E^n(f, \{x_i\})$ to denote the application of $E$ to the board $b$, $n$ times to the squares $x_1, \ldots x_{n}$ respectively. 
+for $ t \in n(s)$. Otherwise  $b(s) < 4$, and $E[b, s] = b$, as the earthquake does not move any chips.
+We use the notation $E^n[b, \{s_k\}]$ to denote the application of $E$ to the board $b$ $n$ times, to the squares $s_1, \ldots s_{n}$ respectively. 
 \end{definition}
-
-%move stuff below somewhere else
-A global earthquake (abbreviated as $GE$) is similar to a standard earthquake applied to every square on the board simultaneously.
-
-\begin{definition} We define the global earthquake operator as a map from the set of all boards to the set of all boards and write $GE: \mathcal{B} \rightarrow \mathcal{B}$. 
-It transforms all of the squares on the board with the following rule.
-\begin{align*}
-GE(f) = \sum_{i, j \in \mathbb{Z}\times \mathbb{Z}} E(f, (i,j))
-\end{align*}
-\end{definition}
-%move stuff above somewhere else
 
@@
 \end{definition}
-This is equivalent to saying an earthquake has no effect on the board. We would also like to formalize the definition of a finite board.
+This is equivalent to saying an earthquake has no effect on the board. We would also to have a formal definition for a finite board.
 
 \begin{definition} 
-We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
+We say that a board is finite if the total number of chips on it is finite. That is a board $b$ is finite if the sum
 \begin{equation}
-C= \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y) 
+C= \sum_{(i,j) \in \mathbb{Z} \times \mathbb{Z}} b(i,j), 
 \end{equation}
@@
 
-In order to show finite boards do not cycle, we define a spread function that maps boards to nonnegative integers.
+In order to show finite boards never revisit a prior arrangement, we define a spread function that maps boards to nonnegative integers.
 The spread function should give some measure of how ``spread out" the chips on a board are.
@@
 \begin{proof}
-For ease of notation we use $x=(x_1, x_2)$. We define our spread function as $S: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
+For ease of notation we use $s=(s_1, s_2)$. We define our spread function as $S: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
 \begin{equation*}
-S(f) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
+\Psi(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
 \end{equation*}
@@
 \begin{equation*}
-E[f, x] \neq f \Rightarrow S(E[f, x]) > S(f)
+E[f, s] \neq f \Rightarrow \Psi(E[f, s]) > S(f)
 \end{equation*}
-That is if an earthquake changes the board, $S$ increases.
-This implies if $E[b, x] \neq b$, then $E^n[b, \{x_n\}] \neq b$ for all $n > 0$ and all sequences $\{x_n\}$ .
-In other words, a series of earthquakes applied to a finite board can never produce a previous board.
+That is if an earthquake changes the board, $\Psi$ increases.
+This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.
 
-We evaluate $S(E[b, x]) - S(f)$, recalling that $E[b, x](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $x$.
+We evaluate $\Psi(E[b, s]) - \Psi(b)$, recalling that $E[b, s](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $x$.
 \begin{align*}
-&S(E[b, x]) - S(f) =\sum_{i,j} E[b, x](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
-&= E[b, x](x_1,x_2)(|x_1|+|x_2|)^2 + \sum_{(i,j) \in n(x_1,x_2)} E[b, x] (i,j) (|i|+|j|)^2 \\
-&- b(x_1, x_2)(|x_1|+|x_2|)^2 + \qquad \;  \; \sum_{(i,j) \in n(x_1, x_2)} b(i,j)(|i|+|j|)^2 \\
-&= 4(|x_1| + |x_2|)^2 +\qquad \qquad \quad \; \; \sum_{(i,j) \in n(x_1, x_2)} (|i| + |j|)^2,
+&\Psi(E[b, s]) - \Psi(b) =\sum_{i,j} E[b, s](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
+&= E[b, s](s_1,s_2)(|s_1|+|s_2|)^2 + \sum_{(i,j) \in n(s)} E[b, s] (i,j) (|i|+|j|)^2 \\
+&- b(s_1, s_2)(|s_1|+|s_2|)^2 + \qquad \;  \; \sum_{(i,j) \in n(s)} b(i,j)(|i|+|j|)^2 \\
+&= -4(|s_1| + |s_2|)^2 +\qquad \qquad \quad \; \; \sum_{(i,j) \in n(s)} (|i| + |j|)^2,
 \end{align*}
-At most two of the neighbors of $(x_1, x_2))$ are a taxicab distance $(|x_1|+|x_2|-1)^2$ from the origin and at least two of the neighbors are a taxicab distance $(|x_1|+|x_2|+1)^2$ from the origin and $(|x_1|+|x_2|+1)^2 > (|x_1|+|x_2|-1)^2$.
-Therefore
+For most two of the neighbors of $(s_1, s_2))$ 
+\begin{equation*}
+(|i|+|j|)= (|s_1|+|s_2|-1)^2,
+\end{equation*}
+and for at least two of the neighbors 
+\begin{equation*}
+(|i|+|j|)=(|s_1|+|s_2|+1)^2,
+\end{equation*}
+ and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
 \begin{align*}
-&4(|x_1| + |x_2|)^2 +\sum_{(i,j) \in n(x_1, x_2)} (|i| + |j|)^2 \\
-& \geq 4(|x_1| + |x_2|)^2 + 2 (|x_1| + |x_2|+1)^2 + 2 (|x_1| + |x_2|-1)^2 = 4 >0.
+&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
+& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
 \end{align*}
 \end{proof}
-
-%Move stuff below somewhere else
-\begin{lemma}
-If $GE(f) \neq f$, then $GE^n(f) \neq f$ for all $n > 0$.
-\end{lemma}
-
-\begin{proof}
-Let $\Phi: \mathcal{B}_F \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
-\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
-We can think of $\Phi(f)$ as the sum of the values weighted by the ``taxi-cab" distance from the origin squared. This is well-defined since boards are finite. We claim
-\[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
-If a global earthquake changes the board, $\Phi$ increases.
-This implies if $GE(f) \neq f$, then for all $n > 0$, $GE^n(f) \neq f$.
-In other words, a series of earthquakes to a finite board can never produce a previous board.
-
-We can evaluate $\Phi(GE(f)) - \Phi(f),$ 
-\[ \Phi(GE(f)) - \Phi(f) = \sum_{ (x,y)} (GE(f)(x,y) -f(x,y))(|x|+|y|)^2 \]
-
-Let $A$ be the set of locations $(x,y)$ with $f(x,y) \geq 4$.
-$A$ is non-empty by assumption.
-We call $A$ the set of active squares.
-We know $GE(f)(x,y) -f(x,y) \neq 0$ if $(x,y) \in A \cup \bigcup_{(x,y) \in A} n(x,y)$.
-
-For each $(x,y) \in A$, four chips are removed from the square and one is placed on each of its four neighbors. Thus we have that 
-\[ \Phi(GE(f)) - \Phi(f)= \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
-
-Since at most two of the neighbors of $(x,y)$ are a taxicab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxicab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that
-\begin{align*}
-& \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \\
-&\geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2+ 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \\
-&= \sum_{(x,y) \in A} -4x^2 -4y^2 -8xy +2x^2 +2y^2 +4xy +2x + \\
-&\hspace{1.7cm} 2y + 2+2x^2 +2y^2 +4xy -2x -2y +2\\
-&= \sum_{(x,y) \in A} 4\\
-&> 0
-\end{align*}
-\end{proof}
-%Move stuff above somewhere else
 
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = C$.
+(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_f$, so $\sum_{i,j} b(i,j) = C$.
 Because $b$ is a finite board we know the locations with a nonzero number of chips are bounded.
 Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then by Lemma~\ref{finiteextension}, we have that $E^n[f, \{x_n\}]$ is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ x_n \}$.
+Then by Lemma~\ref{finiteextension}, we have that $E^n[f, \{s_kj \}]$ is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ s_k \}$ of length $n$.
 If we set 
@@
 then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,.
-We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{x_n\}$ such that $E^N[ f, \{x_n\}] = LE^{N+m}[f, \{x_n\}]$.
+We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{s_k\}$, which is arbitrary after the first $N$ terms, such that $E^N[ f, \{s_k\}] = LE^{N+m}[f, \{s_k\}]$.
 \end{proof}
@@
 
+\end{proof}
+
+\section{Global Earthquakes}
+\label{Global Earthquakes}
+\emph{written by Perry Kleinhenz, edited by Erik Waingarten and Fermi Ma}
+Now that we have shown that the order in which earthquakes occur does not matter we might wonder if the behavior of the system changes when all possible earthquakes occur at the \emph{same} time.
+In this section we will formalize this notion and show that all finite boards stabilize under such earthquakes as well.
+
+We will call these earthquakes global earthquakes. A global earthquake (abbreviated as $GE$) is esentially to a standard earthquake applied to every square on the board simultaneously.
+\begin{definition} We define the global earthquake operator as a map from the set of all boards to the set of all boards and write $GE: \mathcal{B} \rightarrow \mathcal{B}$. 
+It transforms all of the squares on the board with the following rule
+\begin{align*}
+GE[b](i,j) = b(i,j) +\sum_{s \in \mathbb{Z} \times \mathbb{Z}} (E[b, s](i,j) - b(i,j)).
+\end{align*}
+That is the value of the board at a square after applying a global earthquake is equal to the value of the board originally plus the change in the value at the square that would occur from standard earthquakes affecting every square.
+\end{definition}
+
+In order to prove that finite boards stabilize under global earthquakes we will first show that global earthquakes are equivalent to some finite sequence of standard earthquakes.
+Before doing so we will introduce a definition to refer to the squares a global earthquake can effect.
+
+\begin{definition} The active sites $A_b$ of a board $b$ are all those squares which have at least 4 chips on them or written symbolically
+\begin{definition}
+A_b=\{ (i,j); b(i,j) \geq 4 \}
+\end{equation*}
+\end{definition}
+
+We now show that a global earthquake is equivalent to applying standard earthquakes to each of the active sites on a board in any order.
+\begin{lemma} Suppose $b$ is a finite board. If we let arrange the terms of $A_b$ into a sequence $\{a_k}$ with $N$ elements then 
+\begin{equation*}
+GE[b] = E^N[b, \{a_k\}]
+\end{equation*}
+\end{llemma}
+\begin{proof} 
+A standard earthquake is nontrivial only if the square it targets is contained in $A_b$. That is $E[b, s] = b$  if $s \notin A_b$. Therefore $GE[b](i,j)$ becomes
+\begin{equation*}
+GE[b](i,j) = b(i,j) +\sum_{s \in A_b} (E[b, s](i,j) - b(i,j)).
+\end{equation*}
+So a global earthquake can be considered finitely many earthquakes occurring at the same time.
+But now if we consider $E^N[b, \{a_k\}]$, because the order of earthquakes does not matter the change in the value at some $(i,j)$ between the application of the $m+1$th and $m$th earthquake is equal to the change in the value at $(i,j)$ if the $m+1$th earthquake was applied directly to the original board.
+That is
+\begin{equation*}
+E^{m+1}[b, \{a_k\}](i,j) - E^{m}[b, \{a_k\}](i,j) = E[b, a_{m+1} ](i,j) - b(i,j)
+\end{equation*}
+Therefore 
+\begin{equation*}
+E^N[b, \{a_k\}](i,j) = b(i,j) + \sum_{a_k \in A_b} (E[b, a_k](i,j) - b(i,j)),
+\end{equation*}
+but this is the same expression as the value at $(i,j)$ after a global earthquake effects the board $b$. 
+\end{proof}
+
+\begin{myth} Suppose $b$ is a finite board, then there exists $N$ such that for all $m>0$ 
+\begin{equation*}
+GE^{N+m}[b] = GE^{N}[b]
+\end{equation*}
+\end{myth}
+\begin{proof}
+Assume otherwise.
+That is for some board $b, GE^N[b]$ always has at least one square with at least four chips on it.
+But applying a global earthquake is equivalent to applying a sequence of standard earthquakes and so we have a board which does not stabilize under standard earthquakes which is a contradiction.
 \end{proof}

### 2014-10-10 draft0.tex
Continued to make edits from Susan
Up through section 7. Order of earthquakes still needs some work.

@@
 \label{finiteextension}
-Suppose $\sum_{x,y} b(x,y) = C$, and $f$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then   $E^n[b, \{x_n\}$ is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$for all $n$ and for all sequences $\{x_n\}$ .
+Suppose $\sum_{i,j} b(i,j) = C$, and $f$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
+Then   $E^n[b, \{s_k\}]$ is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$for all $n$ and for all sequences $\{s_k\}$ .
 \end{lemma}
@@
 \begin{proof}
+%This proof needs to get fixed
 The most compact stable state of a board with $S$ chips is to have $3$ chips on each square adjacent to each other.
@@
 \emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\
-
-We can in fact show that an infinite board with a bounded number of chips on each square cannot cycle under local earthquakes.
+%Change this to reflect the proof given in the talk
+We can in fact show that infinite boards cannot cycle under earthquakes.
 To show this we will mirror the previous section by defining a potential function on boards and then showing that the functions value only increases after a local earthquake has occurred.
@@
 \begin{definition}
-We say that a board is infinite if the total number of chips on it is finite. That is a board $f$ is infinite if the sum 
+We say that a board is infinite if the total number of chips on it is infinite. That is a board $f$ is infinite if the sum 
 \begin{equation*}
- \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y)
+ \sum_{(i,j) \in \mathbb{Z} \times \mathbb{Z}} b(x,y)
 \end{equation*}
 does not converge.
-We denote the set of infinite boards as $\mathcal{B}_I$.
-We say that an infinite board $f \in \mathcal{B}_I$ is bounded if $\exists M \in \mathbb{Z}$ such that $f(x,y) \leq M$ for all $(x,y) \in \mathbb{Z} \times \mathbb{Z}$.
+We say that an infinite board $b$ is bounded if there exists $M \in \mathbb{Z}$ such that $b(i,j) \leq M$ for all $(i,j) \in \mathbb{Z} \times \mathbb{Z}$.
 We denote the set of bounded infinite boards as $\mathcal{B}_B$
@@
 
-Here, we address the following question posed in the original problem statement: does the order of redistribution matter?
-As there is no order associated with global earthquakes, this question is naturally restricted to the case of local earthquakes.
-More specifically, we consider some finite starting board and some sequence of local earthquakes that hit the board until it finally stabilizes (note that our results from Section~\ref{Stability of Finite Boards} imply that such a sequence of earthquakes exist).
-We will find that the exact sequence of local earthquakes does not matter, and that if we keep hitting the board with local earthquakes anywhere they are allowed to happen, the board will always stabilize to the same final configuration.\\
+Now that we have shown that all finite boards will eventually reach a stable state we would like to know if the order in which these earthquakes occur matters?
+More specifically, we consider some finite starting board and some sequence of local earthquakes that stabilizes the board (note that our results from Section~\ref{Stability of Finite Boards} imply that such a sequence of earthquakes exist).
+We will find that the number of earthquakes that targets a particular square in order to stabilize the board is invariant of the order in which the earthquakes occur.\\
 
@@
 \begin{definition}
-We define a valid earthquake to be a local earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.
-We say that a sequence of local earthquakes $LE_1, LE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
+We say that an earthquake $E$ is valid if the square it targets has at least $4$ chips on it. 
+We say that a sequence of local earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
+\begin{equation*}
+E^{m}[b, \{s_k\}](s_{m+1}  \geq 4
+\end{equation*}
+for all $1\geq m+1 \leq n$.
 \end{definition}
 
-\begin{theorem}
-Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ local earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
-\end{theorem}
+For ease of notation we also introduce the notion of an indicator function
+\begin{definition}
+We define the indicator function $[[\cdot]]$ as 
+\begin{equation*}
+[[A]] = \begin{cases} 1 \text{ if A is true} \\ 0 \text{ if A is false} \end{cases}
+\end{equation*}
+\end{definition}
 
+Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
+\begin{lemma} 
+\label{earthquakeredefine}
+Consider a board $b$ if $E$ is a valid earthquake, then 
+\begin{equation*}
+E[b, s](x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
+\end{equation*}
+\end{lemma}
 \begin{proof}
+This follows directly from the definition of an earthquake and a valid earthquake. 
+\end{proof}
+
+Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
+\begin{lemma}
+\label{swappinglemma}
+Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting squares $s_1$ and $s_2$ respectively.
+If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and
+\begin{equation*}
+E_2[E_1[b, s_1] s_2] = E_1[LE_2[b, s_2], s_1] 
+\end{equation*}
+\end{lemma}
+\begin{proof}
+We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.
+
+We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[B, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
+Therefore we must have $b(s_1)<8$ and $s_2=s_1$.
+But then $E_1[b, s_1](s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes.
+This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.
+
+Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $(i,j) \in \mathbb{Z} \times \mathbb{Z}$.
+Applying Lemma ~\ref{earthquakeredefine} we have
+\begin{align*}
+E_2[E_1[b, s_1], s_2](x) = E_1[b, s_1])(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
+= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
+= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
+= E_2[B, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
+= E_1[E_2 [b, s_2], s_1](x) 
+\end{align*}
+The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
+\end{proof}
+
+\begin{lemma}
+Let $E_1,E_2,E_3,\dots, E_n$ be a sequence of earthquakes, where $E_k$ hits the square $s_k$.
+Consider  the sequence given by moving some earthquake $E_j$ to position $m<j$ in the sequence, such that $E_j$ is a valid earthquake for all positions between $m$ and $j$, inclusive.
+This sequence is valid and produces the same final board configuration.
+\end{lemma}
+\label{shiftlemma}
+\begin{proof}
+It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes.
+In fact we can also ignore the first $m-1$ terms, because their order is not changed so the board state will be identical after $m-1$ terms for both sequences.
+Let us call this board state $b_{m-1}$
+
+Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, LE_j$ applied to $b_{m-1}$.
+
+We can show this by inducting on the length of the sequence $l=j-m+1$.
+The base case of $l=2$ is shown in lemma ~\ref{swappinglemma}.
+Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$.
+So our sequence is
+\begin{equation*}
+E_m, E_{m+1}, \ldots, E_{j-1}, E_j
+\end{equation*}
+where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
+\begin{align*}
+E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
+E_{j}, E_{m+1}, \ldots,  E_{j-1} 
+\end{align*}
+are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma ~\ref{swappinglemma} we know that the two sequences 
+\begin{align*}
+E_{m}, E_{j} \\
+E_{j}, E_{m}
+\end{align*}
+are both valid and produce the same final board configuration.
+If we apply the same reduction as above  we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board.
+\end{proof}
+
 The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
@@
 
-First, we need a lemma in order to do the rearrangements.
+\begin{theorem}
+Let $b$ be a board that becomes a stable board $b$ after $N$ local earthquakes.
+Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $b$.
+Furthermore the number of times that a square $x$ is targeted is invariant of the order in which the earthquakes occur.
+\end{theorem}
+\begin{proof}
+Given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
+We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes in order to stabilize.
+Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquakes that target  square $i$.
+We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a square that starts with $A_i$ chips on it.
+Thus using Lemma ~\ref{shiftlemma}, we can move these earthquakes to the front of the sequence without changing the final positioning of the board.
+We do this for all squares $i$.
 
-\begin{lemma} Let $LE_1$ and $LE_2$ be local earthquakes.
-If $LE_1, LE_2$ is a valid sequence and $LE_2$ is a valid earthquake then when before $LE_1$ then $LE_2, LE_1$ is also a valid sequence of earthquakes and $LE_2( LE_1 ( B ) ) = LE_1( LE_2 ( B ) )$
-\end{lemma}
-\begin{proof}
-Suppose $LE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $LE_2$ affects the square $(x_2, y_2)$ and its neighbors.
-Suppose  $LE_2, LE_1$ was an invalid sequence, that is $LE_2(B)(x_1, y_1)<4$, but since $LE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$.
-Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$.
-But then $LE_1(B)(x_1, y_1)=3$ which means $LE_1, LE_2$ would not be a valid sequence of earthquakes.
-Thus $LE_2, LE_1$ is a valid sequence.
+The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the squares.
 
-Now consider $LE_2( LE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the local earthquake we see that 
-\begin{align*}
-LE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
-LE_2 ( LE_1 (B)) (u,v) = LE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
-= B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]]  \\
-= B(u,v) + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
-= LE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
-= LE_1 (LE_2 (B))(u,v) 
-\end{align*}
-The critical step in this proof is recognizing that because $LE_1$ and $LE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
-\end{proof}
+Now, find the end of the sequence of moved earthquakes and consider the board at this current point in time, noting that we have new values for all the squares.
+We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes.
+We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
 
-\begin{lemma}
-Let $LE_1,LE_2,LE_3,\dots, LE_k$ be a sequence of earthquakes, where $LE_i$ hits the square $(x_i, y_i)$.
-Consider  the sequence given by moving some earthquake $LE_j$ to position $m<j$ in the sequence, such that $LE_j$ is a valid earthquake for all positions between $m$ and $j$.
-This sequence is valid and produces the same final board configuration.
-\end{lemma}
-\begin{proof}
-It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes.
-In fact we can also ignore the first $m-1$ terms, because their order is not changed they will all be valid and the board state will be identical after $m-1$ terms for both sequences.
-Let us call this board state $B_{m-1}$
-
-Therefore we must show that the sequence $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ applied to $B_{m-1}$.
-
-We can show this by inducting on the length of the sequence $l=j-m+1$.
-The base case of $l=2$ is shown in the above lemma.
-Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$.
-So our sequence is
-\begin{equation*}
-LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j
-\end{equation*}
-where $p+1=j-m+1$. Then if we set $B_{m} = LE_{m} (B_{m-1})$ by applying our inductive step we have that the two sequences 
-\begin{align*}
-LE_{m+1}, \ldots, LE_{j-1}, LE_{j} \\ 
-LE_{j}, LE_{m+1}, \ldots,  LE_{j-1} 
-\end{align*}
-are both valid and produce the same final board configuration. Now if we consider the board state $B_{m-1}$ by our above Lemma we know that the two sequences 
-\begin{align*}
-LE_{m}, LE_{j} \\
-LE_{j}, LE_{m}
-\end{align*}
-are both valid and produce the same final board configuration.
-If we apply the same reduction as above  we have  that $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$  and $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ both applied to $B_{m-1}$ are valid and produce the same board.
-\end{proof}
-
-Given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
-We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes or else it will not eventually stabilize.
-Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquake events that correspond to square $i$.
-All these earthquakes can be moved to the front of the sequence.
-We do this for all squares $i$.
-By the lemma, the final positioning of the board does not change.
-
-The order of the earthquakes that have been moved to the front of the sequence can be in any order.
-This holds true because all these earthquakes could have happened with the initial values of the squares (maybe need to explain this better?
-but i think it's clear...).
-
-Now, find the point in the sequence that begins right after this point.
-Consider the board at this current point in time, and note that we have new values for all the squares.
-Then repeat this argument, moving a new set of earthquakes to the front.
-We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
+Thus we have transformed our starting sequence into
 
@@
 \emph{written by Perry Kleinhenz, edited by Erik Waingarten and Fermi Ma}
+
 Now that we have shown that the order in which earthquakes occur does not matter we might wonder if the behavior of the system changes when all possible earthquakes occur at the \emph{same} time.
@@
 \begin{definition} The active sites $A_b$ of a board $b$ are all those squares which have at least 4 chips on them or written symbolically
-\begin{definition}
+\begin{equation*}
 A_b=\{ (i,j); b(i,j) \geq 4 \}
@@
 We now show that a global earthquake is equivalent to applying standard earthquakes to each of the active sites on a board in any order.
-\begin{lemma} Suppose $b$ is a finite board. If we let arrange the terms of $A_b$ into a sequence $\{a_k}$ with $N$ elements then 
+\begin{lemma} Suppose $b$ is a finite board. If we let arrange the terms of $A_b$ into a sequence $\{a_k\}$ with $N$ elements then 
 \begin{equation*}
@@
 \end{equation*}
-\end{llemma}
+\end{lemma}
 \begin{proof} 
@@
 
-\begin{myth} Suppose $b$ is a finite board, then there exists $N$ such that for all $m>0$ 
+\begin{theorem} Suppose $b$ is a finite board, then there exists $N$ such that for all $m>0$ 
 \begin{equation*}
@@
 \end{equation*}
-\end{myth}
+\end{theorem}
 \begin{proof}

### 2014-10-10 draft0.tex
More changes to Order of distribution section

@@
 We also consider extensions of this problem where the earthquake hits all squares on the board at once.
-We will refer to earthquakes that hit only one square as $\emph{local earthquakes}$ and earthquakes that hit the entire board as $\emph{global earthquakes}$.
+We will refer to earthquakes that hit only one square as $\emph{earthquakes}$ and earthquakes that hit the entire board as $\emph{global earthquakes}$.
 
 In Section~\ref{Definitions and Notation}, we formalize the terminology and operations we use to analyze the problem.
-In particular, we formalize the effects of local earthquakes, as well as what it means for a board of chips to be finite or infinite.
+In particular, we formalize the effects of earthquakes, as well as what it means for a board of chips to be finite or infinite.
 In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to $\mathbb{R}$. 
-We then use the fact that local earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
+We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
 In Section~\ref{Acyclicity of Infinite Boards}, we modify the spread function so that it converges to a real number even for infinite boards, with bounded chip stack size. 
 From this, we conclude that such boards never cycle back to previous configurations. 
-In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a local earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
+In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
 
@@
 In this section we show that any finite board will reach a stable state after finitely many earthquakes.
-We do so by showing that if a local earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
+We do so by showing that if a earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
 We will then give a bound for the maximum number of configurations that earthquakes can transform a given board into. 
@@
 We can in fact show that infinite boards cannot cycle under earthquakes.
-To show this we will mirror the previous section by defining a potential function on boards and then showing that the functions value only increases after a local earthquake has occurred.
+To show this we will mirror the previous section by defining a potential function on boards and then showing that the functions value only increases after an earthquake has occurred.
 
@@
 \frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}}  = \frac{1}{4(d)\cdot 4^{d-1}} = \frac{4}{4(d)\cdot 4^d}
-\end{equation*}which is larger in magnitude than $\frac{4}{4(d+1)\cdot 4^d}$. Therefore the value of $\Phi$ strictly increases after any local earthquake.
-It follows that infinite boards hit with local earthquakes can never cycle back to a previous arrangement.
+\end{equation*}which is larger in magnitude than $\frac{4}{4(d+1)\cdot 4^d}$. Therefore the value of $\Phi$ strictly increases after any earthquake.
+It follows that infinite boards hit with earthquakes can never cycle back to a previous arrangement.
 \end{proof}
@@
 
-Now that we have shown that all finite boards will eventually reach a stable state we would like to know if the order in which these earthquakes occur matters?
-More specifically, we consider some finite starting board and some sequence of local earthquakes that stabilizes the board (note that our results from Section~\ref{Stability of Finite Boards} imply that such a sequence of earthquakes exist).
-We will find that the number of earthquakes that targets a particular square in order to stabilize the board is invariant of the order in which the earthquakes occur.\\
+Now that we have shown that all finite boards will eventually reach a stable state we would like to know if the order in which these earthquakes occur matters.
+More specifically, if we consider some finite starting board and two sequences of earthquakes that stabilizes the board, we would like to know if the resultant board is the same for both sequences.
+
+In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular square in order to stabilize the board is invariant of the order in which the earthquakes occur.
+We do so by first describing conditions under which the position of an Earthquake in a sequence can change without effecting the final state.
+We then define a process that rearranges a sequence of earthquakes without changing the final board state and produces the same sequence for any board.\\
 
@@
 We say that an earthquake $E$ is valid if the square it targets has at least $4$ chips on it. 
-We say that a sequence of local earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
+We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
 \begin{equation*}
@@
 
-For ease of notation we also introduce the notion of an indicator function
+For ease of notation we also introduce an indicator function
 \begin{definition}
@@
 
+We can use this lemma along with a simple induction to show that under appropriate conditions we can move an earthquake earlier in the sequence without affecting the final board state.
 \begin{lemma}
@@
 
+Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order.
 The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
@@
 \begin{theorem}
-Let $b$ be a board that becomes a stable board $b$ after $N$ local earthquakes.
+Let $b$ be a board that becomes a stable board $b$ after $N$ earthquakes.
 Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $b$.
@@
 \begin{proof}
-Given the original configuration of the board, let $A_i$ be the number of chips on square $i$.
-We know that square $i$ will be hit by at least $\lfloor \frac{A_i}{4} \rfloor$ earthquakes in order to stabilize.
+Given the original configuration of the board, let $A_i$ be the number of chips on square $i$. We know that square $i$ will be hit by at least 
+\begin{equation*}
+\lfloor \frac{A_i}{4} \rfloor
+\end{equation*}
+earthquakes in order to stabilize.
 Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquakes that target  square $i$.
@@
 
-Now, find the end of the sequence of moved earthquakes and consider the board at this current point in time, noting that we have new values for all the squares.
+Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the squares.
 We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes.
@@
 
-Thus we have transformed our starting sequence into
-
+Thus we have transformed our starting sequence into canonical sequence that is completely determined by the state of the initial board.
+Therefore we can do this for any other sequence that stabilizes the board and get the same sequence.
+This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them).
 \end{proof}

### 2014-10-10 draft0.tex
modified up to acyclicity of infinite boards

@@
 \begin{quote}
-Each node of an infinite chess board contains some nonnegative number of poker chips.
-An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions.
-Does this process reach a stable state (with fewer than four chips on every square)?
-Is it independent of the order of redistribution?
-What can be said about the question when the underlying graph is different?
+Each node of an infinite grid graph contains some nonnegative number of chips.
+An earthquake hits one node that has at least four chips and redistributes them so that one chip moves to each of its four neighbors.
 \end{quote}
 }
-
-We also consider extensions of this problem where the earthquake hits all squares on the board at once.
-We will refer to earthquakes that hit only one square as $\emph{earthquakes}$ and earthquakes that hit the entire board as $\emph{global earthquakes}$.
 
@@
 We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
-In Section~\ref{Acyclicity of Infinite Boards}, we modify the spread function so that it converges to a real number even for infinite boards, with bounded chip stack size. 
+Then we show that this means that no boards can return to their same state.
 From this, we conclude that such boards never cycle back to previous configurations. 
@@
 We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
-In Section~\ref{Board Variants: n-Trees}, we generalize away from the chessboard set up, and consider what happens when a similar process happens on tree graphs of infinite size. 
+In Section~\ref{Board Variants: n-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
 Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits.
@@
 
-First, we formalize the board of poker chips. We consider the integer lattice $\mathbb{Z} \times \mathbb{Z}$ where each coordinate refers to an individual square.
+First, we formalize the board of poker chips.
+We consider the integer lattice $\mathbb{Z} \times \mathbb{Z}$ as the set of nodes, indexed by the corresponding coordinate.
+We made the two nodes have an edge between them if they are a distance of 1 apart.
 
 \begin{definition} A board is a function $b: \mathbb{Z} \times \mathbb{Z} \to \mathbb{Z}_{\geq 0}$.
-The input to this function correspond to the coordinates of a square and the output is the number of chips on that square. Let $\mathcal{B}$ denote the set of all boards.
+The input to this function correspond to the nodes and the output is the number of chips on that node. Let $\mathcal{B}$ denote the set of all boards.
 \end{definition}
 
-In order to keep our notation compact we define a function that takes in the coordinate of a square and returns the coordinates of its neighbors in a set. 
+In order to keep our notation compact we let $n(i,j)$ be the set of neighbors of node $(i,j)$.
 \begin{definition}
@@
 
-Recall from the introduction that an earthquake (abbreviated as $E$) is an operation that hits one square with more than $4$ chips and redistributes 1 chip to each neighbor.
+Recall from the introduction that an earthquake (abbreviated as $E$) is an operation that hits one node with more than $4$ chips and redistributes 1 chip to each neighbor.
 We define this operation in terms of $b$.
@@
 \begin{definition}
-An earthquake takes as input a board and the coordinates of a grid square and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
-If $b(s) \geq 4$, then
+An earthquake takes as input a board and a node and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$. If $b(s) \geq 4$, then
 \begin{align*}
@@
 \begin{definition}
-We say that a board is stable if all squares contain values strictly less than 4. 
+We say that a board is stable if all nodes contain values strictly less than 4. 
 \end{definition}
-This is equivalent to saying an earthquake has no effect on the board. We would also to have a formal definition for a finite board.
+This is equivalent to saying all earthquakes have no effect on the board. We would also to have a formal definition for a finite board.
 
@@
 \begin{equation}
-C= \sum_{(i,j) \in \mathbb{Z} \times \mathbb{Z}} b(i,j), 
+C= \sum_{s \in \mathbb{Z} \times \mathbb{Z}} b(s), 
 \end{equation}
@@
 \label{finitestability}
-If $f \in B_F$, there exists $N> 0$ such that after $N$ earthquakes the resultant board will be stable, or for all $m > 0$, $E^{N+m}[b] = E^N[b]$.
+If $b \in \mathcal{B}_F$, there exists $N> 0$ such that after $N$ earthquakes the resultant board will be stable, or for all $m > 0$ and all $s \in \mathbb{Z} \times \mathbb{Z}$, $E^{N+m}[b,s] = E^N[b,s]$.
 \end{theorem}
@@
 \begin{lemma}
-Suppose $b \in B_F$. If $E[b, x] \neq b$, then $E^n[b, \{x_n\}] \neq b$ for all $n > 0$ and all sequences $\{x_n\}$.
+Suppose $b \in \mathcal{B}_F$. If $E[b, s] \neq b$, then $E^n[b, \{s_n\}] \neq b$ for all $n > 0$ and all sequences $\{s_n\}$.
 \end{lemma}
@@
 \begin{proof}
-For ease of notation we use $s=(s_1, s_2)$. We define our spread function as $S: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
+For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as $S: \mathcal{B}_F \rightarrow \mathbb{Z}_{\geq 0}$ 
 \begin{equation*}
-\Psi(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
+S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
 \end{equation*}
@@
 \begin{equation*}
-E[f, s] \neq f \Rightarrow \Psi(E[f, s]) > S(f)
+E[b, s] \neq b \Rightarrow S(E[b, s]) > S(b)
 \end{equation*}
-That is if an earthquake changes the board, $\Psi$ increases.
+That is if an earthquake changes the board, $S$ increases.
 This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.
 
-We evaluate $\Psi(E[b, s]) - \Psi(b)$, recalling that $E[b, s](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $x$.
+We evaluate $S(E[b, s]) - S(b)$, recalling that $E[b, s](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
 \begin{align*}
-&\Psi(E[b, s]) - \Psi(b) =\sum_{i,j} E[b, s](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
-&= E[b, s](s_1,s_2)(|s_1|+|s_2|)^2 + \sum_{(i,j) \in n(s)} E[b, s] (i,j) (|i|+|j|)^2 \\
-&- b(s_1, s_2)(|s_1|+|s_2|)^2 + \qquad \;  \; \sum_{(i,j) \in n(s)} b(i,j)(|i|+|j|)^2 \\
-&= -4(|s_1| + |s_2|)^2 +\qquad \qquad \quad \; \; \sum_{(i,j) \in n(s)} (|i| + |j|)^2,
+S(E[b, s]) - S(b) &=\sum_{i,j} E[b, s](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
+&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
 \end{align*}
-For most two of the neighbors of $(s_1, s_2))$ 
+For most two of the neighbors of $(s_1, s_2)$ 
 \begin{equation*}
@@
 \label{finiteextension}
-Suppose $\sum_{i,j} b(i,j) = C$, and $f$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
+Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
 Then   $E^n[b, \{s_k\}]$ is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$for all $n$ and for all sequences $\{s_k\}$ .
@@
 \begin{proof}
-%This proof needs to get fixed
-The most compact stable state of a board with $S$ chips is to have $3$ chips on each square adjacent to each other.
-This happens in a square $\sqrt{S/3} \times \sqrt{S/3}$.
-Similarly, a connected component of chips can only extend by a certain amount, where there are no two 0s adjacent to each other.
-This means that the most spread board which started with a connected component of $S$ chips can be $2\sqrt{S} \times 2\sqrt{S}$.
+Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E[b, s](i,j) \geq 1$.
+This is clear since any earthquake will move either bring things into the column $i$ or if it moves some chips out, then there are two chips that remain in that column.
 
-If we do this for each initial connected component and take the total area of the board, we get the desired bound.
+The same is true for each row. This means that if a chip ever goes to a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
 \end{proof}
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_f$, so $\sum_{i,j} b(i,j) = C$.
+(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_F$, so $\sum_{i,j} b(i,j) = C$.
 Because $b$ is a finite board we know the locations with a nonzero number of chips are bounded.
 Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then by Lemma~\ref{finiteextension}, we have that $E^n[f, \{s_kj \}]$ is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ s_k \}$ of length $n$.
+Then by Lemma~\ref{finiteextension}, we have that $E^n[b, \{s_k \}]$ is nonzero in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ s_k \}$ of length $n$.
 If we set 
@@
 then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,.
-We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{s_k\}$, which is arbitrary after the first $N$ terms, such that $E^N[ f, \{s_k\}] = LE^{N+m}[f, \{s_k\}]$.
+We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{s_k\}$, which is arbitrary after the first $N$ terms, such that $E^N[ b, \{s_k\}] = E^{N+m}[b, \{s_k\}]$.
 \end{proof}
 
-\section{Acyclicity of Infinite Boards}
+\subsection{Acyclicity of Infinite Boards}
 \label{Acyclicity of Infinite Boards}
+\emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\
 
-\emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\
-%Change this to reflect the proof given in the talk
 We can in fact show that infinite boards cannot cycle under earthquakes.
-To show this we will mirror the previous section by defining a potential function on boards and then showing that the functions value only increases after an earthquake has occurred.
 
-We make the quick definition of the set of infinite boards
-\begin{definition}
-We say that a board is infinite if the total number of chips on it is infinite. That is a board $f$ is infinite if the sum 
-\begin{equation*}
- \sum_{(i,j) \in \mathbb{Z} \times \mathbb{Z}} b(x,y)
-\end{equation*}
-does not converge.
-We say that an infinite board $b$ is bounded if there exists $M \in \mathbb{Z}$ such that $b(i,j) \leq M$ for all $(i,j) \in \mathbb{Z} \times \mathbb{Z}$.
-We denote the set of bounded infinite boards as $\mathcal{B}_B$
-\end{definition}
+\begin{corollary}
+All boards are can never return to their same state.
+\end{corollary}
 
-We once again define a potential function, although this time it is from bounded infinite boards to the positive reals. \begin{definition} Let $\Phi: \mathcal{B}_B \rightarrow \mathbb{R}_{\geq 0}$
-\[ \Phi(f) = \sum_{(x,y)}\frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \]
-\end{definition}
-%include the +1 in the denominator so there is not a singularity at (0,0)
-Before it was obvious that our potential function converged and was well defined. However we are now considering boards for which the sum of all chips does not converge. Therefore we show 
-\begin{lemma} $\Phi(f)$ converges for all $f \in \mathcal{B}_B$.
-\end{lemma}
 \begin{proof}
-We note that for $n > 0$, there are exactly $4n$ squares $(x,y)$ such that $|x| + |y| = n$. If we assume that the maximum chip stack is $M$, then we have the bound:
-\begin{align*}
-\Phi(f) = \sum_{(x,y)}\frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \leq \sum_{n=1}^\infty \frac{M}{4^n} = \frac{M}{3}
-\end{align*}
-Thus the sum in the potential function converges and the function is well defined. 
-\end{proof}
-
-\begin{theorem} 
-If $f \in \mathcal{B}_B$ and $LE(f, (x_0,y_0)) \neq f$, then $LE^{n}(f) \neq f$ for all $n>0$.
-\end{theorem}
-\begin{proof}
-We once again compute 
-\begin{align*}
-\Phi(LE(f))-\Phi(f) = \sum_{(x,y)} \frac{LE(f, (x_0,y_0))(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} - \frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \\
-=\frac{LE(f, (x_0,y_0))(x_0,y_0)-f(x_0,y_0)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} + \sum_{(x,y) \in n(x_0,y_0)} \frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} 
-\end{align*}
-Suppose $|x|+|y|=d$, then 
-\begin{equation*}
-\frac{LE(f,(x,y))(x,y)}{4(d+1)*4^d} -\frac{f(x,y)}{4(d+1)*4^d} =-\frac{4}{4(d+1)\cdot 4^d}
-\end{equation*}
-and since $LE(f, (x,y))(x_0,y_0)>f(x,y)$ for $(x,y) \in n(x_0,y_0)$ we have
-\begin{equation*}
-\frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} >0
-\end{equation*}
-Furthermore we know that one of the adjacent squares is at distance $d-1$, and that 
-\begin{equation*}
-\frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}}  = \frac{1}{4(d)\cdot 4^{d-1}} = \frac{4}{4(d)\cdot 4^d}
-\end{equation*}which is larger in magnitude than $\frac{4}{4(d+1)\cdot 4^d}$. Therefore the value of $\Phi$ strictly increases after any earthquake.
-It follows that infinite boards hit with earthquakes can never cycle back to a previous arrangement.
+We have shown that finite boards never return to the same state.
+Suppose that an infinite board $b$ reurns to the same state, then it does so after $N$ earthquakes at locations $\{ s_k \}$.
+However, we can bound the size of the locations in $\{ s_k \}$ and so this infinite board acts in the same way that a finite board on this bound acts.
+Therefore, if the infinite board cycles, then the finite board cycles as well.
 \end{proof}

### 2014-10-10 simulate.py

@@ -133,6 +133,8 @@ def drawBoard(board, xrange, yrange, maxValue, recomputeMax=False, showActive=Fa
             r, g, b = point*dr, 0, 0
             if point >= 4 and showActive:
                 r, g, b = 0, 255, 0
+            if point > 1:
+                g = 100
             if r > 255:
                 r = 255
                 g = point*dg
@@ -154,10 +156,11 @@ def showImage(board, xrange, yrange, wait=False, showActive=False):
         changed = act(board)
         img = drawBoard(board, xrange, yrange, maxValue, True, showActive=showActive)
         pl.imshow(img)
-        pl.pause(0.02)
+        if not wait:
+            pl.pause(0.02)
         pl.draw()
-        if wait:
-            raw_input()
+    #    if wait:
+    #        raw_input()
     print "Done"
 
 def wire(x,y):
@@ -230,3 +233,22 @@ def showANDGate(bit1, bit2):
     if bit2 == True:
         gate[(2, -1)] = 4
     showImage(gate, 14, 12, True, True)
+
+def present():
+    showImage({(0,0):30}, 10, 10)
+    print "Press to go on"
+    raw_input()
+    showImage({(0,0):100}, 15, 15)
+    print "Press to go on"
+    raw_input()
+    showWireActing()
+    print "Press to go on"
+    raw_input()
+    showSynchronizer()
+    raw_input()
+    showORGate(True, False)
+    showORGate(False, False)
+    print "Press to go on"
+    raw_input()
+    showANDGate(True, False)
+    showANDGate(True, True)

### 2014-10-11 draft0.tex
Working on d-Trees section

@@
 We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
-In Section~\ref{Board Variants: n-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
+In Section~\ref{Board Variants: d-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
 Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits.
@@
 
-\section{Board Variants: $n$-Trees}
-\label{Board Variants: n-Trees}
+\section{Board Variants: $d$-Trees}
+\label{Board Variants: d-Trees}
 \emph{written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten.}\\
 
-Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.
+\begin{definition}
+A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each have $d-1$ unique children and so on.  
+\end{definition}
+We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
+\begin{definition}
+A joined $d$-tree is two $d$-trees connected at their parent nodes by an edge.
+\end{definition}
 
-We can join two semi infinite $n$-trees by an edge connecting their parent nodes.
-This forms a graph where every vertex has $n+1$ neighbors.
-We would like to adapt our description of chips and earthquakes to this graph.
-
-\begin{definition} We define a coordinate system for our graph, with the following rules:
+We now define a coordinate system so that our method of proof can mirror that of the standard board case.
+\begin{definition} We define a coordinate system for our joined $d$-trees, with the following rules:
 \begin{enumerate}
 	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
-	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq n^{|a|}$
-	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq n^{|a-1|}$
+	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq (d-1)^{|a|}$
+	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq (d-1)^{|a-1|}$
 \end{enumerate}
-We define the set of all valid coordinates $G_n$ as 
+We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
 \begin{equation}
-G_n := \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
+G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
 \end{equation}
@@
 
+In order to keep our notation compact we once again define a function that outputs the neighbors of a node
 \begin{definition} 
-We define a tree to be a map from the set of valid coordinates to the nonnegative integers, that is 
-\begin{equation}
-f:  G\rightarrow \mathbb{Z}_{\geq 0}
-\end{equation}
-We write $\mathcal{T}_n$ to refer to the set of all joined $n$-tree's 
+We define the function $n: G_d \rightarrow G_d^d$ as 
+\begin{equation*}
+n(i,j) = \{ (x_1, y_1); (x_1 = i+1 \text{ or } x_1=i-1), \text{ and }(x_1, y_1) \in G_d\}
+\end{equation*}
 \end{definition}
 
+We now formalize the board of poker chips.
+Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate in the integer lattice.
 \begin{definition}
-\label{bigearthquakedeftree}
-We define the global earthquake operator as  as a map from the set of all $n$ trees to the set of all $n$ trees and write $GE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
-It transforms all of the nodes on the tree with the following rule.
+We define a $d$-board to be a map from the set of valid coordinates to the nonnegative integers, that is 
 \begin{equation}
-GE_T(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
+b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
 \end{equation}
-where $k$ is defined as 
-\begin{equation}
-k = \begin{cases}
-1 \text{ if } x\leq 0 \\
--1 \text{ if } x \geq 1
-\end{cases}
-\end{equation}
-and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$ and $[[A]]$ is defined as 
-\begin{equation}
-[[A]] = 
-\begin{cases} 
-1 \text{ if } A \text{ is true} \\ 
-0 \text{ if } A  \text{ is false}
- \end{cases}.
-\end{equation}
+We write $\mathcal{B}(d)$ to refer to the set of all $d$-boards.  
 \end{definition}
 
+Recalling that $n(s)$ refers to the neighbors of $s$ we define our earthquake operation in terms of $d$-boards.
 \begin{definition}
- We define the local earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $LE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
-It selects one valid node $(x,y)$ with $f(x,y) \geq n+1$ and transforms it and all of its neighbors with the following rule: 
+An earthquake takes as an input an $d$-board and a node and outputs an $d$-board so we write $E: \mathcal{B}(d) \times G_d \rightarrow \mathcal{B}(d)$. If $s$ is the node input and $b(s)>d$ then 
 \begin{align*}
-LE_T( f( x, y ) ) = f( x , y )-n-1 \\
-LE_T( f( x + k, y)) = f( x + k, y )+1 \\
-LE_T( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
+E[b, s](s) = b(s)- d \\
+E[b, s](t) = b(t) +1 
 \end{align*}
-where $k$ is as defined above and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$
-and leaves all other squares unchanged.
+for $t \in n(s)$.
+Otherwise $b(s)<d$ and $E[b,s]=b$ as the earthquake does not move any chips.
+We use the notation $E^r[b, \{s_k\}]$ to denote the application of $E$ to the $b$ $r$ times, to the sites $s_1, \ldots s_r$ respectively.
 \end{definition}
 
+Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a board. 
 \begin{definition}
-We say that a $n$-Tree is stable if its values are strictly less than $n+1$. This is equivalent to both of the earthquake maps having no effect on the board. 
+We say that a $d$-board is stable if its values are strictly less than $n$.
+\end{definition}
+This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite board.
+\begin{definition} 
+We say that a $d$-board is finite if the total number of chips on it is finite. That is a $d$board $b$ is finite if the sum
+\begin{equation}
+C= \sum_{(i,j) \in G_d} b(i,j) 
+\end{equation}
+is finite. 
+We denote the set of finite $d$-boards as $B(d)_F$.
 \end{definition}
 
+We will now show that any finite $d$-board will stabilize under earthquakes.
+Our method of proof will mirror the method used for the square lattice, without only minor changes.
+Therefore we will proceed by shooing that if an earthquake changes the state of a $d$-board, then no sequence of earthquakes can transform the output $d$-board back into the original $d$-board.
+We will then give a bound on the maximum number of configurations a given $d$-board can be transformed into by earthquakes.
+Once again these together will show that all finite $d$-boards will stabilize.
+
+We once again define a spread function 
 \begin{definition}
-We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
+ Let $\Psi: \mathcal{B}(d)_F \rightarrow \mathbb{Z}_{\geq 0}$ be the function
 \begin{equation}
-S= \sum_{(x,y) \in G_n} f(x,y) 
-\end{equation}
-is finite. 
-We denote the set of finite boards as $\mathcal{T}_n^f$.
-\end{definition}
-
-As it turns out we can use a  method of proof nearly identical to the standard board case to show that the state of a tree under this process can never cycle.
-
-We once again define a potential function 
-\begin{definition} Let $\Phi: \mathcal{T}_n^f \rightarrow \mathbb{Z}_{\geq 0}$ be the  function
-\begin{equation}
-\Phi(f) = \sum_{(x,y) \in G_n} f(x,y)(|x|)^2
+\Psi(f) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
 \end{equation}
@@
 \end{definition} 
+We can once again think of this spread function as the sum of the number of chips on every node weighted by the taxi-cab distance from the node to the origin.
 
 \begin{lemma}
-If $LE_T(f) \neq f$, then $LE_T^n(f) \neq f$ for all $n > 0$.
+If $E[b, s_1] \neq b$, then $E^r[b, \{s_k\}] \neq b$ for all $r>0$ and all sequences $\{s_k\}$ of length $r$ that begin with $s_1$. 
 \end{lemma}
@@
 \begin{proof}
-We will show that  if $LE$ acts on a board nontrivially then $\Phi(LE(f))$ will be strictly larger. That is  
+For ease of notation let $s=(s_1, s_2)$. We will show that if $E$ acts on a board nontrivially then $\Psi(E[b, s])$ will be strictly larger. That is  
 \begin{equation}
-LE(f) \neq f \Rightarrow \Phi(LE(f)) > \Phi(f)
+E[b, s] \neq f \Rightarrow \Psi(E[b, s]) > \Psi(b)
 \end{equation}
 
-We can evaluate $\Phi(LE(f)) - \Phi(f)$.
+We can evaluate $\Psi(E[b, s]) - \Psi(b)$.
 \begin{align}
-\Phi(LE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} LE_T(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
+\Psi(E[b, s])-\Psi(b) = \sum_{(i,j) \in G_d} E[b, s](i,j)|i|^2 - \sum_{(i,j) \in G_d} b(i,j)|i|^2 
 \end{align}
-Only one node call it $(x_0,y_0)$ and its neighbors are affected. Therefore the above expression simplifies to
+We note that only $s$ and its neighbors are affected so our expression simplifies to
 \begin{align*}
-\Phi(LE_T(f)-\Phi(f) = LE_T( f(x_0,y_0) )* (x_0)^2 + LE_T( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
-+ \sum_{i} LE_T( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
-= -(n+1)*(|x_0|)^2 + (|x_0+k)^2 + n (|x_0-k|)^2 \\
-\geq-(n+1)*(|x_0|)^2 + (|x_0|-1)^2 + n (|x_0|+1)^2
-= -nx^2-x^2+x^2-2|x|+1+n(x^2+2|x|+1) \\
-= 2x(n-1)+n+1 > 0
+\Psi(E[b, s]) - \Psi(b) &= E[b, s](s) |s_1|^2 - b(s) |s_1|^2 + \sum_{(i,j) \in n(s)} (E[b,s](i,j) |i|^2 - b(i,j)|i|^2) \\
+&= -d |s_1|^2 + \sum_{(i,j) \in n(s)} |i|^2
 \end{align*}
-Where $(x_0-k,y_i)$ refers to the $j$th child of $(x_0,y_0)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
-\end{proof}
-
-\begin{lemma}
-If $GE_T(f) \neq f$, then $GE_T^n(f) \neq f$ for all $n > 0$.
-\end{lemma}
-
-\begin{proof}
-We will show that 
-\begin{equation}
-GE_T(f) \neq f \Rightarrow \Phi(GE_T(f)) > \Phi(f)
-\end{equation}
-This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.
-
-We can evaluate $\Phi(GE_T(f)) - \Phi(f)$.
-\begin{align}
-\Phi(GE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} GE_T(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
-\end{align}
-Let $A$ be the set of active locations, or locations with $f(x,y) \geq (n+1)$, which is non-empty by assumption. 
+We note that exactly $d-1$ of the neighbors of $s$ are a distance $|s_1 +1|$ from the origin and one of the neighbors of $s$ is a distance $(|s_1| - 1)$ from the origin.
+A point to emphasize is that this is the only substantial way in which this proof differs from the similar proof for the square lattice.
+Our expression simplifies to
 \begin{align*}
-\Phi(GE_T(f)) - \Phi(f) =  \sum_{(x,y) \in A}  GE_T(f(x,y))x^2 + GE_T(f(x+k,y))(|x+k|)^2 + \\
-\sum_{i} \left( GE_T(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
-=\sum_{(x,y) \in A} -(n+1)x^2 + (|x+k|)^2 + n(|x-k|)^2 
+&= -d |s_1|^2 + \sum_{(i,j) \in n(s)}-|i|^2 \\
+&= - d|s_1|^2 + (d-1)(|s_1| +1)^2 +(|s_1| -1)^2 \\
+&= d + (d-2)2|s_1| >0
 \end{align*}
-Where $(x-k,y_i)$ refers to the $j$th child of $(x,y)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
-We note that each term in this sum is positive by our proof of the lemma for local earthquakes and so the sum itself must be positive.
-That is
-\begin{equation}
-\Phi(GE_T(f)) - \Phi(f) > 0
-\end{equation}
-
 \end{proof}
@@
 \end{proof}
-
-\section{Computing with Chips}
-\label{Computing with Chips}
-
-\emph{Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz.}
-
-In this section, we show how we can compute different Boolean functions using the poker chip board and the earthquake operator.
-
-A Boolean input is a number in $\{0,1\}$, and a Boolean function is a function that takes $k$ Boolean inputs and outputs a number in $\{0,1\}$.
-A simple Boolean function is the AND function that takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are set to 1.
-Another Boolean function is the OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
-If 0 is associated with ``false" and 1 is associated with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
-More complicated Boolean functions can be made with more than 2 inputs and compositions of these inputs.
-Denoting an AND with $\vee$ an OR with $\wedge$, we can construct composite functions such as $b_1 \wedge (b_2 \vee b_3)$.
-
-Boolean circuits are formed by using wires that transmit 0's and 1's, and feeding them through various \emph{gates}.
-For example, a wire carrying a 0 and a wire carrying a 1 can be fed through an OR gate, which will ``output" a wire carrying a 1.
-In this section, we construct circuits using the poker chips board that use AND an OR gates.
-
-\subsection{Model of Computation}
-
-Our model of computation is as follows: we assume a finite board with finitely many chips.
-The inputs are designated by each two squares, one apart one is the ``clock" square and other the bit square.
-Likewise, the outputs each have a ``clock" square and a bit square.
-The ``clock" squares are set the value 4 and the bit squres are set to 4 if they are 1 and 0 otherwise.
-A computation is done by a series of earthquakes, when an output has the clock square active, then the bit square is read, and that output bit is considered read.
-
-\subsection{Example}
-
-A simple example is given by a board that computes the identity.
-
-\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
-				     4_c & 3 & 3 & 3 & 3 & 3_e \\
-				     0 & 0 & 0 & 0 & 0 & 0 \\
-				     i  & 3 & 3 & 3 & 3 & 3_o \\
-				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
-Where the subscript $c$ indicates the clock, the subscript $e$ indicates the output clock square. $i$ indicates the bit square, and $o$ indicates the output bit.
-
-One can see that $e$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $e$.
-Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active.
-Therefore, we can compute the identity.
-
-This circuit also shows how to make a ``wire" in the circuit. We will represent a wire as two parralel arrows, one containing the clock path and the other carrying the input.
-
-\subsection{Turner}
-
-Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
-this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $e$ is the output clock and $o$ is the output bit.
-
-\[ \begin{array}{cccccc} 0 & 0 & e & 0 & o \\
-				     0 & 3 & 2 & 0 & 3 \\
-			            0 & 2 & 3 & 0 & 3 \\
-				    0 & 3 & 2 & 0 & 3 \\
-				    0 & 2 & 3 & 0 & 3 \\
-				    c & 3 & 3 & 0 & 3 \\
-				    0 & 0 & 0 & 0 & 3 \\
-				    i & 3 & 3 & 3 & 3 \end{array} \]	
-Since the clock snakes along the path while $i$ takes the shorter path. 
-Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
-
-\subsection{Synchronize}
-
-We can synchronize the inputs so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
-\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
-					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
-					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
-					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
-					 0 & 0 & 0 & 0 & o & 0 & e & 3 \end{array} \]
-Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
-$i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.
-
-The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
-
-\subsection{Boolean circuits}
-
-Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
-The following can compute the AND of two inputs.
-
-\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
-					   &    & 3   & 2 &  3    &    &  \\
-					   &    &    & 3 &    & 3 & c_2 \\
-					o & 3 & 3 & 2 & 3 & 2 & 	    \\
-					   &    &    &    &    & 3 & i_2 \end{array} \]
-Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
-So in this case, $o = i_1$ AND $i_2$.
-
-Similarly, we can compute the OR of two circuits with a very simple modification:
-\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
-					   &    & 3   & 2 &  3    &    &  \\
-					   &    &    & 3 &    & 3 & c_2 \\
-					o & 3 & 3 & 3 & 3 & 2 & 	    \\
-					   &    &    &    &    & 3 & i_2 \end{array} \]
-And we can do a similar set up to synchronize the clocks.
-
-So currently, we can compute any boolean circuit which can be drawn in a planar graph with only ANDs and ORs. So the question is whether we can compute the NOT? We argue that this is not possible.
-
-\begin{theorem}
-There does not exists any board configuration which computes the NOT operator in the above model of computation.
-\end{theorem}
-
-\begin{proof}
-Suppose we have a board configuration $B$ with input squares $c$, $i$ and output squares $e, o$ that computes the NOT function.
-Then we can make another board $B'$ that synchronizes the final clock with the input so that we do not need to worry about timing of the clock.
-That is, we can guarantee that if $o$ is active, then $o$ is active at the same time $e$ is active.
-
-Now lets make $B'$ an infinite board by putting 0s on the sides.
-So we have a finite board with finitely many chips.
-Let $S_{c_0i_0}$ be the set of squares which were active (had an earthquake affect them) during the execution of $B$ with the input value $i = i_0$ and $c = c_0$.
-
-The first claim to make is that $S_{c_0i_0} \subset S_{c_1i_1}$ if $c_0 \leq c_1$ and $i_0 \leq i_1$.
-This is clearly true since the order of redistribution does not matter.
-In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with local earthquakes.
-
-However, this arises in a contradiction, since $o \in S_{10}$ but $o \notin S_{11}$. Therefore, computing NOT with a board is not possible.
-\end{proof}
-
-What does this mean?
-Well, in particular, it means that we cannot do much.
-In particular, we can't add, since $x + 1$ will act as the NOT operator on the lowest order bit.
-We can only compute boolean circuits in planar graphs.
-
-\end{document}

### 2014-10-11 draft0.tex
Fixed d-Tree section

@@
 \label{finiteextensiontree}
-Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $GE_T^m(f)$ and $LE_T^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
+Suppose $\sum_{x,y} f(x,y) = C$, and the first coordinate of all nodes which are nonzero are in $[i_{min}, i_{max}]$.
+Then $E^r[b, \{s_k\}]$ only contain chips within the range $[i_{min} - 2C, i_{max} + 2C]$ for all $r>0$ and all sequences $\{s_k\}$ of length $r$.
 \end{lemma}
@@
 \begin{proof}
-It is obvious that a board with $S$ chips on level $l<0$ will have a lower $X_{min}$ than a board with $S$ chips on level 0, and the analogous statement for $X_{max}$ holds as well.
+It is obvious that a board with $C$ chips on nodes with first coordinate $l<0$ will have a lower $i_{min}$ than a board with $S$ chips on level 0, and the analogous statement for $i_{max}$ holds as well.
 
-Now as a very crude lower bound we know that $S$ chips on a single node at level $l<0$ will have an $X_{min}$ no lower than $l-2S$ as there must be at least one chip on every other level, as chips only move one level at a time.
+Now because chips cannot be transferred between two nodes whose first coordinate differs by more than 1, we know that $C$ chips on a single node with first coordinate $l<0$ will have an $i_{min}$ no lower than $l-2C$ as there must be at least one chip on every other level.
 
-If we take any other arrangement of $S$ chips, with $X_{min}=l$ but not all $S$ chips are on a single node.
-We know that $X_{min}$ for this board is bounded below by $l-2S$.
-In order to have an $X_{min}$ below this at some point in time it would be possible to have $S$ chips on the same node on level $l$.
-But this cannot occur because in order to move a chip down a level, one chip must move up a level.
-Therefore if $X_{min}=x_l$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not contain chips below level $x_l -2S$.
+If we take any other arrangement of $C$ chips, with $i_{min}=l$ but not all $C$ chips are on a single node.
+We know that $i_{min}$ for this board is bounded below by $l-2C$.
+In order to have an $i_{min}$ below this at some point in time it would be possible to have $C$ chips on the same node on level $l$.
+But this cannot occur because in order to move a chip to a node with a lower first coordinate, one chip must moved to a node with a higher first coordinate.
+Therefore $E^r[b, \{s_k\}]$ does not have a node with a nonzero number of chips on it with first coordinate less than $i_{min} -2C$.
 
-An analogous argument shows that if $X_{max}=x_h$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not contain chips above level $x_h+2S$.
+An analogous argument shows that if $i_{max}$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not have a node with a nonzero number of chips on it with first coordinate greater than $i_{max}+2C$.
 \end{proof}
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie between levels  $[x_l, x_h]$.
-Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $GE_T^m(f)$ and $LE_T^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$
+(of Theorem~\ref{finitestability}) Suppose $b \in \mathcal{B}(d)_f$ is a finite board, so $\sum_{(i,j) \in G_n} b(i,j) = C$, and suppose all chips lie between levels  $[i_{min}, i_{max}]$.
+Then by Lemma ~\ref{finiteextensiontree}, we have that for all $r$ and sequence $\{s_k\}$ all chips in $E^r[b, \{s_k\}]$ will have first coordinate in $ [i_{min} -2C, i_{max} +2C]$.
 
@@
 Thus the total number of ways that the chips can be arranged on the board is finite. 
-In particular there are fewer than $(x_h - x_l+1 + 4 S)$ levels and there are at most 
+In particular there are fewer than $(i_{max} - i_{min}+1 + 4 C)$  and there are at most 
 \begin{equation}
-n^d,
+d^l,
 \end{equation}
-nodes at each level, where $d= \max(x_h, |x_l|)$. So there are at most 
+nodes at each level, where $l= \max(|i_{min}|, |i_{max}|)$. So there are at most 
 \begin{equation}
-N:=(x_h - x_l+1 + 4 S)n^d
+N:=(i_{max} - i_{min}+1 + 4 C)d^l
 \end{equation}
@@
 \end{equation}
-possible boards $b \in \mathcal{T}_n^f$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \leq N < \binom{N+S-1}{N-1}$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
+possible boards $b \in \mathcal{B}(n)_f$ that satisfy this.
+We know that there are no cycles in the states, so there must be an $0 \leq N < \binom{N+S-1}{N-1}$ and some sequence $\{s_k\}$ such that $E^N[b, \{s_k\}] = E^{N+m}[b, \{s_k\}]$ for all $m \geq 0$.
 \end{proof}
+
+\section{Computing with Chips}
+\label{Computing with Chips}
+
+\emph{Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz.}
+
+In this section, we show how we can compute different Boolean functions using the poker chip board and the earthquake operator.
+
+A Boolean input is a number in $\{0,1\}$, and a Boolean function is a function that takes $k$ Boolean inputs and outputs a number in $\{0,1\}$.
+A simple Boolean function is the AND function that takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are set to 1.
+Another Boolean function is the OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
+If 0 is associated with ``false" and 1 is associated with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
+More complicated Boolean functions can be made with more than 2 inputs and compositions of these inputs.
+Denoting an AND with $\vee$ an OR with $\wedge$, we can construct composite functions such as $b_1 \wedge (b_2 \vee b_3)$.
+
+Boolean circuits are formed by using wires that transmit 0's and 1's, and feeding them through various \emph{gates}.
+For example, a wire carrying a 0 and a wire carrying a 1 can be fed through an OR gate, which will ``output" a wire carrying a 1.
+In this section, we construct circuits using the poker chips board that use AND an OR gates.
+
+\subsection{Model of Computation}
+
+Our model of computation is as follows: we assume a finite board with finitely many chips.
+The inputs are designated by each two squares, one apart one is the ``clock" square and other the bit square.
+Likewise, the outputs each have a ``clock" square and a bit square.
+The ``clock" squares are set the value 4 and the bit squres are set to 4 if they are 1 and 0 otherwise.
+A computation is done by a series of earthquakes, when an output has the clock square active, then the bit square is read, and that output bit is considered read.
+
+\subsection{Example}
+
+A simple example is given by a board that computes the identity.
+
+\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
+				     4_c & 3 & 3 & 3 & 3 & 3_e \\
+				     0 & 0 & 0 & 0 & 0 & 0 \\
+				     i  & 3 & 3 & 3 & 3 & 3_o \\
+				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
+Where the subscript $c$ indicates the clock, the subscript $e$ indicates the output clock square. $i$ indicates the bit square, and $o$ indicates the output bit.
+
+One can see that $e$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $e$.
+Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active.
+Therefore, we can compute the identity.
+
+This circuit also shows how to make a ``wire" in the circuit. We will represent a wire as two parralel arrows, one containing the clock path and the other carrying the input.
+
+\subsection{Turner}
+
+Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
+this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $e$ is the output clock and $o$ is the output bit.
+
+\[ \begin{array}{cccccc} 0 & 0 & e & 0 & o \\
+				     0 & 3 & 2 & 0 & 3 \\
+			            0 & 2 & 3 & 0 & 3 \\
+				    0 & 3 & 2 & 0 & 3 \\
+				    0 & 2 & 3 & 0 & 3 \\
+				    c & 3 & 3 & 0 & 3 \\
+				    0 & 0 & 0 & 0 & 3 \\
+				    i & 3 & 3 & 3 & 3 \end{array} \]	
+Since the clock snakes along the path while $i$ takes the shorter path. 
+Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
+
+\subsection{Synchronize}
+
+We can synchronize the inputs so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
+\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
+					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
+					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
+					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
+					 0 & 0 & 0 & 0 & o & 0 & e & 3 \end{array} \]
+Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
+$i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.
+
+The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
+
+\subsection{Boolean circuits}
+
+Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
+The following can compute the AND of two inputs.
+
+\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
+					   &    & 3   & 2 &  3    &    &  \\
+					   &    &    & 3 &    & 3 & c_2 \\
+					o & 3 & 3 & 2 & 3 & 2 & 	    \\
+					   &    &    &    &    & 3 & i_2 \end{array} \]
+Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
+So in this case, $o = i_1$ AND $i_2$.
+
+Similarly, we can compute the OR of two circuits with a very simple modification:
+\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
+					   &    & 3   & 2 &  3    &    &  \\
+					   &    &    & 3 &    & 3 & c_2 \\
+					o & 3 & 3 & 3 & 3 & 2 & 	    \\
+					   &    &    &    &    & 3 & i_2 \end{array} \]
+And we can do a similar set up to synchronize the clocks.
+
+So currently, we can compute any boolean circuit which can be drawn in a planar graph with only ANDs and ORs. So the question is whether we can compute the NOT? We argue that this is not possible.
+
+\begin{theorem}
+There does not exists any board configuration which computes the NOT operator in the above model of computation.
+\end{theorem}
+
+\begin{proof}
+Suppose we have a board configuration $B$ with input squares $c$, $i$ and output squares $e, o$ that computes the NOT function.
+Then we can make another board $B'$ that synchronizes the final clock with the input so that we do not need to worry about timing of the clock.
+That is, we can guarantee that if $o$ is active, then $o$ is active at the same time $e$ is active.
+
+Now lets make $B'$ an infinite board by putting 0s on the sides.
+So we have a finite board with finitely many chips.
+Let $S_{c_0i_0}$ be the set of squares which were active (had an earthquake affect them) during the execution of $B$ with the input value $i = i_0$ and $c = c_0$.
+
+The first claim to make is that $S_{c_0i_0} \subset S_{c_1i_1}$ if $c_0 \leq c_1$ and $i_0 \leq i_1$.
+This is clearly true since the order of redistribution does not matter.
+In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with local earthquakes.
+
+However, this arises in a contradiction, since $o \in S_{10}$ but $o \notin S_{11}$. Therefore, computing NOT with a board is not possible.
+\end{proof}
+
+What does this mean?
+Well, in particular, it means that we cannot do much.
+In particular, we can't add, since $x + 1$ will act as the NOT operator on the lowest order bit.
+We can only compute boolean circuits in planar graphs.
+
+\end{document}

### 2014-10-11 finaldraft.tex
adding final draft
modified notations section

--- 2014-10-11 draft0.tex
@@
 
-First, we formalize the board of poker chips.
-We consider the integer lattice $\mathbb{Z} \times \mathbb{Z}$ as the set of nodes, indexed by the corresponding coordinate.
-We made the two nodes have an edge between them if they are a distance of 1 apart.
+First, we formalize the definition of a board. The board is an infinite grid graph with a nonnegative number of chips placed at each node
+\begin{center}
+\includegraphics[scale=0.5]{grid1.png}
+\end{center}
 
-\begin{definition} A board is a function $b: \mathbb{Z} \times \mathbb{Z} \to \mathbb{Z}_{\geq 0}$.
-The input to this function correspond to the nodes and the output is the number of chips on that node. Let $\mathcal{B}$ denote the set of all boards.
+We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$.
+We will define the board as a function from the Cartesian grid to the nonnegative integers.
+
+\begin{definition} We write boards as functions $b$ from $\mathbb{Z}^2 \to \mathbb{Z}_{\geq 0}$.
 \end{definition}
 
-In order to keep our notation compact we let $n(i,j)$ be the set of neighbors of node $(i,j)$.
-\begin{definition}
-We define the function $n: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 $ as 
-\begin{equation}
+This function maps node labels given as integer coordinates to chip stack sizes. For example, if a board $b$ has 15 chips at the node with coordinate $(5,4)$, then $b(5,4) = 15$.
+
+\begin{definition} We define $n(i,j)$ to be the set of neighbors of node $(i,j)$
+\begin{equation*}
 n(i,j) = \{ (i+1, j), (i-1, j), (i, j-1), (i, j+1) \}.
-\end{equation}
+\end{equation*}
 \end{definition}

### 2014-10-11 finaldraft.tex
changed some notation stuff

@@
 
-In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular square in order to stabilize the board is invariant of the order in which the earthquakes occur.
+In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular node in order to stabilize the board is invariant of the order in which the earthquakes occur.
 We do so by first describing conditions under which the position of an Earthquake in a sequence can change without effecting the final state.
@@
 \begin{definition}
-We say that an earthquake $E$ is valid if the square it targets has at least $4$ chips on it. 
+We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
 We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
@@
 \label{swappinglemma}
-Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting squares $s_1$ and $s_2$ respectively.
+Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively.
 If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and
@@
 
-We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[B, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
+We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[b, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
 Therefore we must have $b(s_1)<8$ and $s_2=s_1$.
@@
 
-Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $(i,j) \in \mathbb{Z} \times \mathbb{Z}$.
+Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$.
 Applying Lemma ~\ref{earthquakeredefine} we have
@@
 = b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
-= E_2[B, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
+= E_2[b, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
 = E_1[E_2 [b, s_2], s_1](x) 
@@
 \begin{lemma}
-Let $E_1,E_2,E_3,\dots, E_n$ be a sequence of earthquakes, where $E_k$ hits the square $s_k$.
+Let $E_1,E_2,E_3,\dots, E_n$ be a sequence of earthquakes, where $E_k$ hits the node $s_k$.
 Consider  the sequence given by moving some earthquake $E_j$ to position $m<j$ in the sequence, such that $E_j$ is a valid earthquake for all positions between $m$ and $j$, inclusive.

### 2014-10-11 finaldraft.tex
finished section on definitions on notation. still need to add figures and figure references though

--- 2014-10-11 finaldraft.tex
@@
 
+\begin{center}
+\includegraphics[scale=0.8]{cartesiangrid.png}
+\end{center}
 We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$.
@@
 
-This function maps node labels given as integer coordinates to chip stack sizes. For example, if a board $b$ has 15 chips at the node with coordinate $(5,4)$, then $b(5,4) = 15$.
+This function maps node labels given as integer coordinates to chip stack sizes. For example, FIGUREREFERENCE if a board $b$ has 8 chips at the node with coordinate $(3,1)$, then $b(3,1) = 8$.
 
@@
 
-Recall from the introduction that an earthquake (abbreviated as $E$) is an operation that hits one node with more than $4$ chips and redistributes 1 chip to each neighbor.
-We define this operation in terms of $b$.
+We now formally define the earthquake operation. Recall that an earthquake hits a node that has 4 or more chips and moves 1 chip to each of the node's 4 neighbors.
 
 \begin{definition}
-An earthquake takes as input a board and a node and outputs a board, so we write $E: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$. If $b(s) \geq 4$, then
+An earthquake hits a board $b$ at node $s$, and the resulting board is denoted $E_sb$. If $b(s) \geq 4$, then
 \begin{align*}
-E[b, s](s) &= b(s)-4 \\
-E[b, s](t) &= b(t)+1, \\
+E_sb(s) &= b(s)-4,\\
+E_sb(t) &= b(t)+1 \hspace{.5cm} \forall t \in n(s).\\
 \end{align*}
-for $ t \in n(s)$. Otherwise  $b(s) < 4$, and $E[b, s] = b$, as the earthquake does not move any chips.
-We use the notation $E^n[b, \{s_k\}]$ to denote the application of $E$ to the board $b$ $n$ times, to the squares $s_1, \ldots s_{n}$ respectively. 
+if $b(s) < 4$, the earthquake has no effect and $E_sb = b$. 
 \end{definition}
 
-Now that we have formalized the earthquake operator we would like to make a precise definition of stability of a board.
+Often, a number of earthquakes will hit a board $b$.
+We represent the result of multiple earthquakes by a composition of the earthquake functions.
+For example, if an earthquake hits node 1, then node 2, and finally node 3, the resulting board is written as $E_3E_2E_1b$.
+
+Now that we have formalized the earthquake operator, we make precise the definition of stability.
 \begin{definition}
@@
 \end{definition}
-This is equivalent to saying all earthquakes have no effect on the board. We would also to have a formal definition for a finite board.
+This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be affected by earthquake operations.
+
+Finally, we formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.
 
 \begin{definition} 
-We say that a board is finite if the total number of chips on it is finite. That is a board $b$ is finite if the sum
-\begin{equation}
-C= \sum_{s \in \mathbb{Z} \times \mathbb{Z}} b(s), 
-\end{equation}
-is finite. Otherwise we say a board is infinite. 
-We denote the set of finite boards as $\mathcal{B}_F$.
+We say that a board is finite if the total number of chips on it is finite. Otherwise we say a board is infinite. 
 \end{definition}

### 2014-10-11 finaldraft.tex
Unbotched It
Edited First 3 sections

@@
-\documentclass[runningheads,a4paper]{llncs}
+documentclass[runningheads,a4paper]{llncs}
 
@@
 First, we formalize the definition of a board. The board is an infinite grid graph with a nonnegative number of chips placed at each node
+\begin{figure}
 \begin{center}
@@
 \end{center}
+\caption{A finite component of the grid}
+\end{figure}
 
+\begin{figure}
+\label{cartesiangrid}
 \begin{center}
@@
 \end{center}
+\caption{A finite component of the grid with coordinates}
+\end{figure}
 We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$.
@@
 
-This function maps node labels given as integer coordinates to chip stack sizes. For example, FIGUREREFERENCE if a board $b$ has 8 chips at the node with coordinate $(3,1)$, then $b(3,1) = 8$.
+This function maps node labels given as integer coordinates to chip stack sizes.
+For example, in Figure \ref{cartesiangrid} the board $b$ has 8 chips at the node with coordinate $(3,1)$, so we write $b(3,1) = 8$.
 
@@
 \label{finitestability}
-If $b \in \mathcal{B}_F$, there exists $N> 0$ such that after $N$ earthquakes the resultant board will be stable, or for all $m > 0$ and all $s \in \mathbb{Z} \times \mathbb{Z}$, $E^{N+m}[b,s] = E^N[b,s]$.
+If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
 \end{theorem}
@@
 \begin{lemma}
-Suppose $b \in \mathcal{B}_F$. If $E[b, s] \neq b$, then $E^n[b, \{s_n\}] \neq b$ for all $n > 0$ and all sequences $\{s_n\}$.
+Suppose $b$ is a finite board. If $E_{s}b \neq b$, then no application of earthquakes to $E_{s}b$ can produce $b$. 
 \end{lemma}
@@
 \begin{proof}
-For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as $S: \mathcal{B}_F \rightarrow \mathbb{Z}_{\geq 0}$ 
+For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as a map from finite boards to the nonnegative integers
 \begin{equation*}
@@
 This is well-defined since the total number of chips on the board is finite.
-We claim
+We claim that
 \begin{equation*}
-E[b, s] \neq b \Rightarrow S(E[b, s]) > S(b)
+E_{s} b \neq b \Rightarrow S(E[b, s]) > S(b)
 \end{equation*}
@@
 
-We evaluate $S(E[b, s]) - S(b)$, recalling that $E[b, s](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
+We evaluate $S(E_{s} b) - S(b)$, recalling that $E_s b(i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
 \begin{align*}
-S(E[b, s]) - S(b) &=\sum_{i,j} E[b, s](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
+S(E_{s} b) - S(b) &=\sum_{i,j} E_{s} b (i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
 &=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
 \end{align*}
-For most two of the neighbors of $(s_1, s_2)$ 
+For at most two of the neighbors of $(s_1, s_2)$ 
 \begin{equation*}
@@
 
-Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that any number of earthquakes can transform a given finite board into.
+Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that earthquakes can transform a given finite board into.
 \begin{lemma}
@@
 Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then   $E^n[b, \{s_k\}]$ is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$for all $n$ and for all sequences $\{s_k\}$ .
+Then after applying any number of earthquakes the resultant board is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
 \end{lemma}
@@
 \begin{proof}
-Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E[b, s](i,j) \geq 1$.
-This is clear since any earthquake will move either bring things into the column $i$ or if it moves some chips out, then there are two chips that remain in that column.
+Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E_{s}b(i,j) \geq 1$.
+This is clear since an earthquake can bring things into column $i$ or if it moves some chips out, then there are two chips that remain in the column.
 
-The same is true for each row. This means that if a chip ever goes to a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
+The same is true for each row. This means that if a chip  is ever in a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
 \end{proof}
@@
 \begin{proof}
-(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_F$, so $\sum_{i,j} b(i,j) = C$.
-Because $b$ is a finite board we know the locations with a nonzero number of chips are bounded.
+(of Theorem~\ref{finitestability}) Recall that $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.
 Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then by Lemma~\ref{finiteextension}, we have that $E^n[b, \{s_k \}]$ is nonzero in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ s_k \}$ of length $n$.
+Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$.
 If we set
@@
 then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,.
-We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{s_k\}$, which is arbitrary after the first $N$ terms, such that $E^N[ b, \{s_k\}] = E^{N+m}[b, \{s_k\}]$.
+We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
 \end{proof}
 
-\subsection{Acyclicity of Infinite Boards}
-\label{Acyclicity of Infinite Boards}
-\emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\
-
-We can in fact show that infinite boards cannot cycle under earthquakes.
+We can use our result for finite boards to show that earthquakes will never cause infinite boards to revisit a prior arrangement.
 
 \begin{corollary}
-All boards are can never return to their same state.
+Suppose $b$ is an infinite board. If $E_s b \neq b$, then no application of earthquakes to $E_s b$ can produce $b$.
 \end{corollary}
@@
 \begin{proof}
-We have shown that finite boards never return to the same state.
-Suppose that an infinite board $b$ reurns to the same state, then it does so after $N$ earthquakes at locations $\{ s_k \}$.
-However, we can bound the size of the locations in $\{ s_k \}$ and so this infinite board acts in the same way that a finite board on this bound acts.
-Therefore, if the infinite board cycles, then the finite board cycles as well.
+We will proceed by contradiction.
+Assume otherwise, so there exists some sequence $\{s_1, s_2, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $E_s b$ the resultant board is $b$.
+However, we can bound the size of the locations affected by $\{s, s_1, s_2, \ldots, s_k \}$, call it $R$.
+This infinite board behaves the same way as a finite board with the same values on $R$ and 0 outside of $R$.
+But this means that earthquakes transformed a finite board back into a previous state which is a contradiction.
+Therefore infinite boards cannot be transformed back into a previous state.
 \end{proof}

### 2014-10-11 finaldraft.tex
adding example of earthquake image
improved some wording in the notations section

@@
-documentclass[runningheads,a4paper]{llncs}
+\documentclass[runningheads,a4paper]{llncs}
 
@@
 
-\begin{definition} We write boards as functions $b$ from $\mathbb{Z}^2 \to \mathbb{Z}_{\geq 0}$.
+\begin{definition} Boards are functions $b$ from $\mathbb{Z}^2 \to \mathbb{Z}_{\geq 0}$.
 \end{definition}
@@
 
-\begin{definition} We define $n(i,j)$ to be the set of neighbors of node $(i,j)$
+\begin{definition} Let $n(i,j)$ be the set of neighbors of node $(i,j)$
 \begin{equation*}
@@
 \begin{definition}
-An earthquake hits a board $b$ at node $s$, and the resulting board is denoted $E_sb$. If $b(s) \geq 4$, then
+Let $E_sb$ be the board that results from an earthquake hitting board $b$ at node $s$. If $b(s) \geq 4$, then
 \begin{align*}
@@
 
+\begin{figure}
+\centering
+\subfigure[Board before earthquake.]{\includegraphics[width=0.25\linewidth]{action1.pdf}\label{fig:boardbeforeearthquake}}
+\qquad\qquad
+\subfigure[Board after earthquake to active square]{\includegraphics[width=0.25\linewidth]{action2.pdf}\label{fig:boardafterequarthquake}}
+\caption{Earthquake applied to board. Green signifies active nodes}
+\label{fig:earthquakeexample}
+\end{figure}
+
 Often, a number of earthquakes will hit a board $b$.
@@
 \begin{definition}
-We say that a board is stable if all nodes contain values strictly less than 4. 
+A board is stable if all nodes contain values strictly less than 4. 
 \end{definition}
@@
 \begin{definition} 
-We say that a board is finite if the total number of chips on it is finite. Otherwise we say a board is infinite. 
+A board is finite if the total number of chips on it is finite. Otherwise we say a board is infinite. 
 \end{definition}

### 2014-10-11 finaldraft.tex
commented out global earthquakes section

@@
 \end{proof}
-
-\section{Global Earthquakes}
-\label{Global Earthquakes}
-\emph{written by Perry Kleinhenz, edited by Erik Waingarten and Fermi Ma}
-
-Now that we have shown that the order in which earthquakes occur does not matter we might wonder if the behavior of the system changes when all possible earthquakes occur at the \emph{same} time.
+%
+%
+%\section{Global Earthquakes}
+%\label{Global Earthquakes}
+%\emph{written by Perry Kleinhenz, edited by Erik Waingarten and Fermi Ma}
+%
+%Now that we have shown that the order in which earthquakes occur does not matter we might wonder if the behavior of the system changes when all possible earthquakes occur at the \emph{same} time.
 In this section we will formalize this notion and show that all finite boards stabilize under such earthquakes as well.
-
-We will call these earthquakes global earthquakes. A global earthquake (abbreviated as $GE$) is esentially to a standard earthquake applied to every square on the board simultaneously.
-\begin{definition} We define the global earthquake operator as a map from the set of all boards to the set of all boards and write $GE: \mathcal{B} \rightarrow \mathcal{B}$. 
-It transforms all of the squares on the board with the following rule
-\begin{align*}
-GE[b](i,j) = b(i,j) +\sum_{s \in \mathbb{Z} \times \mathbb{Z}} (E[b, s](i,j) - b(i,j)).
-\end{align*}
-That is the value of the board at a square after applying a global earthquake is equal to the value of the board originally plus the change in the value at the square that would occur from standard earthquakes affecting every square.
-\end{definition}
-
-In order to prove that finite boards stabilize under global earthquakes we will first show that global earthquakes are equivalent to some finite sequence of standard earthquakes.
+%
+%We will call these earthquakes global earthquakes. A global earthquake (abbreviated as $GE$) is esentially to a standard earthquake applied to every square on the board simultaneously.
+%\begin{definition} We define the global earthquake operator as a map from the set of all boards to the set of all boards and write $GE: \mathcal{B} \rightarrow \mathcal{B}$. 
+%It transforms all of the squares on the board with the following rule
+%\begin{align*}
+%GE[b](i,j) = b(i,j) +\sum_{s \in \mathbb{Z} \times \mathbb{Z}} (E[b, s](i,j) - b(i,j)).
+%\end{align*}
+%That is the value of the board at a square after applying a global earthquake is equal to the value of the board originally plus the change in the value at the square that would occur from standard earthquakes affecting every square.
+%\end{definition}
+%
+%In order to prove that finite boards stabilize under global earthquakes we will first show that global earthquakes are equivalent to some finite sequence of standard earthquakes.
 Before doing so we will introduce a definition to refer to the squares a global earthquake can effect.
-
-\begin{definition} The active sites $A_b$ of a board $b$ are all those squares which have at least 4 chips on them or written symbolically
-\begin{equation*}
-A_b=\{ (i,j); b(i,j) \geq 4 \}
-\end{equation*}
-\end{definition}
-
-We now show that a global earthquake is equivalent to applying standard earthquakes to each of the active sites on a board in any order.
-\begin{lemma} Suppose $b$ is a finite board. If we let arrange the terms of $A_b$ into a sequence $\{a_k\}$ with $N$ elements then 
-\begin{equation*}
-GE[b] = E^N[b, \{a_k\}]
-\end{equation*}
-\end{lemma}
-\begin{proof} 
-A standard earthquake is nontrivial only if the square it targets is contained in $A_b$. That is $E[b, s] = b$  if $s \notin A_b$. Therefore $GE[b](i,j)$ becomes
-\begin{equation*}
-GE[b](i,j) = b(i,j) +\sum_{s \in A_b} (E[b, s](i,j) - b(i,j)).
-\end{equation*}
-So a global earthquake can be considered finitely many earthquakes occurring at the same time.
+%
+%\begin{definition} The active sites $A_b$ of a board $b$ are all those squares which have at least 4 chips on them or written symbolically
+%\begin{equation*}
+%A_b=\{ (i,j); b(i,j) \geq 4 \}
+%\end{equation*}
+%\end{definition}
+%
+%We now show that a global earthquake is equivalent to applying standard earthquakes to each of the active sites on a board in any order.
+%\begin{lemma} Suppose $b$ is a finite board. If we let arrange the terms of $A_b$ into a sequence $\{a_k\}$ with $N$ elements then 
+%\begin{equation*}
+%GE[b] = E^N[b, \{a_k\}]
+%\end{equation*}
+%\end{lemma}
+%\begin{proof} 
+%A standard earthquake is nontrivial only if the square it targets is contained in $A_b$. That is $E[b, s] = b$  if $s \notin A_b$. Therefore $GE[b](i,j)$ becomes
+%\begin{equation*}
+%GE[b](i,j) = b(i,j) +\sum_{s \in A_b} (E[b, s](i,j) - b(i,j)).
+%\end{equation*}
+%So a global earthquake can be considered finitely many earthquakes occurring at the same time.
 But now if we consider $E^N[b, \{a_k\}]$, because the order of earthquakes does not matter the change in the value at some $(i,j)$ between the application of the $m+1$th and $m$th earthquake is equal to the change in the value at $(i,j)$ if the $m+1$th earthquake was applied directly to the original board.
 That is
-\begin{equation*}
-E^{m+1}[b, \{a_k\}](i,j) - E^{m}[b, \{a_k\}](i,j) = E[b, a_{m+1} ](i,j) - b(i,j)
-\end{equation*}
-Therefore 
-\begin{equation*}
-E^N[b, \{a_k\}](i,j) = b(i,j) + \sum_{a_k \in A_b} (E[b, a_k](i,j) - b(i,j)),
-\end{equation*}
-but this is the same expression as the value at $(i,j)$ after a global earthquake effects the board $b$. 
-\end{proof}
-
-\begin{theorem} Suppose $b$ is a finite board, then there exists $N$ such that for all $m>0$ 
-\begin{equation*}
-GE^{N+m}[b] = GE^{N}[b]
-\end{equation*}
-\end{theorem}
-\begin{proof}
-Assume otherwise.
+%\begin{equation*}
+%E^{m+1}[b, \{a_k\}](i,j) - E^{m}[b, \{a_k\}](i,j) = E[b, a_{m+1} ](i,j) - b(i,j)
+%\end{equation*}
+%Therefore 
+%\begin{equation*}
+%E^N[b, \{a_k\}](i,j) = b(i,j) + \sum_{a_k \in A_b} (E[b, a_k](i,j) - b(i,j)),
+%\end{equation*}
+%but this is the same expression as the value at $(i,j)$ after a global earthquake effects the board $b$. 
+%\end{proof}
+%
+%\begin{theorem} Suppose $b$ is a finite board, then there exists $N$ such that for all $m>0$ 
+%\begin{equation*}
+%GE^{N+m}[b] = GE^{N}[b]
+%\end{equation*}
+%\end{theorem}
+%\begin{proof}
+%Assume otherwise.
 That is for some board $b, GE^N[b]$ always has at least one square with at least four chips on it.
 But applying a global earthquake is equivalent to applying a sequence of standard earthquakes and so we have a board which does not stabilize under standard earthquakes which is a contradiction.
-\end{proof}
+%\end{proof}
 

### 2014-10-11 finaldraft.tex
starting computation section

--- 2014-10-11 finaldraft.tex
@@
 \end{definition}
+
+\begin{figure}
+\centering
+\subfigure[Board before earthquake.]{\includegraphics[width=0.25\linewidth]{action1.pdf}\label{fig:boardbeforeearthquake}}
+\qquad\qquad
+\subfigure[Board after earthquake to active square]{\includegraphics[width=0.25\linewidth]{action2.pdf}\label{fig:boardafterequarthquake}}
+\caption{Earthquake applied to board. Green signifies active nodes}
+\label{fig:earthquakeexample}
+\end{figure}
 
@@
 
-\subsection{Model of Computation}
+\subsection{Simple Model}
+
+A simple model of computation can be defined as follows.
+We assume a finite board.
+We specify some input nodes and some output nodes.
+If an input bit is to be set to 1, then we place 4 chips in the input node.
+We apply earthquakes to the board until it stabilizes.
+If an output node ever had an earthquake, then we mark that output bit as 1; otherwise, the output bit is set to 0.
+
+A simple example is a board that computes the identity function of one output. We can compute 
+\[ f(x) = x \]
+By having a board with no chips, and we let the input node be the same as the output node.
+Then if $x = 0$, this means that the input node will not contain any chips, so the board is stable.
+The output node never received an earthquake, so the output is 0.
+If $x = 1$, then the input node gets 4 chips and an earthquake affects it.
+So the output node is activated, and so the output is 1.
+With this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
+
+\begin{figure}
+\centering
+\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wire.pdf}\label{fig:wire}}
+\qquad\qquad
+\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andgate.pdf}\label{fig:AND}}
+\qquad\qquad
+\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
+\caption{Gates in the simple model. Green are the inputs, red are the outputs}
+\label{fig:simplegates}
+\end{figure}
+
+Composing these gates, we can make circuits which represent some Boolean function. In particular we can compute any Boolean function that can be represented with a planar circuit with only AND and OR.
+
+\begin{theorem}
+There does not exists a board which computes $f(x) = \neg x$.
+\end{theorem}
+
+\begin{proof}
+Suppose for contradiction that we have a board $b$ with one input $i$ and one output $o$ which computes NOT.
+Then we know that there exists some node which initially has over $4$ chips, since otherwise, the board would be stable and so the output never gets activated.
+\end{proof}
 

### 2014-10-11 finaldraft.tex
starting computation section 2

--- 2014-10-11 finaldraft.tex
@@
 
-\subsection{Model of Computation}
+\subsection{Simple Model}
+
+A simple model of computation can be defined as follows.
+We assume a finite board.
+We specify some input nodes and some output nodes.
+If an input bit is to be set to 1, then we place 4 chips in the input node.
+We apply earthquakes to the board until it stabilizes.
+If an output node ever had an earthquake, then we mark that output bit as 1; otherwise, the output bit is set to 0.
+
+A simple example is a board that computes the identity function of one output. We can compute 
+\[ f(x) = x \]
+By having a board with no chips, and we let the input node be the same as the output node.
+Then if $x = 0$, this means that the input node will not contain any chips, so the board is stable.
+The output node never received an earthquake, so the output is 0.
+If $x = 1$, then the input node gets 4 chips and an earthquake affects it.
+So the output node is activated, and so the output is 1.
+With this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
+
+\begin{figure}
+\centering
+\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wire.pdf}\label{fig:wire}}
+\qquad\qquad
+\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andgate.pdf}\label{fig:AND}}
+\qquad\qquad
+\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
+\caption{Gates in the simple model. Green are the inputs, red are the outputs}
+\label{fig:simplegates}
+\end{figure}
+
+Composing these gates, we can make circuits which represent some Boolean function. In particular we can compute any Boolean function that can be represented with a planar circuit with only AND and OR.
+
+\begin{theorem}
+There does not exists a board which computes $f(x) = \neg x$.
+\end{theorem}
+
+\begin{proof}
+Suppose for contradiction that we have a board $b$ with one input $i$ and one output $o$ which computes NOT.
+Then we know that there exists some node which initially has over $4$ chips, since otherwise, the board would be stable and so the output never gets activated.
+
+Let $A(0)$ be the set of nodes that get activated when $i = 0$ and $A(1)$ be the set of nodes that get activated when $i=1$.
+Then we claim that $A(0) \subset A(1)$.
+This would yield a contradiction since $o \in A(0)$ but $o \notin A(1)$.
+\end{proof}
 

### 2014-10-11 finaldraft.tex
finished simple model computation section

@@
 \begin{theorem}
+\label{thm:order}
 Let $b$ be a board that becomes a stable board $b$ after $N$ earthquakes.
@@
 
+\begin{lemma}
+Let $A(i)$ be the set of active nodes in an execution of board $b$ with input node $i$. Then $A(0) \subset A(1)$.
+\end{lemma}
+
+\begin{proof}
+By Theorem~\ref{thm:order}, we can simulate an execution of the board with input be 0 before we apply earthquakes to the input node.
+We start with a board $b$ where the input is 1, so it has 4 chips.
+We simulate by applying earthquakes to the active nodes that are not the input.
+If the input node ever receives more than $8$ chips, we apply one earthquake to the input node.
+This way, we guarantee that the input node always contains more than $4$ chips.
+In addition, the input node does not contain more than $7$ chips for more than one consecutive earthquake.
+We do this until every node except the input node has less than 4 chips.
+
+Note that this execution is indistinguishable from the execution with input set to 0.
+This means that all nodes activated in the execution with input 0 are the same as the nodes activated in the above mentioned execution.
+
+Once this is done, then we continue with the execution applying the earthquake to the input node.
+By Theorem~\ref{thm:order}, the final configuration will be the same as any execution of the board.
+So $A(0) \subset A(1)$.
+\end{proof}
+
+\subsection{More powerful model}
+
 Our model of computation is as follows: we assume a finite board with finitely many chips.

### 2014-10-11 finaldraft.tex
finished computation section

@@
 \subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
-\caption{Gates in the simple model. Green are the inputs, red are the outputs}
+\caption{Gates in the simple model. Green are the inputs, red are the outputs. We disregard edges and assume that any two nodes that are touching have an edge between them.}
 \label{fig:simplegates}
@@
 
-Our model of computation is as follows: we assume a finite board with finitely many chips.
-The inputs are designated by each two squares, one apart one is the ``clock" square and other the bit square.
-Likewise, the outputs each have a ``clock" square and a bit square.
-The ``clock" squares are set the value 4 and the bit squres are set to 4 if they are 1 and 0 otherwise.
-A computation is done by a series of earthquakes, when an output has the clock square active, then the bit square is read, and that output bit is considered read.
+As before, we assume a finite board with finitely many chips.
+The inputs are designated by having one input node.
+The outputs also each have one node.
+We now make the model of computation stronger by adding a special node $r$.
+When $r$ gets activated, it ends the execution and reads the outputs.
+We assume that in one ``round" we apply an earthquake to each valid node and that $r$ is the last node which gets applied an earthquake.
 
-\subsection{Example}
+In order to make it easier to compute, we will make a special \textit{clock} node, designated as $c$ which always starts out with $4$ chips.
+We will have the clock go through each input and travel in parallel to each input.
+We will build \textit{gadgets} that allow us to compute with the clock so that we can build gates in a modular and then put them together.
 
-A simple example is given by a board that computes the identity.
+We will also make a wire path that takes $c$ to $r$, so that we know exactly at what point the outputs are read.
+
+As shown before, simple example is given by a board that computes the identity.
+For simplicity, we write the boards as a board of numbers, which every space is a node and two nodes have an edge between them if they are either above, below, left, or right.
 
 \[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
-				     4_c & 3 & 3 & 3 & 3 & 3_e \\
+				     4_c & 3 & 3 & 3 & 3 & 3_r \\
 				     0 & 0 & 0 & 0 & 0 & 0 \\
@@
 				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
-Where the subscript $c$ indicates the clock, the subscript $e$ indicates the output clock square. $i$ indicates the bit square, and $o$ indicates the output bit.
+Where $i$ indicates the input bit node, and $o$ indicates the output bit.
 
-One can see that $e$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $e$.
+One can see that $r$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $r$.
 Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active.
@@
 Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
-this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $e$ is the output clock and $o$ is the output bit.
+this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $r$ is the output read and $o$ is the output bit.
 
-\[ \begin{array}{cccccc} 0 & 0 & e & 0 & o \\
+\[ \begin{array}{cccccc} 0 & 0 & r & 0 & o \\
 				     0 & 3 & 2 & 0 & 3 \\
@@
 
-We can synchronize the inputs so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
+We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
 \[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
@@
 					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
-					 0 & 0 & 0 & 0 & o & 0 & e & 3 \end{array} \]
+					 0 & 0 & 0 & 0 & o & 0 & r & 3 \end{array} \]
 Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
@@
 
-So currently, we can compute any boolean circuit which can be drawn in a planar graph with only ANDs and ORs. So the question is whether we can compute the NOT? We argue that this is not possible.
+We can build a NOT board with the following circuit:
+\begin{figure}
+\centering
+\includegraphics[width=0.25\linewidth]{notgate.pdf}
+\caption{Possible board to compute $f(x) = \neg x$}
+\label{fig:NOT}
+\end{figure}
+In Figure~\ref{fig:NOT}, the wire to $c$ and $r$ is the same length as the wire from $c$ to $o$, but the wire from $i$ to $r$ is shorter than the length of the wire from $c$ to $o$.
+If $i$ is set to $1$, then $r$ will be activated before the clock reaches $o$, so $o$ will be read as 0.
+If $i$ is set to 0, then the clock will reach $o$ and $r$ at the same time, so the output is $1$.
+This computes $f(x) = \neg x$; however, we have switched the order of $r$ and $o$.
+We also do not know how to compose this board with other boards.
 
-\begin{theorem}
-There does not exists any board configuration which computes the NOT operator in the above model of computation.
-\end{theorem}
-
-\begin{proof}
-Suppose we have a board configuration $B$ with input squares $c$, $i$ and output squares $e, o$ that computes the NOT function.
-Then we can make another board $B'$ that synchronizes the final clock with the input so that we do not need to worry about timing of the clock.
-That is, we can guarantee that if $o$ is active, then $o$ is active at the same time $e$ is active.
-
-Now lets make $B'$ an infinite board by putting 0s on the sides.
-So we have a finite board with finitely many chips.
-Let $S_{c_0i_0}$ be the set of squares which were active (had an earthquake affect them) during the execution of $B$ with the input value $i = i_0$ and $c = c_0$.
-
-The first claim to make is that $S_{c_0i_0} \subset S_{c_1i_1}$ if $c_0 \leq c_1$ and $i_0 \leq i_1$.
-This is clearly true since the order of redistribution does not matter.
-In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with local earthquakes.
-
-However, this arises in a contradiction, since $o \in S_{10}$ but $o \notin S_{11}$. Therefore, computing NOT with a board is not possible.
-\end{proof}
-
-What does this mean?
-Well, in particular, it means that we cannot do much.
-In particular, we can't add, since $x + 1$ will act as the NOT operator on the lowest order bit.
-We can only compute boolean circuits in planar graphs.
+We would also like to build a crossover gadget to allow us to compute circuits which are not necesarily.
 

### 2014-10-11 finaldraft.tex
started changing stability section. rewrote first three or four paragraphs

--- 2014-10-11 finaldraft.tex
@@
 
+Generally, when we refer to an earthquake, we restrict our attention to earthquakes that actually change the state of the board.
+
 Often, a number of earthquakes will hit a board $b$.
@@
 
-In this section we show that any finite board will reach a stable state after finitely many earthquakes.
-We do so by showing that if a earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
-We will then give a bound for the maximum number of configurations that earthquakes can transform a given board into. 
-Together these show that all finite boards stabilize. 
+In this section, we show that any finite board will eventually become stable after a finite number of earthquakes.
+We first introduce a spread function that maps boards to positive numbers.
+We demonstrate that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversably changes a board configuration.
+In other words, we can never return to a given board arrangement.
+
+We then give a bound on the maximum number of board configurations a finite board can ever reach.
+We will argue that since a finite number of boards are reachable and board states are never revisited, that eventually this process must terminate.
+Thus, we eventually reach a stable board where no earthquakes are allowable.
+
+To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$.
+Then the spread function is
+\[ \psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 \]
+
+The motivation for this definition is that the spread function should give a measure of how ``spread out" the chips on a board are.
+Intuitvely, earthquakes have the function of moving chips that are condensed in a single piile and spreading them out over a greater area.
+This function, which sums up the chip stack sizes weighted by the taxicab distance squared to the origin, sufficiently captures the idea of spread.
+
 \begin{theorem}
@@
 
-\subsection{Simple Model}
-
-A simple model of computation can be defined as follows.
-We assume a finite board.
-We specify some input nodes and some output nodes.
-If an input bit is to be set to 1, then we place 4 chips in the input node.
-We apply earthquakes to the board until it stabilizes.
-If an output node ever had an earthquake, then we mark that output bit as 1; otherwise, the output bit is set to 0.
-
-A simple example is a board that computes the identity function of one output. We can compute 
-\[ f(x) = x \]
-By having a board with no chips, and we let the input node be the same as the output node.
-Then if $x = 0$, this means that the input node will not contain any chips, so the board is stable.
-The output node never received an earthquake, so the output is 0.
-If $x = 1$, then the input node gets 4 chips and an earthquake affects it.
-So the output node is activated, and so the output is 1.
-With this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
-
-\begin{figure}
-\centering
-\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wire.pdf}\label{fig:wire}}
-\qquad\qquad
-\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andgate.pdf}\label{fig:AND}}
-\qquad\qquad
-\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
-\caption{Gates in the simple model. Green are the inputs, red are the outputs}
-\label{fig:simplegates}
-\end{figure}
-
-Composing these gates, we can make circuits which represent some Boolean function. In particular we can compute any Boolean function that can be represented with a planar circuit with only AND and OR.
-
-\begin{theorem}
-There does not exists a board which computes $f(x) = \neg x$.
-\end{theorem}
-
-\begin{proof}
-Suppose for contradiction that we have a board $b$ with one input $i$ and one output $o$ which computes NOT.
-Then we know that there exists some node which initially has over $4$ chips, since otherwise, the board would be stable and so the output never gets activated.
-
-Let $A(0)$ be the set of nodes that get activated when $i = 0$ and $A(1)$ be the set of nodes that get activated when $i=1$.
-Then we claim that $A(0) \subset A(1)$.
-This would yield a contradiction since $o \in A(0)$ but $o \notin A(1)$.
-\end{proof}
+\subsection{Model of Computation}
 

### 2014-10-11 finaldraft.tex
modified half of stability section, up to bound on board size

@@
 
-The motivation for this definition is that the spread function should give a measure of how ``spread out" the chips on a board are.
-Intuitvely, earthquakes have the function of moving chips that are condensed in a single piile and spreading them out over a greater area.
-This function, which sums up the chip stack sizes weighted by the taxicab distance squared to the origin, sufficiently captures the idea of spread.
-
-\begin{theorem}
-\label{finitestability}
-If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
-\end{theorem}
-
-In order to show finite boards never revisit a prior arrangement, we define a spread function that maps boards to nonnegative integers.
-The spread function should give some measure of how ``spread out" the chips on a board are.
-Earthquakes seem to ``spread out" the chips on a board and so we would like earthquakes to increase the value of the function.
+Thus, the function sums up the chip stack sizes weighted by the taxicab distance to the origin squared.
+The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are.
+Intuitvely, earthquakes move chips that are condensed in a single pile and spreads them out over a greater area.
+We capture this in the following lemma:
 
 \begin{lemma}
-Suppose $b$ is a finite board. If $E_{s}b \neq b$, then no application of earthquakes to $E_{s}b$ can produce $b$. 
+\label{spreadfunction}
+Suppose an earthquake hits node $s$ of a board $b$, producing the board $E_sb$. Then
+\[ \Psi(E_sb) > \Psi(b) \]
+so the earthquake strictly increases the spread of the board.
 \end{lemma}
+\begin{proof}
+The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.
 
+Let the taxicab distance to $s$ is $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.
+
+Out of the 4 nodes in $n(s)$, some have a taxicab distance of $d_s+1$, and the rest are at distance $d_s-1$.
+In particular, if $s$ is not exactly on the $x$ or $y$ axis, exactly 2 neighboring nodes are at $d_s+1$, and exactly 2 are at distance $d_s-1$.
+If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$.
+Since we want to lower bound the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.
+
+Suppose that the two neighboring nodes at distance $d_s+1$ have chip stacks of $n_1$ and $n_2$ originally.
+Then their contribution to the spread function goes from $n_1(d_s+1)^2$ and $n_2(d_s+1)^2$ to $(n_1+1)(d_s+1)^2$ and $(n_2+1)(d_s+1)^2$.
+If the other two nodes at distance $d_s-1$ have chip stacks of $n_3$ and $n_4$, their contribution to the spread function goes from $n_3(d_s-1)^2$ and $n_4(d_s-1)^2$ to $(n_3+1)(d_s-1)^2$ and $(n_4+1)(d_s-1)^2$.
+
+Omitting the algebraic details, we find that summing up the changes in the spread function gives
+\[ \Psi(E_sb) - \Psi(b) = 4\]
+which completes the proof.
+\end{proof}
+
+\begin{corollary}
+\label{finitestability}
+A board can never return to a previous state via a sequence of earthquakes.
+\end{corollary}
 \begin{proof}
-For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as a map from finite boards to the nonnegative integers
-\begin{equation*}
-S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
-\end{equation*}
-We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared.
-This is well-defined since the total number of chips on the board is finite.
-We claim that
-\begin{equation*}
-E_{s} b \neq b \Rightarrow S(E[b, s]) > S(b)
-\end{equation*}
-That is if an earthquake changes the board, $S$ increases.
-This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.
-
-We evaluate $S(E_{s} b) - S(b)$, recalling that $E_s b(i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
-\begin{align*}
-S(E_{s} b) - S(b) &=\sum_{i,j} E_{s} b (i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
-&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
-\end{align*}
-For at most two of the neighbors of $(s_1, s_2)$ 
-\begin{equation*}
-(|i|+|j|)= (|s_1|+|s_2|-1)^2,
-\end{equation*}
-and for at least two of the neighbors 
-\begin{equation*}
-(|i|+|j|)=(|s_1|+|s_2|+1)^2,
-\end{equation*}
- and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
-\begin{align*}
-&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
-& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
-\end{align*}
+Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$.
+Lemma~\ref{spreadfunction} shows that $\psi(b) > \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
 \end{proof}

### 2014-10-11 finaldraft.tex
Fixed some errors
Also started working on bringing d tree section in line on notation.

@@
 Then the spread function is
-\[ \psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 \]
+\begin{equation*}
+\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 
+\end{equation*}
 
@@
 The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are.
-Intuitvely, earthquakes move chips that are condensed in a single pile and spreads them out over a greater area.
+Intuitively, earthquakes takes chips that are condensed in a single pile and spreads them out over a greater area.
 We capture this in the following lemma:
@@
 \label{spreadfunction}
-Suppose an earthquake hits node $s$ of a board $b$, producing the board $E_sb$. Then
-\[ \Psi(E_sb) > \Psi(b) \]
+Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb$. Then
+\begin{equation*}
+\Psi(E_sb) > \Psi(b) 
+\end{equation*}
 so the earthquake strictly increases the spread of the board.
@@
 
-Let the taxicab distance to $s$ is $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.
+Let the taxicab distance to $s$ be $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.
 
@@
 If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$.
-Since we want to lower bound the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.
+Since we only need a lower bound for the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.
 
@@
 Omitting the algebraic details, we find that summing up the changes in the spread function gives
-\[ \Psi(E_sb) - \Psi(b) = 4\]
+\begin{equation*}
+ \Psi(E_sb) - \Psi(b) = 4
+ \end{equation*}
 which completes the proof.
@@
 \begin{corollary}
-\label{finitestability}
-A board can never return to a previous state via a sequence of earthquakes.
+A finite board can never return to a previous state via a sequence of earthquakes.
 \end{corollary}
@@
 Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$.
-Lemma~\ref{spreadfunction} shows that $\psi(b) > \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
+Lemma~\ref{spreadfunction} shows that $\psi(b) < \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
 \end{proof}
@@
 
-We are now ready to prove the theorem that all finite boards will reach a stable state.
+We are now ready to show that all finite boards will reach a stable state.
 
+\begin{theorem} If $b$ is a finite board then there exists some sequence $\{s_1, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $b$ the resultant board is stable. 
+\end{theorem}
 \begin{proof}
-(of Theorem~\ref{finitestability}) Recall that $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.
+ Since $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.
 Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
@@
 We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
-We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
-\begin{equation*}
-E^{m}[b, \{s_k\}](s_{m+1}  \geq 4
-\end{equation*}
-for all $1\geq m+1 \leq n$.
+We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid when applied one after another. 
 \end{definition}
@@
 \begin{equation*}
-E[b, s](x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
+E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
 \end{equation*}
@@
 \begin{equation*}
-E_2[E_1[b, s_1] s_2] = E_1[LE_2[b, s_2], s_1] 
+E_2 E_1 b = E_1 E_2 b
 \end{equation*}
@@
 
-We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[b, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
+We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
 Therefore we must have $b(s_1)<8$ and $s_2=s_1$.
-But then $E_1[b, s_1](s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes.
+But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes.
 This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.
 
-Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$.
+Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$.
 Applying Lemma ~\ref{earthquakeredefine} we have
 \begin{align*}
-E_2[E_1[b, s_1], s_2](x) = E_1[b, s_1])(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
+E_2 E_1 b (x) = E_1(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
 = b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
 = b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
-= E_2[b, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
-= E_1[E_2 [b, s_2], s_1](x) 
+= E_2 b(x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
+= E_1 E_2 b(x) 
 \end{align*}
@@
 \end{definition}
+\begin{figure}[!h]
+\label{4tree}
+\centering
+\includegraphics[width=0.4\linewidth]{4tree}
+\caption{A 4-tree}
+\end{figure}
 We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
@@
 \end{definition}
+\begin{figure}[!h]
+\label{joined4tree}
+\centering
+\includegraphics[width=0.8\linewidth]{joined4tree}
+\caption{A joined 4-tree}
+\end{figure}
 
@@
 \begin{definition} 
-We define the function $n: G_d \rightarrow G_d^d$ as 
+Let $n(i,j)d$ be the set of neighbors of node $(i,j)$ 
 \begin{equation*}
@@
 We now formalize the board of poker chips.
-Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate in the integer lattice.
+Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
 \begin{definition}
@@
 Recalling that $n(s)$ refers to the neighbors of $s$ we define our earthquake operation in terms of $d$-boards.
+It functions analogously to the earthquake for the standard board, but instead of hitting nodes that have more than 4 chips it hits nodes that have more than $d$ chips.
 \begin{definition}
@@
 \begin{align*}
-E[b, s](s) = b(s)- d \\
-E[b, s](t) = b(t) +1 
+E_s b(s) = b(s)- d \\
+E_s b(t) = b(t) +1 \quad \forall t \in n(s)
 \end{align*}
-for $t \in n(s)$.
-Otherwise $b(s)<d$ and $E[b,s]=b$ as the earthquake does not move any chips.
-We use the notation $E^r[b, \{s_k\}]$ to denote the application of $E$ to the $b$ $r$ times, to the sites $s_1, \ldots s_r$ respectively.
+Otherwise $b(s)<d$ and $E_s b=b$ as the earthquake does not move any chips.
 \end{definition}
@@
 \begin{definition}
-We say that a $d$-board is stable if its values are strictly less than $n$.
+We say that a $d$-board is stable if its values are strictly less than $d$.
 \end{definition}
@@
 \begin{definition} 
-We say that a $d$-board is finite if the total number of chips on it is finite. That is a $d$board $b$ is finite if the sum
-\begin{equation}
-C= \sum_{(i,j) \in G_d} b(i,j) 
-\end{equation}
-is finite. 
-We denote the set of finite $d$-boards as $B(d)_F$.
+We say that a $d$-board is finite if the total number of chips on it is finite.
 \end{definition}
@@
 \begin{definition}
- Let $\Psi: \mathcal{B}(d)_F \rightarrow \mathbb{Z}_{\geq 0}$ be the function
+To define the spread function we once again define a distance on our graph.  The distance for a node $s$ 
+$\Psi$ is a function from the finite $d$-boards and is defined
 \begin{equation}
-\Psi(f) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
+\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
 \end{equation}

### 2014-10-12 finaldraft.tex
Finished changing notation for d-Trees

@@
 \begin{corollary}
+\label{norepeats}
 A finite board can never return to a previous state via a sequence of earthquakes.
@@
 \end{equation}
-We write $\mathcal{B}(d)$ to refer to the set of all $d$-boards.  
 \end{definition}
@@
 \begin{definition}
-An earthquake takes as an input an $d$-board and a node and outputs an $d$-board so we write $E: \mathcal{B}(d) \times G_d \rightarrow \mathcal{B}(d)$. If $s$ is the node input and $b(s)>d$ then 
+An earthquake takes as an input an $d$-board and a node and outputs an $d$-board. If $s$ is the node input and $b(s)>d$ then 
 \begin{align*}
-E_s b(s) = b(s)- d \\
-E_s b(t) = b(t) +1 \quad \forall t \in n(s)
+&E_s b(s) = b(s)- d \\
+&E_s b(t) = b(t) +1 \quad \forall t \in n(s)
 \end{align*}
@@
 \begin{definition}
-To define the spread function we once again define a distance on our graph.  The distance for a node $s$ 
-$\Psi$ is a function from the finite $d$-boards and is defined
+$\Psi$ is a function from the finite $d$-boards and is defined as
 \begin{equation}
@@
 \end{equation}
-Note that this is well defined for trees which contain a finite number of chips. 
+Note that this is well defined for boards which contain a finite number of chips. 
 \end{definition} 
@@
 \begin{lemma} 
-If $E[b, s_1] \neq b$, then $E^r[b, \{s_k\}] \neq b$ for all $r>0$ and all sequences $\{s_k\}$ of length $r$ that begin with $s_1$. 
+Suppose an earthquake hits node $s$ of a finite $d$-board $b$ producing the board $E_s b$. Then
+\begin{equation*}
+\Psi( E_s b) > \Psi (b)
+\end{equation*}
 \end{lemma}
-
 \begin{proof}
-For ease of notation let $s=(s_1, s_2)$. We will show that if $E$ acts on a board nontrivially then $\Psi(E[b, s])$ will be strictly larger. That is  
-\begin{equation}
-E[b, s] \neq f \Rightarrow \Psi(E[b, s]) > \Psi(b)
-\end{equation}
-
-We can evaluate $\Psi(E[b, s]) - \Psi(b)$.
+For ease of notation let $s=(s_1, s_2)$. We can evaluate $\Psi(E_s b) - \Psi(b)$.
 \begin{align}
-\Psi(E[b, s])-\Psi(b) = \sum_{(i,j) \in G_d} E[b, s](i,j)|i|^2 - \sum_{(i,j) \in G_d} b(i,j)|i|^2 
+\Psi(E[b, s])-\Psi(b) = \sum_{(i,j) \in G_d} E_s b (i,j)|i|^2 - \sum_{(i,j) \in G_d} b(i,j)|i|^2 
 \end{align}
@@
 \begin{align*}
-\Psi(E[b, s]) - \Psi(b) &= E[b, s](s) |s_1|^2 - b(s) |s_1|^2 + \sum_{(i,j) \in n(s)} (E[b,s](i,j) |i|^2 - b(i,j)|i|^2) \\
+\Psi(E_s b ) - \Psi(b) &= E_s b (s) |s_1|^2 - b(s) |s_1|^2 + \sum_{(i,j) \in n(s)} (E_s b (i,j) |i|^2 - b(i,j)|i|^2) \\
 &= -d |s_1|^2 + \sum_{(i,j) \in n(s)} |i|^2
@@
 A point to emphasize is that this is the only substantial way in which this proof differs from the similar proof for the square lattice.
-Our expression simplifies to
+Omitting the algebraic details we find that
 \begin{align*}
-&= -d |s_1|^2 + \sum_{(i,j) \in n(s)}-|i|^2 \\
-&= - d|s_1|^2 + (d-1)(|s_1| +1)^2 +(|s_1| -1)^2 \\
-&= d + (d-2)2|s_1| >0
+\Psi( E_s b) - \Psi(b) > d >0
 \end{align*}
+which completes the proof.
+\end{proof}
+
+\begin{lemma}
+A finite $d$-board can never return to a previous state via a sequence of earthquakes
+\end{lemma}
+\begin{proof}
+The proof is identical to that of Corollary ~\ref{norepeats}
 \end{proof}
@@
 Suppose $\sum_{x,y} f(x,y) = C$, and the first coordinate of all nodes which are nonzero are in $[i_{min}, i_{max}]$.
-Then $E^r[b, \{s_k\}]$ only contain chips within the range $[i_{min} - 2C, i_{max} + 2C]$ for all $r>0$ and all sequences $\{s_k\}$ of length $r$.
+Then after applying any number of earthquakes the first coordinates of the nonzero nodes of the resultant board will be in range $[i_{min} - 2C, i_{max} + 2C]$.
 \end{lemma}
@@
 But this cannot occur because in order to move a chip to a node with a lower first coordinate, one chip must moved to a node with a higher first coordinate.
-Therefore $E^r[b, \{s_k\}]$ does not have a node with a nonzero number of chips on it with first coordinate less than $i_{min} -2C$.
+Therefore the resultant board does not have a node with a nonzero number of chips on it with first coordinate less than $i_{min} -2C$.
 
-An analogous argument shows that if $i_{max}$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not have a node with a nonzero number of chips on it with first coordinate greater than $i_{max}+2C$.
+An analogous argument shows that  and after applying any sequences of earthquakes to $b$ the resultant board does not have a node with a nonzero number of chips on it with first coordinate greater than $i_{max}+2C$.
 \end{proof}
 
-So now we are ready to prove the theorem that all finite boards will reach a stable state.
-
+\begin{theorem} If $b$ is a finite $d$-board then there exists some sequence $\{s_1, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $b$ the resultant $d$-board is stable.
+\end{theorem}
 \begin{proof}
-(of Theorem~\ref{finitestability}) Suppose $b \in \mathcal{B}(d)_f$ is a finite board, so $\sum_{(i,j) \in G_n} b(i,j) = C$, and suppose all chips lie between levels  $[i_{min}, i_{max}]$.
+ Suppose $b \in \mathcal{B}(d)_f$ is a finite board, so $\sum_{(i,j) \in G_n} b(i,j) = C$, and suppose all chips lie between levels  $[i_{min}, i_{max}]$.
 Then by Lemma ~\ref{finiteextensiontree}, we have that for all $r$ and sequence $\{s_k\}$ all chips in $E^r[b, \{s_k\}]$ will have first coordinate in $ [i_{min} -2C, i_{max} +2C]$.
@@
 \end{equation}
-possible boards $b \in \mathcal{B}(n)_f$ that satisfy this.
-We know that there are no cycles in the states, so there must be an $0 \leq N < \binom{N+S-1}{N-1}$ and some sequence $\{s_k\}$ such that $E^N[b, \{s_k\}] = E^{N+m}[b, \{s_k\}]$ for all $m \geq 0$.
+possible boards $b \in \mathcal{B}(n)_f$ that satisfy this. We know that there are no cycles in the states, so eventually the board must stabilize.
 \end{proof}

### 2014-10-12 finaldraft.tex
Fixed stuff in d-Trees

@@
 \begin{proof}
- Suppose $b \in \mathcal{B}(d)_f$ is a finite board, so $\sum_{(i,j) \in G_n} b(i,j) = C$, and suppose all chips lie between levels  $[i_{min}, i_{max}]$.
-Then by Lemma ~\ref{finiteextensiontree}, we have that for all $r$ and sequence $\{s_k\}$ all chips in $E^r[b, \{s_k\}]$ will have first coordinate in $ [i_{min} -2C, i_{max} +2C]$.
+ Since $b$ is finite we know $\sum_{(i,j) \in G_n} b(i,j) = C$.
+Suppose all chips lie between levels  $[i_{min}, i_{max}]$.
+Then by Lemma ~\ref{finiteextensiontree}, we have that after applying any sequence of earthquakes to $b$ the any node of the resultant board with a nonzero number of chips on it has a first coordinate contained in $ [i_{min} -2C, i_{max} +2C]$.
 
@@
 \end{equation}
-nodes that are potentially occupied. Therefore there are at most 
+nodes that could have chips on them. Therefore there are at most 
 \begin{equation}
@@
 \end{equation}
-possible boards $b \in \mathcal{B}(n)_f$ that satisfy this. We know that there are no cycles in the states, so eventually the board must stabilize.
+possible boards that satisfy this. We know that there are no cycles in the states, so eventually the board must stabilize.
 \end{proof}

### 2014-10-12 finaldraft.tex
Made Changes to Time Until Stability

@@
 
-Another question of interest is how many global earthquakes it would take for a finite board to reach a stable state.
-We will first focus on the case of the board with only one stack containing $n$ chips.
+Another question of interest is how many global earthquakes must be applied to a finite board in order for it to stabilize.
+We will confine our analysis to the case of the board with only one stack containing $n$ chips.
+We are able to demonstrate upper bounds on the number of earthquakes it takes to stabilize the board and we conjecture on the existence of tighter ones.
 
@@
 \begin{definition}
-We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes.
-We also define $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
+We define $T(n)$ to be the number of global earthquakes that must be applied to the board for it to stabilize.
 \end{definition}
-Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$, since that many earthquakes will reduce the number of chips on $(0,0)$ to less than $4$.
 
-We can get another upper bound for $T(n)$ by looking at the maximum value of the potential function.
-Recall that the board will be nonzero only in locations in the range $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
-So we know that when the board is stable at some $f'$, 
+We can get an upper board for $T(n)$ by looking at the maximum value of our spread function.
+Recall that by Lemma ~\ref{finiteextension} after any number of earthquakes the resultant board will be nonzero only on nodes in the range $[-n, n] \times [-n, n]$. 
+So we know that when the board is stable at some $b'$, 
 \begin{equation*}
-\Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n = 3*4n*4n= O(n^2)
+\Psi(b') \leq \sum_{x,y} b'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n^2 = 3*4n^2*4n^2= O(n^4)
 \end{equation*}
 We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
-\[ T(n) = O(n^2) \]
+\begin{equation*}
+T(n) = O(n^4)
+\end{equation*}
 We can get tighter bounds by looking at the number of earthquakes more carefully.
+
 \begin{definition}
- Let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
+ Let $T'(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
  \end{definition} 
@@
 \begin{equation*}
-T(x,y) = \begin{dcases} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) \neq (0,0) \\
- \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) = (0,0) .
+T'(x,y) = \begin{dcases} \lfloor\frac{T'(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) \neq (0,0) \\
+ \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) = (0,0) .
  \end{dcases}
  \end{equation*}
-So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
+So solving for $T'$ is very similar to solving a discrete version of the Laplacian.
+We note that it is not exactly the same as a discrete Laplacian as the value of the function at a coordinate is not the average of the value of the function at its neighbors.
+
+We were able to experiment different scenarios to understand the growth of $T(n)$. 
 \begin{figure}[!ht]
 \centering
-\subfigure[Global earthquakes until stable.
-Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
-\subfigure[Global earthquakes until stable.
-Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
+\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
+\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes divided by the number of chips on $(0,0)$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
 \caption{Number of global earthquakes until stability}
@@
 \end{figure}
+
 By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$.
-Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.
+Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(0,0)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(0,0) = \Theta(n\log n)$.
 
@@
 \centering
-\subfigure[Local earthquakes of center until stable.
-Horizontal axis is $n$ and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
-\subfigure[Local earthquakes of center until stable.
-Horizontal axis is $n$ and vertical axis is the number of local earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
-\caption{Number of local earthquakes in center until stability}
+\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
+\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes divided by the number of chips on (0,0).]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
+\caption{Number of local earthquakes on (0,0) until stability}
 \label{fig:growthofT'}

### 2014-10-12 finaldraft.tex
Commented out Image
added global earthquakes to definitions and notation

@@
 
-Finally, we formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.
+We also formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.
 
@@
 \end{definition}
+
+Finally, we use the above definitions to define global earthquakes, which will be defined only for finite boards. 
+\begin{definition}
+Given a finite board $b$, a global earthquake is given by applying one earthquake to every square in $b$ that initially has 4 or more chips.
+\end{definition}
+
+We will see in Section~\ref{Order of Redistribution} that the order in which earthquakes are applied does not matter, and thus global earthquakes are well-defined without the need for an ordering on the individual earthquakes.
+In order to avoid the case where there may be infinitely many squares with 4 or more chips, we only define global earthquakes for finite boards.
 
@@
 \end{definition}
-\begin{figure}[!h]
-\label{4tree}
-\centering
-\includegraphics[width=0.4\linewidth]{4tree}
-\caption{A 4-tree}
-\end{figure}
+%\begin{figure}[!h]
+%\label{4tree}
+%\centering
+%\includegraphics[width=0.4\linewidth]{4tree}
+%\caption{A 4-tree}
+%\end{figure}
 We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
@@
 \end{definition}
-\begin{figure}[!h]
-\label{joined4tree}
-\centering
-\includegraphics[width=0.8\linewidth]{joined4tree}
-\caption{A joined 4-tree}
-\end{figure}
+%\begin{figure}[!h]
+%\label{joined4tree}
+%\centering
+%\includegraphics[width=0.8\linewidth]{joined4tree}
+%\caption{A joined 4-tree}
+%\end{figure}
 

### 2014-10-12 finaldraft.tex
commented out time until stability seciton

@@
 \end{proof}
+
+Note that the results of this section imply that global earthquakes are a well-defined operation.
+Recall that a global earthquake is an operation defined on finite boards that hits every square of a finite board with 4 or more chips with an earthquake.
+By Theorem~\ref{thm:order}, the order of local earthquakes does not matter, and thus global earthquakes are unambiguously defined.
+This is important as we will refer to global earthquakes in Section~\ref{Computing with Chips}.
 %
@@
 %\end{proof}
-
-\section{Time until Stability}
-\label{Time until Stability}
-
-Another question of interest is how many global earthquakes must be applied to a finite board in order for it to stabilize.
-We will confine our analysis to the case of the board with only one stack containing $n$ chips.
+%
+%\section{Time until Stability}
+%\label{Time until Stability}
+%
+%Another question of interest is how many earthquakes must be applied to a finite board in order for it to stabilize.
+In order to simplify our simulations and analysis, we restricted our attention to global earthquakes, which hit entire boards and lead to a high degree of symmetry.
+We decided that this restriction was reasonable, as global earthquakes
+%
+%We will confine our analysis to the case of the board with only one stack containing $n$ chips.
 We are able to demonstrate upper bounds on the number of earthquakes it takes to stabilize the board and we conjecture on the existence of tighter ones.
-
-Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
-\begin{definition}
-We define $T(n)$ to be the number of global earthquakes that must be applied to the board for it to stabilize.
-\end{definition}
-
-We can get an upper board for $T(n)$ by looking at the maximum value of our spread function.
-Recall that by Lemma ~\ref{finiteextension} after any number of earthquakes the resultant board will be nonzero only on nodes in the range $[-n, n] \times [-n, n]$. 
-So we know that when the board is stable at some $b'$, 
-\begin{equation*}
-\Psi(b') \leq \sum_{x,y} b'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n^2 = 3*4n^2*4n^2= O(n^4)
-\end{equation*}
-We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
-\begin{equation*}
-T(n) = O(n^4)
-\end{equation*}
-We can get tighter bounds by looking at the number of earthquakes more carefully.
-
-\begin{definition}
- Let $T'(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
- \end{definition} 
- If $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (after taking floors).
+%
+%Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
+%\begin{definition}
+%We define $T(n)$ to be the number of global earthquakes that must be applied to the board for it to stabilize.
+%\end{definition}
+%
+%We can get an upper board for $T(n)$ by looking at the maximum value of our spread function.
+%Recall that by Lemma ~\ref{finiteextension} after any number of earthquakes the resultant board will be nonzero only on nodes in the range $[-n, n] \times [-n, n]$. 
+%So we know that when the board is stable at some $b'$, 
+%\begin{equation*}
+%\Psi(b') \leq \sum_{x,y} b'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n^2 = 3*4n^2*4n^2= O(n^4)
+%\end{equation*}
+%We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
+%\begin{equation*}
+%T(n) = O(n^4)
+%\end{equation*}
+%We can get tighter bounds by looking at the number of earthquakes more carefully.
+%
+%\begin{definition}
+% Let $T'(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
+% \end{definition} 
+% If $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (after taking floors).
 For $(0,0)$ the number of earthquakes that must hit the square equals the number of chips entering $(x,y)$ divided by 4 (after taking floors) plus the number of earthquakes needed to clear the original $n$ chips off the square.
- Since a chip only enters a square when an earthquake hits an adjacent square, we have the following formula
-\begin{equation*}
-T'(x,y) = \begin{dcases} \lfloor\frac{T'(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) \neq (0,0) \\
- \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) = (0,0) .
- \end{dcases}
- \end{equation*}
-So solving for $T'$ is very similar to solving a discrete version of the Laplacian.
+% Since a chip only enters a square when an earthquake hits an adjacent square, we have the following formula
+%\begin{equation*}
+%T'(x,y) = \begin{dcases} \lfloor\frac{T'(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) \neq (0,0) \\
+% \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) = (0,0) .
+% \end{dcases}
+% \end{equation*}
+%So solving for $T'$ is very similar to solving a discrete version of the Laplacian.
 We note that it is not exactly the same as a discrete Laplacian as the value of the function at a coordinate is not the average of the value of the function at its neighbors.
-
-We were able to experiment different scenarios to understand the growth of $T(n)$. 
-\begin{figure}[!ht]
-\centering
-\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
-\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes divided by the number of chips on $(0,0)$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
-\caption{Number of global earthquakes until stability}
-\label{fig:growthofT}
-\end{figure}
-
-By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$.
+%
+%We were able to experiment different scenarios to understand the growth of $T(n)$. 
+%\begin{figure}[!ht]
+%\centering
+%\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
+%\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes divided by the number of chips on $(0,0)$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
+%\caption{Number of global earthquakes until stability}
+%\label{fig:growthofT}
+%\end{figure}
+%
+%By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$.
 Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(0,0)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(0,0) = \Theta(n\log n)$.
-
-\begin{figure}[!ht]
-\centering
-\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
-\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes divided by the number of chips on (0,0).]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
-\caption{Number of local earthquakes on (0,0) until stability}
-\label{fig:growthofT'}
-\end{figure}
+%
+%\begin{figure}[!ht]
+%\centering
+%\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
+%\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes divided by the number of chips on (0,0).]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
+%\caption{Number of local earthquakes on (0,0) until stability}
+%\label{fig:growthofT'}
+%\end{figure}
+%
+%
 

### 2014-10-12 finaldraft.tex
got rid of subsections
changed second to last and third to last proofs of stability of finite boards

@@
 \label{finiteextension}
-Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then after applying any number of earthquakes the resultant board is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
+Suppose there are a total of $C$ chips on the board, and $b$ is nonzero in the region $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$ (equivalently, all chips on the board are in this region).
+Then after any sequence of earthquakes, the resultant board is nonzero only in the region $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
 \end{lemma}
@@
 \begin{proof}
-Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E_{s}b(i,j) \geq 1$.
-This is clear since an earthquake can bring things into column $i$ or if it moves some chips out, then there are two chips that remain in the column.
+This follows from the observation that if a row / column of the board contains at least one chip, it will always contain at least one chip.
+Note that this follows from the fact that an earthquake operation can never cause a board to break this property; if a row / column is affected by an earthquake operation, at least 2 chips stay inside that row / column.
 
-The same is true for each row. This means that if a chip  is ever in a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
+Now consider the maximum value $x$ coordinate of the board that chips can ever ``reach".
+An earthquake must first move a chip to the column given by $x$ coordinate $i_{max} +1$, and a chip will always stay there.
+The same argument follows for $i_{max} + 2$, and so on, but we can only do this up until $i_{max} +C$.
+The same argument works in any direction, so we can never leave the region $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$.
 \end{proof}
@@
 
-\begin{theorem} If $b$ is a finite board then there exists some sequence $\{s_1, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $b$ the resultant board is stable. 
+\begin{theorem} If $b$ is a finite board that is repeatedly hit by earthquakes, it must eventually reach a stable state. 
 \end{theorem}
 \begin{proof}
- Since $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.
-Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
-Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$.
-If we set
-\begin{equation*}
-D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
-\end{equation*}
-then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,.
-We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
+Lemma~\ref{finiteextension} implies that there a finite board with $C$ total chips can only ever occupy a finite number $D$ of squares.
+This implies a finite upper bound of $\binom{D+C-1}{D-1}$ possible boards (via an elementary combinatorics argument of how to distribute $C$ identical objects among $D$ distinguishable slots).
+
+In addition, Corollary~\ref{norepeats} tells us that we never visit a board state more than once.
+Thus, since we have a finite number of reachable board states, and the earthquake operation always takes a board to previously unvisited board, a sequence of earthquake operations must eventually end.
+At this end state, the board cannot allow further earthquakes, and is stable by definition.
 \end{proof}
@@
 
-\subsection{Simple Model}
-
 A simple model of computation can be defined as follows.
@@
 
-\subsection{More powerful model}
-
 As before, we assume a finite board with finitely many chips.
@@
 
-\subsection{Turner}
-
 Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
@@
 
-\subsection{Synchronize}
-
 We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
@@
 The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
-
-\subsection{Boolean circuits}
 

### 2014-10-12 finaldraft.tex
Brought finite d-trees in line
with proofs from finite

@@
 \end{theorem}
+\label{finitestability}
 \begin{proof}
@@
 We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
-\begin{equation}
+\begin{equation*}
 G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
-\end{equation}
+\end{equation*}
+\end{definition}
+
+We now formalize the board of poker chips.
+Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
+\begin{definition}
+We define a $d$-board to be a function from the set of valid coordinates to the nonnegative integers, that is 
+\begin{equation*}
+b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
+\end{equation*}
 \end{definition}
@@
 
-We now formalize the board of poker chips.
-Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
-\begin{definition}
-We define a $d$-board to be a map from the set of valid coordinates to the nonnegative integers, that is 
-\begin{equation}
-b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
-\end{equation}
-\end{definition}
-
-Recalling that $n(s)$ refers to the neighbors of $s$ we define our earthquake operation in terms of $d$-boards.
+We now define our earthquake operation in terms of $d$-boards.
 It functions analogously to the earthquake for the standard board, but instead of hitting nodes that have more than 4 chips it hits nodes that have more than $d$ chips.
 \begin{definition}
-An earthquake takes as an input an $d$-board and a node and outputs an $d$-board. If $s$ is the node input and $b(s)>d$ then 
+Let $E_s b$ be the board that results from an earthquake hitting the $d$-board $b$ at node $s$. If $b(s)>d$ then 
 \begin{align*}
@@
 \end{align*}
-Otherwise $b(s)<d$ and $E_s b=b$ as the earthquake does not move any chips.
+Otherwise $b(s)<d$ the earthquake does not move any chips and $E_s b=b$.
 \end{definition}
 
-Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a board. 
+Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a $d$-board.
+Noting that it is almost identical to definition of stability for a standard board.
 \begin{definition}
@@
 \end{definition}
-This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite board.
+This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite $d$-board.
 \begin{definition} 
@@
 
-We will now show that any finite $d$-board will stabilize under earthquakes.
+We will now show that any finite $d$-board will stabilize after a finite number of earthquakes.
 Our method of proof will mirror the method used for the square lattice, without only minor changes.
-Therefore we will proceed by shooing that if an earthquake changes the state of a $d$-board, then no sequence of earthquakes can transform the output $d$-board back into the original $d$-board.
+Therefore we will introduce a new spread function that maps $d$-boards to positive numbers.
+We once again show that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversibly changes a board configuration.
 We will then give a bound on the maximum number of configurations a given $d$-board can be transformed into by earthquakes.
@@
 $\Psi$ is a function from the finite $d$-boards and is defined as
-\begin{equation}
+\begin{equation*}
 \Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
-\end{equation}
+\end{equation*}
 Note that this is well defined for boards which contain a finite number of chips. 
@@
 
+We again demonstrate that nontrivial earthquakes increase the value of the spread function, using a nearly identical proof.
 \begin{lemma} 
@@
 \begin{proof}
-For ease of notation let $s=(s_1, s_2)$. We can evaluate $\Psi(E_s b) - \Psi(b)$.
-\begin{align}
-\Psi(E[b, s])-\Psi(b) = \sum_{(i,j) \in G_d} E_s b (i,j)|i|^2 - \sum_{(i,j) \in G_d} b(i,j)|i|^2 
-\end{align}
-We note that only $s$ and its neighbors are affected so our expression simplifies to
-\begin{align*}
-\Psi(E_s b ) - \Psi(b) &= E_s b (s) |s_1|^2 - b(s) |s_1|^2 + \sum_{(i,j) \in n(s)} (E_s b (i,j) |i|^2 - b(i,j)|i|^2) \\
-&= -d |s_1|^2 + \sum_{(i,j) \in n(s)} |i|^2
-\end{align*}
-We note that exactly $d-1$ of the neighbors of $s$ are a distance $|s_1 +1|$ from the origin and one of the neighbors of $s$ is a distance $(|s_1| - 1)$ from the origin.
+Once again the chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$.
+Thus the contribution to $\Psi$ is the same at all nodes except for those $d+1$.
+
+If the first coordinate of $s$ is $s_1$ then the contribution to the spread function from node $s$ decreases from $b(s)|s_1|^2$ to $(b(s)-d)|s_1|^2$.
+
+Out of the $d$ nodes in $n(s)$ exactly $d-1$ of the neighbors of $s$ have a first coordinate with absolute value $||s_1| +1|$ and the remaining neighbor has a first coordinate with absolute value $||s_1| - 1|$ from the origin.
 A point to emphasize is that this is the only substantial way in which this proof differs from the similar proof for the square lattice.
+
+We note that the contributions of these neighboring nodes increase by $||s_1| +1|$ and $||s_1|-1|$ respectively.
+
 Omitting the algebraic details we find that 
@@
 
-\begin{lemma}
+\begin{corollary}
+\label{notreerepeats}
 A finite $d$-board can never return to a previous state via a sequence of earthquakes
-\end{lemma}
+\end{corollary}
 \begin{proof}
@@
 \begin{proof}
- Since $b$ is finite we know $\sum_{(i,j) \in G_n} b(i,j) = C$.
-Suppose all chips lie between levels  $[i_{min}, i_{max}]$.
-Then by Lemma ~\ref{finiteextensiontree}, we have that after applying any sequence of earthquakes to $b$ the any node of the resultant board with a nonzero number of chips on it has a first coordinate contained in $ [i_{min} -2C, i_{max} +2C]$.
-
-We have restricted the size of the board to be finite, and the number of chips that can be placed on the board to be finite. 
-Thus the total number of ways that the chips can be arranged on the board is finite. 
-In particular there are fewer than $(i_{max} - i_{min}+1 + 4 C)$  and there are at most 
-\begin{equation}
-d^l,
-\end{equation}
-nodes at each level, where $l= \max(|i_{min}|, |i_{max}|)$. So there are at most 
-\begin{equation}
-N:=(i_{max} - i_{min}+1 + 4 C)d^l
-\end{equation}
-nodes that could have chips on them. Therefore there are at most 
-\begin{equation}
-\binom{N+S-1}{N-1}
-\end{equation}
-possible boards that satisfy this. We know that there are no cycles in the states, so eventually the board must stabilize.
+The proof is identical to that of Theorem ~\ref{finitestability} with Lemma ~\ref{finiteextension} replaced by Lemma ~\ref{finiteextensiontree} and Corollary ~\ref{norepeats} replaced by ~\ref{notreerepeats}.
 \end{proof}

### 2014-10-12 finaldraft.tex
Edited Computation with Chips
Up through the first theorem.

@@
 
-In this section, we show how we can compute different Boolean functions using the poker chip board and the earthquake operator.
+In this section, we show how we can compute introduce the ideas of Boolean functions and Boolean circuits and show how we can compute these use the poker chip board and the earthquake operator.
 
-A Boolean input is a number in $\{0,1\}$, and a Boolean function is a function that takes $k$ Boolean inputs and outputs a number in $\{0,1\}$.
-A simple Boolean function is the AND function that takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are set to 1.
-Another Boolean function is the OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
-If 0 is associated with ``false" and 1 is associated with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
-More complicated Boolean functions can be made with more than 2 inputs and compositions of these inputs.
-Denoting an AND with $\vee$ an OR with $\wedge$, we can construct composite functions such as $b_1 \wedge (b_2 \vee b_3)$.
+We first introduce conventional definitions of boolean variables and functions.
+\begin{definition} 
+A Boolean variable is a number in $\{0,1\}$, and a Boolean function is a function that takes in a finite number of Boolean variable and outputs a Boolean variable.
+\end{definition}
 
-Boolean circuits are formed by using wires that transmit 0's and 1's, and feeding them through various \emph{gates}.
-For example, a wire carrying a 0 and a wire carrying a 1 can be fed through an OR gate, which will ``output" a wire carrying a 1.
-In this section, we construct circuits using the poker chips board that use AND an OR gates.
+We now introduce the standard Boolean functions. 
+\begin{definition}
+ The AND function takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are 1.
+We will use $\vee$ to denote AND.
+The OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
+We will use $\wedge$ to denote OR.
+The NOT function takes in 1 Boolean input; it returns a 1 if and only if the input value is a 0.
+ \end{definition}
+ If we think of associating 0 with ``false" and 1 with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
+We can make more complicated Boolean functions with more than 2 inputs and compositions of these inputs.
 
-A simple model of computation can be defined as follows.
-We assume a finite board.
-We specify some input nodes and some output nodes.
-If an input bit is to be set to 1, then we place 4 chips in the input node.
-We apply earthquakes to the board until it stabilizes.
-If an output node ever had an earthquake, then we mark that output bit as 1; otherwise, the output bit is set to 0.
+ We now begin to define Boolean circuits, first defining the two component parts, wires and gates.
+\begin{definition}
+ A wire is a device that in theory is capable of transmitting Boolean variables over arbitrarily long distances.
+If it takes in a 1 it will output a 1 and if it takes in a 0 it will output a 0.
+A gate is a boolean function that takes in an arbitrary number of wires and outputs an arbitrary number of wires.
+\end{definition}
+Examples of gates include AND gates and OR gates. In this section we will show that we can use the poker chips board to construct circuits that use AND and OR gates.
 
-A simple example is a board that computes the identity function of one output. We can compute 
-\[ f(x) = x \]
-By having a board with no chips, and we let the input node be the same as the output node.
-Then if $x = 0$, this means that the input node will not contain any chips, so the board is stable.
-The output node never received an earthquake, so the output is 0.
-If $x = 1$, then the input node gets 4 chips and an earthquake affects it.
-So the output node is activated, and so the output is 1.
-With this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
+We now define a simple model of computation.
+We assume a finite board and specify some nodes as inputs nodes and some as outputs.
+If an input bit is to be set to 1, then we place 4 chips on the input node.
+Otherwise we do not place any chips on it.
+We then apply earthquakes to the board until it stabilizes.
+If an earthquake ever occurred on an output node, then we set the output bit to 1; otherwise, the output bit is set to 0.
+
+A simple example of this is a board that computes the identity function of one input. We can compute 
+\begin{equation*}
+f(x) = x
+\end{equation*}
+If our board has no chips on it, other than perhaps at the input node.
+We set the input node and the output node to be the same.
+If $x = 0$, then the input node does not have any chips on it and the board is stable.
+An earthquake never occurred on the output node and so the output is 0.
+If $x = 1$, then we place 4 chips on the input node and in order for the board to stabilize an earthquake must effect it.
+Because the input and output node are the same, this means an earthquake occurred on the output node and so the output is 1.
+
+Using this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
 
@@
 
-Composing these gates, we can make circuits which represent some Boolean function. In particular we can compute any Boolean function that can be represented with a planar circuit with only AND and OR.
+Composing these gates, we can make circuits which certain Boolean functions.
+In particular we can compute any Boolean function that can be represented by a planar circuit composed of only AND and OR gates.
+However using this model we cannot construct a NOT gate.
 
@@
 \begin{proof}
-Suppose for contradiction that we have a board $b$ with one input $i$ and one output $o$ which computes NOT.
-Then we know that there exists some node which initially has over $4$ chips, since otherwise, the board would be stable and so the output never gets activated.
+Suppose for contradiction that we have a board with one input $i$ and one output $o$ which computes NOT.
+Then we know that there exists some node which initially has over $4$ chips otherwise the board would be stable and the output could never be effected by an earthquake.
 
-Let $A(0)$ be the set of nodes that get activated when $i = 0$ and $A(1)$ be the set of nodes that get activated when $i=1$.
-Then we claim that $A(0) \subset A(1)$.
-This would yield a contradiction since $o \in A(0)$ but $o \notin A(1)$.
+Let $A(0)$ be the set of nodes that are targeted by earthquakes when $i = 0$ and $A(1)$ be the set of nodes that are targeted by earthquakes when $i=1$.
+We claim that $A(0) \subset A(1)$.
+This is true as when we set the input square to 1 we only increase the locations that can be targeted by nontrivial earthquakes.
+However this produces a contradiction as the output square is not in $A(1)$ by definition of our function.
+Therefore we cannot have a board that computes NOT.
 \end{proof}

### 2014-10-12 finaldraft.tex
More changes to Computation

@@
 \begin{definition} 
-Let $n(i,j)d$ be the set of neighbors of node $(i,j)$ 
-\begin{equation*}
-n(i,j) = \{ (x_1, y_1); (x_1 = i+1 \text{ or } x_1=i-1), \text{ and }(x_1, y_1) \in G_d\}
-\end{equation*}
+Let $n(i,j)d$ be the set of neighbors of node $(i,j)$.
 \end{definition}
@@
 
-\begin{lemma}
-Let $A(i)$ be the set of active nodes in an execution of board $b$ with input node $i$. Then $A(0) \subset A(1)$.
-\end{lemma}
+%\begin{lemma}
+%Let $A(i)$ be the set of active nodes in an execution of board $b$ with input node $i$. Then $A(0) \subset A(1)$.
+%\end{lemma}
 
-\begin{proof}
-By Theorem~\ref{thm:order}, we can simulate an execution of the board with input be 0 before we apply earthquakes to the input node.
+%\begin{proof}
+%By Theorem~\ref{thm:order}, we can simulate an execution of the board with input be 0 before we apply earthquakes to the input node.
 We start with a board $b$ where the input is 1, so it has 4 chips.
@@
 In addition, the input node does not contain more than $7$ chips for more than one consecutive earthquake.
-We do this until every node except the input node has less than 4 chips.
+%We do this until every node except the input node has less than 4 chips.
 
-Note that this execution is indistinguishable from the execution with input set to 0.
+%Note that this execution is indistinguishable from the execution with input set to 0.
 This means that all nodes activated in the execution with input 0 are the same as the nodes activated in the above mentioned execution.
 
-Once this is done, then we continue with the execution applying the earthquake to the input node.
+%Once this is done, then we continue with the execution applying the earthquake to the input node.
 By Theorem~\ref{thm:order}, the final configuration will be the same as any execution of the board.
 So $A(0) \subset A(1)$.
-\end{proof}
+%\end{proof}
 

### 2014-10-13 finaldraft.tex
Additional Changes to Comp
Also added better images for circuits

@@
 Otherwise we do not place any chips on it.
-We then apply earthquakes to the board until it stabilizes.
-If an earthquake ever occurred on an output node, then we set the output bit to 1; otherwise, the output bit is set to 0.
+We then apply global earthquakes to the board until it stabilizes.
+If an output node ever has chips redistributed from it by an earthquake, then we set the output bit to 1; otherwise, the output bit is set to 0.
 
@@
 \end{equation*}
-If our board has no chips on it, other than perhaps at the input node.
-We set the input node and the output node to be the same.
+Our board has no chips on it, other than perhaps at the input node.
+We set the output node to be the same as the input node.
 If $x = 0$, then the input node does not have any chips on it and the board is stable.
@@
 
-Composing these gates, we can make circuits which certain Boolean functions.
+Composing these gates, we can make circuits which can be used to compute certain Boolean functions.
 In particular we can compute any Boolean function that can be represented by a planar circuit composed of only AND and OR gates.
@@
 We claim that $A(0) \subset A(1)$.
-This is true as when we set the input square to 1 we only increase the locations that can be targeted by nontrivial earthquakes.
-However this produces a contradiction as the output square is not in $A(1)$ by definition of our function.
-Therefore we cannot have a board that computes NOT.
+This is true as when we set the input square to 1 we only increase the locations that can chips redistributed from them by earthquakes.
+However the output square is not in $A(1)$ by definition of our function, which produces a contradiction.
+Therefore our current model of computation cannot have a board that computes NOT.
 \end{proof}
@@
 
+We now introduce a slightly more complicated model for computation that will allow us to build NOT gates as well as AND and OR gates.
 As before, we assume a finite board with finitely many chips.
-The inputs are designated by having one input node.
-The outputs also each have one node.
-We now make the model of computation stronger by adding a special node $r$.
-When $r$ gets activated, it ends the execution and reads the outputs.
-We assume that in one ``round" we apply an earthquake to each valid node and that $r$ is the last node which gets applied an earthquake.
+We again specify certain nodes as  inputs and outputs, placing 4 chips on input bits that are to be set to 1.
+We also designate a specific node $r$, it can be any node including inputs and outputs.
+When $r$ is effected by an earthquake, we finish redistributing all of the chips and read the outputs, setting an output to be 1 if it has 4 chips on it and 0 otherwise.
 
-In order to make it easier to compute, we will make a special \textit{clock} node, designated as $c$ which always starts out with $4$ chips.
-We will have the clock go through each input and travel in parallel to each input.
-We will build \textit{gadgets} that allow us to compute with the clock so that we can build gates in a modular and then put them together.
+We show how to make a board that computes the identity, as well as AND and OR gates. They are essentially the same as for our simpler model of computation but they include now include an $r$ node.
 
-We will also make a wire path that takes $c$ to $r$, so that we know exactly at what point the outputs are read.
+\begin{figure}
+\centering
+\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wirewithr}\label{fig:wirer}}
+\qquad\qquad
+\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andwithr}\label{fig:ANDr}}
+\qquad\qquad
+\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orwithr}\label{fig:ORr}}
+\caption{Gates in the our more complicated model.
+Green are the inputs, red are the outputs and yellow is the node $r$.
+We disregard edges and assume that any two nodes that are touching have an edge between them.}
+\label{fig:simplegateswithr}
+\end{figure}
 
-As shown before, simple example is given by a board that computes the identity.
-For simplicity, we write the boards as a board of numbers, which every space is a node and two nodes have an edge between them if they are either above, below, left, or right.
-
-\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
-				     4_c & 3 & 3 & 3 & 3 & 3_r \\
-				     0 & 0 & 0 & 0 & 0 & 0 \\
-				     i  & 3 & 3 & 3 & 3 & 3_o \\
-				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
-Where $i$ indicates the input bit node, and $o$ indicates the output bit.
-
-One can see that $r$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $r$.
-Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active.
-Therefore, we can compute the identity.
-
-This circuit also shows how to make a ``wire" in the circuit. We will represent a wire as two parralel arrows, one containing the clock path and the other carrying the input.
-
-Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
+%Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
 this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $r$ is the output read and $o$ is the output bit.
 
-\[ \begin{array}{cccccc} 0 & 0 & r & 0 & o \\
-				     0 & 3 & 2 & 0 & 3 \\
-			            0 & 2 & 3 & 0 & 3 \\
-				    0 & 3 & 2 & 0 & 3 \\
-				    0 & 2 & 3 & 0 & 3 \\
-				    c & 3 & 3 & 0 & 3 \\
-				    0 & 0 & 0 & 0 & 3 \\
-				    i & 3 & 3 & 3 & 3 \end{array} \]	
-Since the clock snakes along the path while $i$ takes the shorter path. 
-Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
+%Since the clock snakes along the path while $i$ takes the shorter path. 
+%Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
 
-We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
-\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
-					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
-					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
-					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
-					 0 & 0 & 0 & 0 & o & 0 & r & 3 \end{array} \]
-Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
+%We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
+
+%Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
 $i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.
 
-The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
+%The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
 
-Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
+%Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
 The following can compute the AND of two inputs.
 
-\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
-					   &    & 3   & 2 &  3    &    &  \\
-					   &    &    & 3 &    & 3 & c_2 \\
-					o & 3 & 3 & 2 & 3 & 2 & 	    \\
-					   &    &    &    &    & 3 & i_2 \end{array} \]
-Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
+%Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
 So in this case, $o = i_1$ AND $i_2$.
 
-Similarly, we can compute the OR of two circuits with a very simple modification:
-\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
-					   &    & 3   & 2 &  3    &    &  \\
-					   &    &    & 3 &    & 3 & c_2 \\
-					o & 3 & 3 & 3 & 3 & 2 & 	    \\
-					   &    &    &    &    & 3 & i_2 \end{array} \]
-And we can do a similar set up to synchronize the clocks.
+%Similarly, we can compute the OR of two circuits with a very simple modification:
+%And we can do a similar set up to synchronize the clocks.
 
-We can build a NOT board with the following circuit:
+Our goal in introducing this more complicated model of computation is to allow ourselves to build a board that would compute NOT. The following board does so:
 \begin{figure}
 \centering
-\includegraphics[width=0.25\linewidth]{notgate.pdf}
-\caption{Possible board to compute $f(x) = \neg x$}
+\includegraphics[width=0.5\linewidth]{notgate}
+\caption{Board which computes $f(x) = \neg x$}
 \label{fig:NOT}
 \end{figure}
-In Figure~\ref{fig:NOT}, the wire to $c$ and $r$ is the same length as the wire from $c$ to $o$, but the wire from $i$ to $r$ is shorter than the length of the wire from $c$ to $o$.
-If $i$ is set to $1$, then $r$ will be activated before the clock reaches $o$, so $o$ will be read as 0.
-If $i$ is set to 0, then the clock will reach $o$ and $r$ at the same time, so the output is $1$.
-This computes $f(x) = \neg x$; however, we have switched the order of $r$ and $o$.
-We also do not know how to compose this board with other boards.
-
-We would also like to build a crossover gadget to allow us to compute circuits which are not necesarily.
-
+We note that in Figure~\ref{fig:NOT}, the number of nodes between the 4 and $r$ is the as the wire from the 4 and the output node, but the number of nodes separating the input node and $r$ is shorter than the number of nodes separating the 4 and the output node.
+If $i$ is set to $1$, then $r$ will be activated while the output only has 3 chips on it, so the output will be read as 0.
+If the input is set to 0, then $r$ and  the square adjacent to $o$ will both have 4 chips on them at same time so the output will be $1$.
+This computes $f(x) = \neg x$.
+however, we have switched the standard positions of $r$ and the output node, so it is unclear how to compose this board with other ones.
 \end{document}

### 2014-10-13 finaldraft.tex
Minor Changes to first 3 sections

@@
 In particular, we formalize the effects of earthquakes, as well as what it means for a board of chips to be finite or infinite.
-In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to $\mathbb{R}$. 
-We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
-Then we show that this means that no boards can return to their same state.
-From this, we conclude that such boards never cycle back to previous configurations. 
+In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to positive integers. 
+We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations.
+We also show that the number of possible configurations a finite board can be transformed into by earthquake is bounded.
+These two results together will show that all finite boards reach a steady state.
 In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
@@
 For the remainder of the paper, we consider other approaches and extensions of the problem. 
-In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes.
+%In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes.
 In particular, we look at the starting arrangement where all of the board's $n$ chips are initially on one central square.
-We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
+%We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
 In Section~\ref{Board Variants: d-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
 Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits.
-We will also give impossibility results that show the limits of our model of computation.
 \section{Definitions and Notation}
@@
 \subfigure[Board after earthquake to active square]{\includegraphics[width=0.25\linewidth]{action2.pdf}\label{fig:boardafterequarthquake}}
-\caption{Earthquake applied to board. Green signifies active nodes}
+\caption{Earthquake applied to board. Green signifies nodes with more than 4 chips}
 \label{fig:earthquakeexample}
@@
 \end{definition}
-This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be affected by earthquake operations.
+This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be changed by earthquake operations.
 
@@
 
-In this section, we show that any finite board will eventually become stable after a finite number of earthquakes.
-We first introduce a spread function that maps boards to positive numbers.
+In this section, we show that any finite board will become stable after a finite number of earthquakes.
+We first introduce a spread function that maps boards to the positive integers.
 We demonstrate that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversably changes a board configuration.
@@
 
-We then give a bound on the maximum number of board configurations a finite board can ever reach.
-We will argue that since a finite number of boards are reachable and board states are never revisited, that eventually this process must terminate.
-Thus, we eventually reach a stable board where no earthquakes are allowable.
+We then give a bound on the maximum number of board configurations a finite board can be transformed into by earthquakes.
+We will argue that since a finite number of boards are reachable and board states are never revisited, eventually this process must terminate.
+Thus, we eventually reach a stable board where no earthquakes will change the board.
 
@@
 
-Thus, the function sums up the chip stack sizes weighted by the taxicab distance to the origin squared.
+Essentially the function sums up the number of chips on a given node weighted by the taxicab distance to the origin squared.
 The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are.
@@
 \label{spreadfunction}
-Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb$. Then
+Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb \neq b$. Then
 \begin{equation*}
@@
 If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$.
-Since we only need a lower bound for the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.
+We only need a lower bound for the increase in the spread function, so we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.
 
@@
 In addition, Corollary~\ref{norepeats} tells us that we never visit a board state more than once.
-Thus, since we have a finite number of reachable board states, and the earthquake operation always takes a board to previously unvisited board, a sequence of earthquake operations must eventually end.
-At this end state, the board cannot allow further earthquakes, and is stable by definition.
+Thus, since we have a finite number of reachable board states, and a nontrivial earthquake operation always outputs a previously unvisited board, a sequence of earthquake operations must eventually end.
+At this end state, the board cannot allow further nontrivial earthquakes, and is stable by definition.
 \end{proof}
@@
 This infinite board behaves the same way as a finite board with the same values on $R$ and 0 outside of $R$.
-But this means that earthquakes transformed a finite board back into a previous state which is a contradiction.
+But this means that earthquakes transformed a finite board back into a previous state which contradicts Lemma ~\ref{finiteextension}.
 Therefore infinite boards cannot be transformed back into a previous state.

### 2014-10-13 finaldraft.tex
Minor changes to final sections

@@
 In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular node in order to stabilize the board is invariant of the order in which the earthquakes occur.
-We do so by first describing conditions under which the position of an Earthquake in a sequence can change without effecting the final state.
-We then define a process that rearranges a sequence of earthquakes without changing the final board state and produces the same sequence for any board.\\
+We show this by first describing conditions under which we can rearrange a sequence earthquakes without changing the resultant board.
+We then define a process that rearranges a sequence of earthquakes without changing the final board state and for a given board will always produce the same sequence.\\
 
@@
 
-Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, LE_j$ applied to $b_{m-1}$.
+Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ applied to $b_{m-1}$.
 
@@
 are both valid and produce the same final board configuration.
-If we apply the same reduction as above  we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board.
+If we apply the same reduction we used in our first step of this proof we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board.
 \end{proof}
@@
 Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order.
-The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
-This technique will then specify an order in which the earthquakes will occur, which completes the proof.
+The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, based on the initial board and then move these earthquakes to the front of the sequence without changing the final board.
+This technique will specify an order in which the earthquakes will occur, depending only on the original board which completes the proof.
 
@@
 Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $b$.
-Furthermore the number of times that a square $x$ is targeted is invariant of the order in which the earthquakes occur.
+Furthermore the number of times that a node $s$ is targeted is invariant of the order in which the earthquakes occur.
 \end{theorem}
 \begin{proof}
-Given the original configuration of the board, let $A_i$ be the number of chips on square $i$. We know that square $i$ will be hit by at least 
+Given the original configuration of the board, let $A_i$ be the number of chips on node $i$. We know that node $i$ will be hit by at least 
 \begin{equation*}
@@
 Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquakes that target  square $i$.
-We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a square that starts with $A_i$ chips on it.
+We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a node that starts with $A_i$ chips on it.
 Thus using Lemma ~\ref{shiftlemma}, we can move these earthquakes to the front of the sequence without changing the final positioning of the board.
-We do this for all squares $i$.
+We do this for all nodes $i$.
 
-The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the squares.
+The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the nodes.
 
-Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the squares.
+Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the nodes.
 We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes.
-We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
+We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma ~\ref{shiftlemma}).
 
-Thus we have transformed our starting sequence into canonical sequence that is completely determined by the state of the initial board.
-Therefore we can do this for any other sequence that stabilizes the board and get the same sequence.
+Thus we have transformed our starting sequence into a sequence that is completely determined by the state of the initial board, noting that the final state of the board is still the same.
+We can do this for any other sequence that stabilizes the board and get the same sequence.
 This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them).
@@
 
-Note that the results of this section imply that global earthquakes are a well-defined operation.
+Note that the results of this section imply that global earthquakes are a well-defined operation on finite boards.
 Recall that a global earthquake is an operation defined on finite boards that hits every square of a finite board with 4 or more chips with an earthquake.
-By Theorem~\ref{thm:order}, the order of local earthquakes does not matter, and thus global earthquakes are unambiguously defined.
+By Theorem~\ref{thm:order}, the order of standard earthquakes does not matter, and thus global earthquakes are unambiguously defined.
 This is important as we will refer to global earthquakes in Section~\ref{Computing with Chips}.
@@
 \begin{definition}
-A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each have $d-1$ unique children and so on.  
+A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each are connected by edges to $d-1$ unique children and so on.  
 \end{definition}

### 2014-10-13 finaldraft.tex
split up some definitions

@@
 \begin{definition}
- The AND function takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are 1.
-We will use $\vee$ to denote AND.
-The OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
-We will use $\wedge$ to denote OR.
-The NOT function takes in 1 Boolean input; it returns a 1 if and only if the input value is a 0.
+ The AND function takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are 1. We will use $\vee$ to denote AND.
+\end{definition}
+\begin{definition}
+The OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0. We will use $\wedge$ to denote OR.
+\end{definition}
+\begin{definition}
+The NOT function takes in 1 Boolean input; it returns a 1 if and only if the input value is a 0. We will use $\neg$ to denote NOT.
  \end{definition}

### 2014-10-13 finaldraft.tex
edited some minor things
added guiding text to d trees section

@@
 Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$.
-Lemma~\ref{spreadfunction} shows that $\psi(b) < \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
+Lemma~\ref{spreadfunction} shows that $\Psi(b) < \Psi(E_1E_2\cdots E_kb)$, which is a contradiction.
 \end{proof}
@@
 Now that we have shown that all finite boards will eventually reach a stable state we would like to know if the order in which these earthquakes occur matters.
-More specifically, if we consider some finite starting board and two sequences of earthquakes that stabilizes the board, we would like to know if the resultant board is the same for both sequences.
+More specifically, if we consider some finite starting board and two sequences of earthquakes that stabilize the board, we would like to know if the resultant board is the same for both sequences.
 
@@
 \label{earthquakeredefine}
-Consider a board $b$ if $E$ is a valid earthquake, then 
+Consider a board $b$. If $E$ is a valid earthquake, then 
 \begin{equation*}
@@
 Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
+Note that we will frequently refer to sequences of earthquakes.
+Earthquake sequences correspond to those earthquakes being applied in the order the sequence specifies.
+For example, if we have a sequence of earthquakes $E_1,E_2,E_3$ applied to board $b$, the resulting board is $E_3E_2E_1b$.
+
 \begin{lemma}
@@
 
-We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
+We prove this by contradiction. Assume $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$, we know $b(s_1) \geq 4 $. 
 Therefore we must have $b(s_1)<8$ and $s_2=s_1$.
@@
 Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$.
-Applying Lemma ~\ref{earthquakeredefine} we have
+Applying Lemma~\ref{earthquakeredefine} we have
 \begin{align*}
@@
 \end{align*}
-The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
+The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering, the algebra above is valid.
 \end{proof}
 
-We can use this lemma along with a simple induction to show that under appropriate conditions we can move an earthquake earlier in the sequence without affecting the final board state.
+We can use this lemma along with a simple induction to show that under appropriate conditions we can move an earthquake to an earlier position in the sequence without affecting the final board state.
 \begin{lemma}
@@
 We can show this by inducting on the length of the sequence $l=j-m+1$.
-The base case of $l=2$ is shown in lemma ~\ref{swappinglemma}.
+The base case of $l=2$ is shown in Lemma~\ref{swappinglemma}.
 Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$.
@@
 \end{align*}
-are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma ~\ref{swappinglemma} we know that the two sequences 
+are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma~\ref{swappinglemma} we know that the two sequences 
 \begin{align*}
@@
 
-Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order.
-The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, based on the initial board and then move these earthquakes to the front of the sequence without changing the final board.
-This technique will specify an order in which the earthquakes will occur, depending only on the original board which completes the proof.
+Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board, we can show that the number of earthquakes hitting a square for a given board is the same regardless of order.
+
+The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen based on the initial board.
+We can then move these earthquakes to the front of the sequence without changing the final board.
+This technique will specify an order in which the earthquakes will occur, depending only on the original board, which completes the proof.
 
@@
 We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes.
-We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma ~\ref{shiftlemma}).
+We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma~\ref{shiftlemma}).
 
@@
 
+So far, we have analyzed the poker chips problem assuming that the underlying graph is an infinite grid graph.
+One important feature of this graph is that it is \emph{4-regular}, which means that each node has exactly 4 neighbors.
+Graphs where every node has an equal number of neighbors allow for an easy definition of the earthquake operation; if every node has $d$ neighbors, then an earthquake redistributes $d$ chips to each of the $d$ neighbors.
+
+Thus, in this section we analyze how the problem changes when played over a large class of graphs that are regular, but are not grid-shaped.
+
 \begin{definition}
@@
 \end{definition}
-%\begin{figure}[!h]
-%\label{4tree}
-%\centering
-%\includegraphics[width=0.4\linewidth]{4tree}
-%\caption{A 4-tree}
-%\end{figure}
+\begin{figure}[!h]
+\label{4tree}
+\centering
+\includegraphics[width=0.4\linewidth]{4tree}
+\caption{A 4-tree}
+\end{figure}
 We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
@@
 \end{definition}
-%\begin{figure}[!h]
-%\label{joined4tree}
-%\centering
-%\includegraphics[width=0.8\linewidth]{joined4tree}
-%\caption{A joined 4-tree}
-%\end{figure}
+\begin{figure}[!h]
+\label{joined4tree}
+\centering
+\includegraphics[width=0.8\linewidth]{joined4tree}
+\caption{A joined 4-tree}
+\end{figure}
 
@@
 We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
+\end{definition}
 \begin{equation*}
@@
 \end{equation*}
-\end{definition}
 

### 2014-10-13 finaldraft.tex
merging

--- 2014-10-13 finaldraft.tex
@@
 In particular, we formalize the effects of earthquakes, as well as what it means for a board of chips to be finite or infinite.
-In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to positive integers. 
-We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations.
-We also show that the number of possible configurations a finite board can be transformed into by earthquake is bounded.
-These two results together will show that all finite boards reach a steady state.
+In Section~\ref{Stability of Finite Boards}, we introduce a spread function that maps finite board configurations to $\mathbb{R}$. 
+We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations, from which we will conclude that all finite boards reach a steady state.
+Then we show that this means that no boards can return to their same state.
+From this, we conclude that such boards never cycle back to previous configurations. 
 In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
@@
 For the remainder of the paper, we consider other approaches and extensions of the problem. 
-%In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes.
+In  Section~\ref{Time until Stability}, we present results from our computer simulations of repeated earthquakes.
 In particular, we look at the starting arrangement where all of the board's $n$ chips are initially on one central square.
-%We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
+We conjecture based on experimental evidence that such arrangements stabilize after $\Theta(n \log n)$ global earthquakes. 
 In Section~\ref{Board Variants: d-Trees}, we generalize away from the infinite grid graph and consider what happens when a similar process happens on tree graphs of infinite size. 
 Finally, in Section~\ref{Computing with Chips} we present a model of computation based on the poker chip board and show how to construct simple Boolean circuits.
+We will also give impossibility results that show the limits of our model of computation.
 \section{Definitions and Notation}
@@
 \end{figure}
+
 We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$.
@@
 \subfigure[Board after earthquake to active square]{\includegraphics[width=0.25\linewidth]{action2.pdf}\label{fig:boardafterequarthquake}}
-\caption{Earthquake applied to board. Green signifies nodes with more than 4 chips}
+\caption{Earthquake applied to board. Green signifies active nodes}
 \label{fig:earthquakeexample}
@@
 \end{definition}
-This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be changed by earthquake operations.
+This definition corresponds with the intuitive notion of a process that has stabilized, as stable boards can no longer be affected by earthquake operations.
 
-We also formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.
+Finally, we formalize the notion of $\emph{finite}$ and $\emph{infinite}$ boards, a distinction we will refer to extensively throughout this paper.
 
@@
 
-Finally, we use the above definitions to define global earthquakes, which will be defined only for finite boards. 
-\begin{definition}
-Given a finite board $b$, a global earthquake is given by applying one earthquake to every square in $b$ that initially has 4 or more chips.
-\end{definition}
-
-We will see in Section~\ref{Order of Redistribution} that the order in which earthquakes are applied does not matter, and thus global earthquakes are well-defined without the need for an ordering on the individual earthquakes.
-In order to avoid the case where there may be infinitely many squares with 4 or more chips, we only define global earthquakes for finite boards.
-
 \section{Stability of Finite Boards}
@@
 
-In this section, we show that any finite board will become stable after a finite number of earthquakes.
-We first introduce a spread function that maps boards to the positive integers.
+In this section, we show that any finite board will eventually become stable after a finite number of earthquakes.
+We first introduce a spread function that maps boards to positive numbers.
 We demonstrate that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversably changes a board configuration.
@@
 
-We then give a bound on the maximum number of board configurations a finite board can be transformed into by earthquakes.
-We will argue that since a finite number of boards are reachable and board states are never revisited, eventually this process must terminate.
-Thus, we eventually reach a stable board where no earthquakes will change the board.
+We then give a bound on the maximum number of board configurations a finite board can ever reach.
+We will argue that since a finite number of boards are reachable and board states are never revisited, that eventually this process must terminate.
+Thus, we eventually reach a stable board where no earthquakes are allowable.
 
@@
 Then the spread function is
-\begin{equation*}
-\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 
-\end{equation*}
+\[ \psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 \]
 
-Essentially the function sums up the number of chips on a given node weighted by the taxicab distance to the origin squared.
-The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are.
-Intuitively, earthquakes takes chips that are condensed in a single pile and spreads them out over a greater area.
-We capture this in the following lemma:
+The motivation for this definition is that the spread function should give a measure of how ``spread out" the chips on a board are.
+Intuitvely, earthquakes have the function of moving chips that are condensed in a single piile and spreading them out over a greater area.
+This function, which sums up the chip stack sizes weighted by the taxicab distance squared to the origin, sufficiently captures the idea of spread.
+
+\begin{theorem}
+\label{finitestability}
+If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
+\end{theorem}
+
+In order to show finite boards never revisit a prior arrangement, we define a spread function that maps boards to nonnegative integers.
+The spread function should give some measure of how ``spread out" the chips on a board are.
+Earthquakes seem to ``spread out" the chips on a board and so we would like earthquakes to increase the value of the function.
 
 \begin{lemma}
-\label{spreadfunction}
-Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb \neq b$. Then
+Suppose $b$ is a finite board. If $E_{s}b \neq b$, then no application of earthquakes to $E_{s}b$ can produce $b$. 
+\end{lemma}
+
+\begin{proof}
+For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as a map from finite boards to the nonnegative integers
 \begin{equation*}
-\Psi(E_sb) > \Psi(b) 
+S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
 \end{equation*}
-so the earthquake strictly increases the spread of the board.
-\end{lemma}
-\begin{proof}
-The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.
+We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared.
+This is well-defined since the total number of chips on the board is finite.
+We claim that
+\begin{equation*}
+E_{s} b \neq b \Rightarrow S(E[b, s]) > S(b)
+\end{equation*}
+That is if an earthquake changes the board, $S$ increases.
+This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.
 
-Let the taxicab distance to $s$ be $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.
-
-Out of the 4 nodes in $n(s)$, some have a taxicab distance of $d_s+1$, and the rest are at distance $d_s-1$.
-In particular, if $s$ is not exactly on the $x$ or $y$ axis, exactly 2 neighboring nodes are at $d_s+1$, and exactly 2 are at distance $d_s-1$.
-If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$.
-We only need a lower bound for the increase in the spread function, so we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.
-
-Suppose that the two neighboring nodes at distance $d_s+1$ have chip stacks of $n_1$ and $n_2$ originally.
-Then their contribution to the spread function goes from $n_1(d_s+1)^2$ and $n_2(d_s+1)^2$ to $(n_1+1)(d_s+1)^2$ and $(n_2+1)(d_s+1)^2$.
-If the other two nodes at distance $d_s-1$ have chip stacks of $n_3$ and $n_4$, their contribution to the spread function goes from $n_3(d_s-1)^2$ and $n_4(d_s-1)^2$ to $(n_3+1)(d_s-1)^2$ and $(n_4+1)(d_s-1)^2$.
-
-Omitting the algebraic details, we find that summing up the changes in the spread function gives
+We evaluate $S(E_{s} b) - S(b)$, recalling that $E_s b(i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
+\begin{align*}
+S(E_{s} b) - S(b) &=\sum_{i,j} E_{s} b (i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
+&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
+\end{align*}
+For at most two of the neighbors of $(s_1, s_2)$ 
 \begin{equation*}
- \Psi(E_sb) - \Psi(b) = 4
+(|i|+|j|)= (|s_1|+|s_2|-1)^2,
 \end{equation*}
-which completes the proof.
-\end{proof}
-
-\begin{corollary}
-\label{norepeats}
-A finite board can never return to a previous state via a sequence of earthquakes.
-\end{corollary}
-\begin{proof}
-Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$.
-Lemma~\ref{spreadfunction} shows that $\psi(b) < \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
+and for at least two of the neighbors 
+\begin{equation*}
+(|i|+|j|)=(|s_1|+|s_2|+1)^2,
+\end{equation*}
+ and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
+\begin{align*}
+&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
+& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
+\end{align*}
 \end{proof}
@@
 \label{finiteextension}
-Suppose there are a total of $C$ chips on the board, and $b$ is nonzero in the region $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$ (equivalently, all chips on the board are in this region).
-Then after any sequence of earthquakes, the resultant board is nonzero only in the region $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
+Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
+Then after applying any number of earthquakes the resultant board is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
 \end{lemma}
@@
 \begin{proof}
-This follows from the observation that if a row / column of the board contains at least one chip, it will always contain at least one chip.
-Note that this follows from the fact that an earthquake operation can never cause a board to break this property; if a row / column is affected by an earthquake operation, at least 2 chips stay inside that row / column.
+Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E_{s}b(i,j) \geq 1$.
+This is clear since an earthquake can bring things into column $i$ or if it moves some chips out, then there are two chips that remain in the column.
 
-Now consider the maximum value $x$ coordinate of the board that chips can ever ``reach".
-An earthquake must first move a chip to the column given by $x$ coordinate $i_{max} +1$, and a chip will always stay there.
-The same argument follows for $i_{max} + 2$, and so on, but we can only do this up until $i_{max} +C$.
-The same argument works in any direction, so we can never leave the region $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$.
+The same is true for each row. This means that if a chip  is ever in a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
 \end{proof}
 
-We are now ready to show that all finite boards will reach a stable state.
+We are now ready to prove the theorem that all finite boards will reach a stable state.
 
-\begin{theorem} If $b$ is a finite board that is repeatedly hit by earthquakes, it must eventually reach a stable state. 
-\end{theorem}
-\label{finitestability}
 \begin{proof}
-Lemma~\ref{finiteextension} implies that there a finite board with $C$ total chips can only ever occupy a finite number $D$ of squares.
-This implies a finite upper bound of $\binom{D+C-1}{D-1}$ possible boards (via an elementary combinatorics argument of how to distribute $C$ identical objects among $D$ distinguishable slots).
-
-In addition, Corollary~\ref{norepeats} tells us that we never visit a board state more than once.
-Thus, since we have a finite number of reachable board states, and a nontrivial earthquake operation always outputs a previously unvisited board, a sequence of earthquake operations must eventually end.
-At this end state, the board cannot allow further nontrivial earthquakes, and is stable by definition.
+(of Theorem~\ref{finitestability}) Recall that $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.
+Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$.
+Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$.
+If we set
+\begin{equation*}
+D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
+\end{equation*}
+then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,.
+We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
 \end{proof}
@@
 This infinite board behaves the same way as a finite board with the same values on $R$ and 0 outside of $R$.
-But this means that earthquakes transformed a finite board back into a previous state which contradicts Lemma ~\ref{finiteextension}.
+But this means that earthquakes transformed a finite board back into a previous state which is a contradiction.
 Therefore infinite boards cannot be transformed back into a previous state.
@@
 In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular node in order to stabilize the board is invariant of the order in which the earthquakes occur.
-We show this by first describing conditions under which we can rearrange a sequence earthquakes without changing the resultant board.
-We then define a process that rearranges a sequence of earthquakes without changing the final board state and for a given board will always produce the same sequence.\\
+We do so by first describing conditions under which the position of an Earthquake in a sequence can change without effecting the final state.
+We then define a process that rearranges a sequence of earthquakes without changing the final board state and produces the same sequence for any board.\\
 
@@
 We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
-We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid when applied one after another. 
+We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid in the order applied. That is 
+\begin{equation*}
+E^{m}[b, \{s_k\}](s_{m+1}  \geq 4
+\end{equation*}
+for all $1\geq m+1 \leq n$.
 \end{definition}
@@
 \begin{equation*}
-E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
+E[b, s](x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
 \end{equation*}
@@
 \begin{equation*}
-E_2 E_1 b = E_1 E_2 b
+E_2[E_1[b, s_1] s_2] = E_1[LE_2[b, s_2], s_1] 
 \end{equation*}
@@
 
-We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
+We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2[b, s_2](s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
 Therefore we must have $b(s_1)<8$ and $s_2=s_1$.
-But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes.
+But then $E_1[b, s_1](s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes.
 This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.
 
-Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$.
+Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2[ E_1[b, s_1], s_2](x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$.
 Applying Lemma ~\ref{earthquakeredefine} we have
 \begin{align*}
-E_2 E_1 b (x) = E_1(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
+E_2[E_1[b, s_1], s_2](x) = E_1[b, s_1])(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
 = b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
 = b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
-= E_2 b(x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
-= E_1 E_2 b(x) 
+= E_2[b, s_2](x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
+= E_1[E_2 [b, s_2], s_1](x) 
 \end{align*}
@@
 
-Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ applied to $b_{m-1}$.
+Therefore we must show that the sequence $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ applied to $b_{m-1}$ is valid and produces the same board as the sequence $E_m, E_{m+1}, \ldots, E_{j-1}, LE_j$ applied to $b_{m-1}$.
 
@@
 are both valid and produce the same final board configuration.
-If we apply the same reduction we used in our first step of this proof we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board.
+If we apply the same reduction as above  we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board.
 \end{proof}
@@
 Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order.
-The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, based on the initial board and then move these earthquakes to the front of the sequence without changing the final board.
-This technique will specify an order in which the earthquakes will occur, depending only on the original board which completes the proof.
+The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, and that we can move it to the front of the sequence of earthquakes without changing the final board.
+This technique will then specify an order in which the earthquakes will occur, which completes the proof.
 
@@
 Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $b$.
-Furthermore the number of times that a node $s$ is targeted is invariant of the order in which the earthquakes occur.
+Furthermore the number of times that a square $x$ is targeted is invariant of the order in which the earthquakes occur.
 \end{theorem}
 \begin{proof}
-Given the original configuration of the board, let $A_i$ be the number of chips on node $i$. We know that node $i$ will be hit by at least 
+Given the original configuration of the board, let $A_i$ be the number of chips on square $i$. We know that square $i$ will be hit by at least 
 \begin{equation*}
@@
 Thus, somewhere in the sequence $\{E_i\}$, there exist $\lfloor \frac{A_i}{4} \rfloor$ earthquakes that target  square $i$.
-We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a node that starts with $A_i$ chips on it.
+We note that these earthquakes will be valid in any position to the left of their original one because they are among the first $\lfloor \frac{A_i}{4} \rfloor$ to target a square that starts with $A_i$ chips on it.
 Thus using Lemma ~\ref{shiftlemma}, we can move these earthquakes to the front of the sequence without changing the final positioning of the board.
-We do this for all nodes $i$.
+We do this for all squares $i$.
 
-The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the nodes.
+The earthquakes that have been moved to the front of the sequence can be in any order because all these earthquakes could have happened with the initial values of the squares.
 
-Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the nodes.
+Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the squares.
 We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes.
-We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma ~\ref{shiftlemma}).
+We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of our lemma).
 
-Thus we have transformed our starting sequence into a sequence that is completely determined by the state of the initial board, noting that the final state of the board is still the same.
-We can do this for any other sequence that stabilizes the board and get the same sequence.
+Thus we have transformed our starting sequence into canonical sequence that is completely determined by the state of the initial board.
+Therefore we can do this for any other sequence that stabilizes the board and get the same sequence.
 This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them).
 \end{proof}
-
-Note that the results of this section imply that global earthquakes are a well-defined operation on finite boards.
-Recall that a global earthquake is an operation defined on finite boards that hits every square of a finite board with 4 or more chips with an earthquake.
-By Theorem~\ref{thm:order}, the order of standard earthquakes does not matter, and thus global earthquakes are unambiguously defined.
-This is important as we will refer to global earthquakes in Section~\ref{Computing with Chips}.
 %
@@
 %\end{proof}
-%
-%\section{Time until Stability}
-%\label{Time until Stability}
-%
-%Another question of interest is how many earthquakes must be applied to a finite board in order for it to stabilize.
-In order to simplify our simulations and analysis, we restricted our attention to global earthquakes, which hit entire boards and lead to a high degree of symmetry.
-We decided that this restriction was reasonable, as global earthquakes
-%
-%We will confine our analysis to the case of the board with only one stack containing $n$ chips.
-We are able to demonstrate upper bounds on the number of earthquakes it takes to stabilize the board and we conjecture on the existence of tighter ones.
-%
-%Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
-%\begin{definition}
-%We define $T(n)$ to be the number of global earthquakes that must be applied to the board for it to stabilize.
-%\end{definition}
-%
-%We can get an upper board for $T(n)$ by looking at the maximum value of our spread function.
-%Recall that by Lemma ~\ref{finiteextension} after any number of earthquakes the resultant board will be nonzero only on nodes in the range $[-n, n] \times [-n, n]$. 
-%So we know that when the board is stable at some $b'$, 
-%\begin{equation*}
-%\Psi(b') \leq \sum_{x,y} b'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n^2 = 3*4n^2*4n^2= O(n^4)
-%\end{equation*}
-%We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
-%\begin{equation*}
-%T(n) = O(n^4)
-%\end{equation*}
-%We can get tighter bounds by looking at the number of earthquakes more carefully.
-%
-%\begin{definition}
-% Let $T'(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
-% \end{definition} 
-% If $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (after taking floors).
+
+\section{Time until Stability}
+\label{Time until Stability}
+
+Another question of interest is how many global earthquakes it would take for a finite board to reach a stable state.
+We will first focus on the case of the board with only one stack containing $n$ chips.
+
+Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
+\begin{definition}
+We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes.
+We also define $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize.
+\end{definition}
+Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$, since that many earthquakes will reduce the number of chips on $(0,0)$ to less than $4$.
+
+We can get another upper bound for $T(n)$ by looking at the maximum value of the potential function.
+Recall that the board will be nonzero only in locations in the range $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$.
+So we know that when the board is stable at some $f'$, 
+\begin{equation*}
+\Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n = 3*4n*4n= O(n^2)
+\end{equation*}
+We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
+\[ T(n) = O(n^2) \]
+We can get tighter bounds by looking at the number of earthquakes more carefully.
+\begin{definition}
+ Let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
+ \end{definition}
+ If $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (after taking floors).
 For $(0,0)$ the number of earthquakes that must hit the square equals the number of chips entering $(x,y)$ divided by 4 (after taking floors) plus the number of earthquakes needed to clear the original $n$ chips off the square.
-% Since a chip only enters a square when an earthquake hits an adjacent square, we have the following formula
-%\begin{equation*}
-%T'(x,y) = \begin{dcases} \lfloor\frac{T'(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) \neq (0,0) \\
-% \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  \quad (x,y) = (0,0) .
-% \end{dcases}
-% \end{equation*}
-%So solving for $T'$ is very similar to solving a discrete version of the Laplacian.
-We note that it is not exactly the same as a discrete Laplacian as the value of the function at a coordinate is not the average of the value of the function at its neighbors.
-%
-%We were able to experiment different scenarios to understand the growth of $T(n)$. 
-%\begin{figure}[!ht]
-%\centering
-%\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
-%\subfigure[Horizontal axis is the number of chips on $(0,0)$ and vertical axis is the number of global earthquakes divided by the number of chips on $(0,0)$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
-%\caption{Number of global earthquakes until stability}
-%\label{fig:growthofT}
-%\end{figure}
-%
-%By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$.
-Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(0,0)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(0,0) = \Theta(n\log n)$.
-%
-%\begin{figure}[!ht]
-%\centering
-%\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
-%\subfigure[Horizontal axis is the number of chips on (0,0) and vertical axis is the number of local earthquakes divided by the number of chips on (0,0).]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
-%\caption{Number of local earthquakes on (0,0) until stability}
-%\label{fig:growthofT'}
-%\end{figure}
-%
-%
+ Since a chip only enters a square when an earthquake hits an adjacent square, we have the following formula
+\begin{equation*}
+T(x,y) = \begin{dcases} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) \neq (0,0) \\
+ \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) = (0,0) .
+ \end{dcases}
+ \end{equation*}
+So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
+\begin{figure}[!ht]
+\centering
+\subfigure[Global earthquakes until stable.
+Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
+\subfigure[Global earthquakes until stable.
+Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
+\caption{Number of global earthquakes until stability}
+\label{fig:growthofT}
+\end{figure}
+By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$.
+Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.
+
+\begin{figure}[!ht]
+\centering
+\subfigure[Local earthquakes of center until stable.
+Horizontal axis is $n$ and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
+\subfigure[Local earthquakes of center until stable.
+Horizontal axis is $n$ and vertical axis is the number of local earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
+\caption{Number of local earthquakes in center until stability}
+\label{fig:growthofT'}
+\end{figure}
 
@@
 \begin{definition}
-A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each are connected by edges to $d-1$ unique children and so on.  
+A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each have $d-1$ unique children and so on.  
 \end{definition}
-%\begin{figure}[!h]
-%\label{4tree}
-%\centering
-%\includegraphics[width=0.4\linewidth]{4tree}
-%\caption{A 4-tree}
-%\end{figure}
 We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
@@
 \end{definition}
-%\begin{figure}[!h]
-%\label{joined4tree}
-%\centering
-%\includegraphics[width=0.8\linewidth]{joined4tree}
-%\caption{A joined 4-tree}
-%\end{figure}
 
@@
 We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
+\begin{equation}
+G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
+\end{equation}
+\end{definition}
+
+In order to keep our notation compact we once again define a function that outputs the neighbors of a node
+\begin{definition} 
+We define the function $n: G_d \rightarrow G_d^d$ as 
 \begin{equation*}
-G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
+n(i,j) = \{ (x_1, y_1); (x_1 = i+1 \text{ or } x_1=i-1), \text{ and }(x_1, y_1) \in G_d\}
 \end{equation*}
@@
 We now formalize the board of poker chips.
-Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
+Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate in the integer lattice.
 \begin{definition}
-We define a $d$-board to be a function from the set of valid coordinates to the nonnegative integers, that is 
-\begin{equation*}
+We define a $d$-board to be a map from the set of valid coordinates to the nonnegative integers, that is 
+\begin{equation}
 b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
-\end{equation*}
+\end{equation}
+We write $\mathcal{B}(d)$ to refer to the set of all $d$-boards.  
 \end{definition}
 
-In order to keep our notation compact we once again define a function that outputs the neighbors of a node
+Recalling that $n(s)$ refers to the neighbors of $s$ we define our earthquake operation in terms of $d$-boards.
 \begin{definition}
-Let $n(i,j)d$ be the set of neighbors of node $(i,j)$.
+An earthquake takes as an input an $d$-board and a node and outputs an $d$-board so we write $E: \mathcal{B}(d) \times G_d \rightarrow \mathcal{B}(d)$. If $s$ is the node input and $b(s)>d$ then 
+\begin{align*}
+E[b, s](s) = b(s)- d \\
+E[b, s](t) = b(t) +1 
+\end{align*}
+for $t \in n(s)$.
+Otherwise $b(s)<d$ and $E[b,s]=b$ as the earthquake does not move any chips.
+We use the notation $E^r[b, \{s_k\}]$ to denote the application of $E$ to the $b$ $r$ times, to the sites $s_1, \ldots s_r$ respectively.
 \end{definition}
 
-We now define our earthquake operation in terms of $d$-boards.
-It functions analogously to the earthquake for the standard board, but instead of hitting nodes that have more than 4 chips it hits nodes that have more than $d$ chips.
+Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a board. 
 \begin{definition}
-Let $E_s b$ be the board that results from an earthquake hitting the $d$-board $b$ at node $s$. If $b(s)>d$ then 
-\begin{align*}
-&E_s b(s) = b(s)- d \\
-&E_s b(t) = b(t) +1 \quad \forall t \in n(s)
-\end{align*}
-Otherwise $b(s)<d$ the earthquake does not move any chips and $E_s b=b$.
+We say that a $d$-board is stable if its values are strictly less than $n$.
+\end{definition}
+This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite board.
+\begin{definition} 
+We say that a $d$-board is finite if the total number of chips on it is finite. That is a $d$board $b$ is finite if the sum
+\begin{equation}
+C= \sum_{(i,j) \in G_d} b(i,j) 
+\end{equation}
+is finite. 
+We denote the set of finite $d$-boards as $B(d)_F$.
 \end{definition}
 
-Now that we have formalized the earthquake operator we can introduce a precise definition of stability of a $d$-board.
-Noting that it is almost identical to definition of stability for a standard board.
-\begin{definition}
-We say that a $d$-board is stable if its values are strictly less than $d$.
-\end{definition}
-This is equivalent to saying all earthquakes have no effect on the board. We also make a precise definition for a finite $d$-board.
-\begin{definition} 
-We say that a $d$-board is finite if the total number of chips on it is finite.
-\end{definition}
-
-We will now show that any finite $d$-board will stabilize after a finite number of earthquakes.
+We will now show that any finite $d$-board will stabilize under earthquakes.
 Our method of proof will mirror the method used for the square lattice, without only minor changes.
-Therefore we will introduce a new spread function that maps $d$-boards to positive numbers.
-We once again show that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversibly changes a board configuration.
+Therefore we will proceed by shooing that if an earthquake changes the state of a $d$-board, then no sequence of earthquakes can transform the output $d$-board back into the original $d$-board.
 We will then give a bound on the maximum number of configurations a given $d$-board can be transformed into by earthquakes.
@@
 \begin{definition}
-$\Psi$ is a function from the finite $d$-boards and is defined as
-\begin{equation*}
-\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
-\end{equation*}
-Note that this is well defined for boards which contain a finite number of chips. 
+ Let $\Psi: \mathcal{B}(d)_F \rightarrow \mathbb{Z}_{\geq 0}$ be the function
+\begin{equation}
+\Psi(f) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
+\end{equation}
+Note that this is well defined for trees which contain a finite number of chips. 
 \end{definition} 
@@
 
-We again demonstrate that nontrivial earthquakes increase the value of the spread function, using a nearly identical proof.
 \begin{lemma}
-Suppose an earthquake hits node $s$ of a finite $d$-board $b$ producing the board $E_s b$. Then
-\begin{equation*}
-\Psi( E_s b) > \Psi (b)
-\end{equation*}
+If $E[b, s_1] \neq b$, then $E^r[b, \{s_k\}] \neq b$ for all $r>0$ and all sequences $\{s_k\}$ of length $r$ that begin with $s_1$. 
 \end{lemma}
+
 \begin{proof}
-Once again the chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$.
-Thus the contribution to $\Psi$ is the same at all nodes except for those $d+1$.
+For ease of notation let $s=(s_1, s_2)$. We will show that if $E$ acts on a board nontrivially then $\Psi(E[b, s])$ will be strictly larger. That is  
+\begin{equation}
+E[b, s] \neq f \Rightarrow \Psi(E[b, s]) > \Psi(b)
+\end{equation}
 
-If the first coordinate of $s$ is $s_1$ then the contribution to the spread function from node $s$ decreases from $b(s)|s_1|^2$ to $(b(s)-d)|s_1|^2$.
-
-Out of the $d$ nodes in $n(s)$ exactly $d-1$ of the neighbors of $s$ have a first coordinate with absolute value $||s_1| +1|$ and the remaining neighbor has a first coordinate with absolute value $||s_1| - 1|$ from the origin.
+We can evaluate $\Psi(E[b, s]) - \Psi(b)$.
+\begin{align}
+\Psi(E[b, s])-\Psi(b) = \sum_{(i,j) \in G_d} E[b, s](i,j)|i|^2 - \sum_{(i,j) \in G_d} b(i,j)|i|^2 
+\end{align}
+We note that only $s$ and its neighbors are affected so our expression simplifies to
+\begin{align*}
+\Psi(E[b, s]) - \Psi(b) &= E[b, s](s) |s_1|^2 - b(s) |s_1|^2 + \sum_{(i,j) \in n(s)} (E[b,s](i,j) |i|^2 - b(i,j)|i|^2) \\
+&= -d |s_1|^2 + \sum_{(i,j) \in n(s)} |i|^2
+\end{align*}
+We note that exactly $d-1$ of the neighbors of $s$ are a distance $|s_1 +1|$ from the origin and one of the neighbors of $s$ is a distance $(|s_1| - 1)$ from the origin.
 A point to emphasize is that this is the only substantial way in which this proof differs from the similar proof for the square lattice.
-
-We note that the contributions of these neighboring nodes increase by $||s_1| +1|$ and $||s_1|-1|$ respectively.
-
-Omitting the algebraic details we find that 
+Our expression simplifies to
 \begin{align*}
-\Psi( E_s b) - \Psi(b) > d >0
+&= -d |s_1|^2 + \sum_{(i,j) \in n(s)}-|i|^2 \\
+&= - d|s_1|^2 + (d-1)(|s_1| +1)^2 +(|s_1| -1)^2 \\
+&= d + (d-2)2|s_1| >0
 \end{align*}
-which completes the proof.
-\end{proof}
-
-\begin{corollary}
-\label{notreerepeats}
-A finite $d$-board can never return to a previous state via a sequence of earthquakes
-\end{corollary}
-\begin{proof}
-The proof is identical to that of Corollary ~\ref{norepeats}
 \end{proof}
@@
 Suppose $\sum_{x,y} f(x,y) = C$, and the first coordinate of all nodes which are nonzero are in $[i_{min}, i_{max}]$.
-Then after applying any number of earthquakes the first coordinates of the nonzero nodes of the resultant board will be in range $[i_{min} - 2C, i_{max} + 2C]$.
+Then $E^r[b, \{s_k\}]$ only contain chips within the range $[i_{min} - 2C, i_{max} + 2C]$ for all $r>0$ and all sequences $\{s_k\}$ of length $r$.
 \end{lemma}
@@
 But this cannot occur because in order to move a chip to a node with a lower first coordinate, one chip must moved to a node with a higher first coordinate.
-Therefore the resultant board does not have a node with a nonzero number of chips on it with first coordinate less than $i_{min} -2C$.
+Therefore $E^r[b, \{s_k\}]$ does not have a node with a nonzero number of chips on it with first coordinate less than $i_{min} -2C$.
 
-An analogous argument shows that  and after applying any sequences of earthquakes to $b$ the resultant board does not have a node with a nonzero number of chips on it with first coordinate greater than $i_{max}+2C$.
+An analogous argument shows that if $i_{max}$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not have a node with a nonzero number of chips on it with first coordinate greater than $i_{max}+2C$.
 \end{proof}
 
-\begin{theorem} If $b$ is a finite $d$-board then there exists some sequence $\{s_1, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $b$ the resultant $d$-board is stable.
-\end{theorem}
+So now we are ready to prove the theorem that all finite boards will reach a stable state.
+
 \begin{proof}
-The proof is identical to that of Theorem ~\ref{finitestability} with Lemma ~\ref{finiteextension} replaced by Lemma ~\ref{finiteextensiontree} and Corollary ~\ref{norepeats} replaced by ~\ref{notreerepeats}.
+(of Theorem~\ref{finitestability}) Suppose $b \in \mathcal{B}(d)_f$ is a finite board, so $\sum_{(i,j) \in G_n} b(i,j) = C$, and suppose all chips lie between levels  $[i_{min}, i_{max}]$.
+Then by Lemma ~\ref{finiteextensiontree}, we have that for all $r$ and sequence $\{s_k\}$ all chips in $E^r[b, \{s_k\}]$ will have first coordinate in $ [i_{min} -2C, i_{max} +2C]$.
+
+We have restricted the size of the board to be finite, and the number of chips that can be placed on the board to be finite. 
+Thus the total number of ways that the chips can be arranged on the board is finite. 
+In particular there are fewer than $(i_{max} - i_{min}+1 + 4 C)$  and there are at most 
+\begin{equation}
+d^l,
+\end{equation}
+nodes at each level, where $l= \max(|i_{min}|, |i_{max}|)$. So there are at most 
+\begin{equation}
+N:=(i_{max} - i_{min}+1 + 4 C)d^l
+\end{equation}
+nodes that are potentially occupied. Therefore there are at most 
+\begin{equation}
+\binom{N+S-1}{N-1}
+\end{equation}
+possible boards $b \in \mathcal{B}(n)_f$ that satisfy this.
+We know that there are no cycles in the states, so there must be an $0 \leq N < \binom{N+S-1}{N-1}$ and some sequence $\{s_k\}$ such that $E^N[b, \{s_k\}] = E^{N+m}[b, \{s_k\}]$ for all $m \geq 0$.
 \end{proof}
@@
 
-In this section, we show how we can compute introduce the ideas of Boolean functions and Boolean circuits and show how we can compute these use the poker chip board and the earthquake operator.
+In this section, we show how we can compute different Boolean functions using the poker chip board and the earthquake operator.
 
-We first introduce conventional definitions of boolean variables and functions.
-\begin{definition} 
-A Boolean variable is a number in $\{0,1\}$, and a Boolean function is a function that takes in a finite number of Boolean variable and outputs a Boolean variable.
-\end{definition}
+A Boolean input is a number in $\{0,1\}$, and a Boolean function is a function that takes $k$ Boolean inputs and outputs a number in $\{0,1\}$.
+A simple Boolean function is the AND function that takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are set to 1.
+Another Boolean function is the OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
+If 0 is associated with ``false" and 1 is associated with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
+More complicated Boolean functions can be made with more than 2 inputs and compositions of these inputs.
+Denoting an AND with $\vee$ an OR with $\wedge$, we can construct composite functions such as $b_1 \wedge (b_2 \vee b_3)$.
 
-We now introduce the standard Boolean functions. 
-\begin{definition}
- The AND function takes in 2 Boolean inputs, and returns a 1 if and only if both the input values are 1.
-We will use $\vee$ to denote AND.
-The OR function, that takes in 2 Boolean inputs, and returns a 0 if and only if both input values are set to 0.
-We will use $\wedge$ to denote OR.
-The NOT function takes in 1 Boolean input; it returns a 1 if and only if the input value is a 0.
- \end{definition}
- If we think of associating 0 with ``false" and 1 with '``true", then the AND function returns ``true" if both inputs are ``true", and the OR function returns ``true" if at least one input is true.
-We can make more complicated Boolean functions with more than 2 inputs and compositions of these inputs.
+Boolean circuits are formed by using wires that transmit 0's and 1's, and feeding them through various \emph{gates}.
+For example, a wire carrying a 0 and a wire carrying a 1 can be fed through an OR gate, which will ``output" a wire carrying a 1.
+In this section, we construct circuits using the poker chips board that use AND an OR gates.
 
- We now begin to define Boolean circuits, first defining the two component parts, wires and gates.
-\begin{definition}
- A wire is a device that in theory is capable of transmitting Boolean variables over arbitrarily long distances.
-If it takes in a 1 it will output a 1 and if it takes in a 0 it will output a 0.
-A gate is a boolean function that takes in an arbitrary number of wires and outputs an arbitrary number of wires.
-\end{definition}
-Examples of gates include AND gates and OR gates. In this section we will show that we can use the poker chips board to construct circuits that use AND and OR gates.
+\subsection{Simple Model}
 
-We now define a simple model of computation.
-We assume a finite board and specify some nodes as inputs nodes and some as outputs.
-If an input bit is to be set to 1, then we place 4 chips on the input node.
-Otherwise we do not place any chips on it.
-We then apply global earthquakes to the board until it stabilizes.
-If an output node ever has chips redistributed from it by an earthquake, then we set the output bit to 1; otherwise, the output bit is set to 0.
+A simple model of computation can be defined as follows.
+We assume a finite board.
+We specify some input nodes and some output nodes.
+If an input bit is to be set to 1, then we place 4 chips in the input node.
+We apply earthquakes to the board until it stabilizes.
+If an output node ever had an earthquake, then we mark that output bit as 1; otherwise, the output bit is set to 0.
 
-A simple example of this is a board that computes the identity function of one input. We can compute 
-\begin{equation*}
-f(x) = x
-\end{equation*}
-Our board has no chips on it, other than perhaps at the input node.
-We set the output node to be the same as the input node.
-If $x = 0$, then the input node does not have any chips on it and the board is stable.
-An earthquake never occurred on the output node and so the output is 0.
-If $x = 1$, then we place 4 chips on the input node and in order for the board to stabilize an earthquake must effect it.
-Because the input and output node are the same, this means an earthquake occurred on the output node and so the output is 1.
-
-Using this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
+A simple example is a board that computes the identity function of one output. We can compute 
+\[ f(x) = x \]
+By having a board with no chips, and we let the input node be the same as the output node.
+Then if $x = 0$, this means that the input node will not contain any chips, so the board is stable.
+The output node never received an earthquake, so the output is 0.
+If $x = 1$, then the input node gets 4 chips and an earthquake affects it.
+So the output node is activated, and so the output is 1.
+With this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
 
@@
 
-Composing these gates, we can make circuits which can be used to compute certain Boolean functions.
-In particular we can compute any Boolean function that can be represented by a planar circuit composed of only AND and OR gates.
-However using this model we cannot construct a NOT gate.
+Composing these gates, we can make circuits which represent some Boolean function. In particular we can compute any Boolean function that can be represented with a planar circuit with only AND and OR.
 
@@
 \begin{proof}
-Suppose for contradiction that we have a board with one input $i$ and one output $o$ which computes NOT.
-Then we know that there exists some node which initially has over $4$ chips otherwise the board would be stable and the output could never be effected by an earthquake.
+Suppose for contradiction that we have a board $b$ with one input $i$ and one output $o$ which computes NOT.
+Then we know that there exists some node which initially has over $4$ chips, since otherwise, the board would be stable and so the output never gets activated.
 
-Let $A(0)$ be the set of nodes that are targeted by earthquakes when $i = 0$ and $A(1)$ be the set of nodes that are targeted by earthquakes when $i=1$.
-We claim that $A(0) \subset A(1)$.
-This is true as when we set the input square to 1 we only increase the locations that can chips redistributed from them by earthquakes.
-However the output square is not in $A(1)$ by definition of our function, which produces a contradiction.
-Therefore our current model of computation cannot have a board that computes NOT.
+Let $A(0)$ be the set of nodes that get activated when $i = 0$ and $A(1)$ be the set of nodes that get activated when $i=1$.
+Then we claim that $A(0) \subset A(1)$.
+This would yield a contradiction since $o \in A(0)$ but $o \notin A(1)$.
 \end{proof}
 
-%\begin{lemma}
-%Let $A(i)$ be the set of active nodes in an execution of board $b$ with input node $i$. Then $A(0) \subset A(1)$.
-%\end{lemma}
+\begin{lemma}
+Let $A(i)$ be the set of active nodes in an execution of board $b$ with input node $i$. Then $A(0) \subset A(1)$.
+\end{lemma}
 
-%\begin{proof}
-%By Theorem~\ref{thm:order}, we can simulate an execution of the board with input be 0 before we apply earthquakes to the input node.
+\begin{proof}
+By Theorem~\ref{thm:order}, we can simulate an execution of the board with input be 0 before we apply earthquakes to the input node.
 We start with a board $b$ where the input is 1, so it has 4 chips.
@@
 In addition, the input node does not contain more than $7$ chips for more than one consecutive earthquake.
-%We do this until every node except the input node has less than 4 chips.
+We do this until every node except the input node has less than 4 chips.
 
-%Note that this execution is indistinguishable from the execution with input set to 0.
+Note that this execution is indistinguishable from the execution with input set to 0.
 This means that all nodes activated in the execution with input 0 are the same as the nodes activated in the above mentioned execution.
 
-%Once this is done, then we continue with the execution applying the earthquake to the input node.
+Once this is done, then we continue with the execution applying the earthquake to the input node.
 By Theorem~\ref{thm:order}, the final configuration will be the same as any execution of the board.
 So $A(0) \subset A(1)$.
-%\end{proof}
+\end{proof}
 
-We now introduce a slightly more complicated model for computation that will allow us to build NOT gates as well as AND and OR gates.
+\subsection{More powerful model}
+
 As before, we assume a finite board with finitely many chips.
-We again specify certain nodes as  inputs and outputs, placing 4 chips on input bits that are to be set to 1.
-We also designate a specific node $r$, it can be any node including inputs and outputs.
-When $r$ is effected by an earthquake, we finish redistributing all of the chips and read the outputs, setting an output to be 1 if it has 4 chips on it and 0 otherwise.
+The inputs are designated by having one input node.
+The outputs also each have one node.
+We now make the model of computation stronger by adding a special node $r$.
+When $r$ gets activated, it ends the execution and reads the outputs.
+We assume that in one ``round" we apply an earthquake to each valid node and that $r$ is the last node which gets applied an earthquake.
 
-We show how to make a board that computes the identity, as well as AND and OR gates. They are essentially the same as for our simpler model of computation but they include now include an $r$ node.
+In order to make it easier to compute, we will make a special \textit{clock} node, designated as $c$ which always starts out with $4$ chips.
+We will have the clock go through each input and travel in parallel to each input.
+We will build \textit{gadgets} that allow us to compute with the clock so that we can build gates in a modular and then put them together.
 
+We will also make a wire path that takes $c$ to $r$, so that we know exactly at what point the outputs are read.
+
+As shown before, simple example is given by a board that computes the identity.
+For simplicity, we write the boards as a board of numbers, which every space is a node and two nodes have an edge between them if they are either above, below, left, or right.
+
+\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
+				     4_c & 3 & 3 & 3 & 3 & 3_r \\
+				     0 & 0 & 0 & 0 & 0 & 0 \\
+				     i  & 3 & 3 & 3 & 3 & 3_o \\
+				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
+Where $i$ indicates the input bit node, and $o$ indicates the output bit.
+
+One can see that $r$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $r$.
+Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active.
+Therefore, we can compute the identity.
+
+This circuit also shows how to make a ``wire" in the circuit. We will represent a wire as two parralel arrows, one containing the clock path and the other carrying the input.
+
+\subsection{Turner}
+
+Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
+this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $r$ is the output read and $o$ is the output bit.
+
+\[ \begin{array}{cccccc} 0 & 0 & r & 0 & o \\
+				     0 & 3 & 2 & 0 & 3 \\
+			            0 & 2 & 3 & 0 & 3 \\
+				    0 & 3 & 2 & 0 & 3 \\
+				    0 & 2 & 3 & 0 & 3 \\
+				    c & 3 & 3 & 0 & 3 \\
+				    0 & 0 & 0 & 0 & 3 \\
+				    i & 3 & 3 & 3 & 3 \end{array} \]	
+Since the clock snakes along the path while $i$ takes the shorter path. 
+Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
+
+\subsection{Synchronize}
+
+We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
+\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
+					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
+					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
+					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
+					 0 & 0 & 0 & 0 & o & 0 & r & 3 \end{array} \]
+Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
+$i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.
+
+The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
+
+\subsection{Boolean circuits}
+
+Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
+The following can compute the AND of two inputs.
+
+\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
+					   &    & 3   & 2 &  3    &    &  \\
+					   &    &    & 3 &    & 3 & c_2 \\
+					o & 3 & 3 & 2 & 3 & 2 & 	    \\
+					   &    &    &    &    & 3 & i_2 \end{array} \]
+Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
+So in this case, $o = i_1$ AND $i_2$.
+
+Similarly, we can compute the OR of two circuits with a very simple modification:
+\[ \begin{array}{cccccccc}    &    & c_1&   &  i_1 &     &  \\
+					   &    & 3   & 2 &  3    &    &  \\
+					   &    &    & 3 &    & 3 & c_2 \\
+					o & 3 & 3 & 3 & 3 & 2 & 	    \\
+					   &    &    &    &    & 3 & i_2 \end{array} \]
+And we can do a similar set up to synchronize the clocks.
+
+We can build a NOT board with the following circuit:
 \begin{figure}
 \centering
-\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wirewithr}\label{fig:wirer}}
-\qquad\qquad
-\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andwithr}\label{fig:ANDr}}
-\qquad\qquad
-\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orwithr}\label{fig:ORr}}
-\caption{Gates in the our more complicated model.
-Green are the inputs, red are the outputs and yellow is the node $r$.
-We disregard edges and assume that any two nodes that are touching have an edge between them.}
-\label{fig:simplegateswithr}
-\end{figure}
-
-%Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance.
-this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $r$ is the output read and $o$ is the output bit.
-
-%Since the clock snakes along the path while $i$ takes the shorter path. 
-%Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.
-
-%We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
-
-%Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long.
-$i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.
-
-%The fact that we have a synchronizer means that the model of computation is equivalent to one where there is only one clock for all input bits.
-
-%Now we can show how to compute some basic boolean functions and since we have turns and synchronizers, we can assume that we have synchronized the inputs already.
-The following can compute the AND of two inputs.
-
-%Where we can make the $c_1$ and $c_2$ snake around before hand to guarantee that they are synchronized, and then we can make $c_1$ snake around so that it becomes the clock of the output.
-So in this case, $o = i_1$ AND $i_2$.
-
-%Similarly, we can compute the OR of two circuits with a very simple modification:
-%And we can do a similar set up to synchronize the clocks.
-
-Our goal in introducing this more complicated model of computation is to allow ourselves to build a board that would compute NOT. The following board does so:
-\begin{figure}
-\centering
-\includegraphics[width=0.5\linewidth]{notgate}
-\caption{Board which computes $f(x) = \neg x$}
+\includegraphics[width=0.25\linewidth]{notgate.pdf}
+\caption{Possible board to compute $f(x) = \neg x$}
 \label{fig:NOT}
 \end{figure}
-We note that in Figure~\ref{fig:NOT}, the number of nodes between the 4 and $r$ is the as the wire from the 4 and the output node, but the number of nodes separating the input node and $r$ is shorter than the number of nodes separating the 4 and the output node.
-If $i$ is set to $1$, then $r$ will be activated while the output only has 3 chips on it, so the output will be read as 0.
-If the input is set to 0, then $r$ and  the square adjacent to $o$ will both have 4 chips on them at same time so the output will be $1$.
-This computes $f(x) = \neg x$.
-however, we have switched the standard positions of $r$ and the output node, so it is unclear how to compose this board with other ones.
+In Figure~\ref{fig:NOT}, the wire to $c$ and $r$ is the same length as the wire from $c$ to $o$, but the wire from $i$ to $r$ is shorter than the length of the wire from $c$ to $o$.
+If $i$ is set to $1$, then $r$ will be activated before the clock reaches $o$, so $o$ will be read as 0.
+If $i$ is set to 0, then the clock will reach $o$ and $r$ at the same time, so the output is $1$.
+This computes $f(x) = \neg x$; however, we have switched the order of $r$ and $o$.
+We also do not know how to compose this board with other boards.
+
+We would also like to build a crossover gadget to allow us to compute circuits which are not necesarily.
+
 \end{document}

### 2014-10-13 finaldraft.tex
fixed typos
fixed more typos

--- 2014-10-13 finaldraft.tex
@@
 We then use the fact that earthquakes strictly increase the value of this spread function to show that finite boards cannot cycle back to previous configurations.
-We also show that the number of possible configurations a finite board can be transformed into by earthquake is bounded.
+We also show that the number of possible configurations a finite board can be transformed into by earthquakes is bounded.
 These two results together will show that all finite boards reach a steady state.
-In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by a earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
+In Section~\ref{Order of Redistribution}, we show that for a given board the number of times a particular square must be hit by an earthquake in order for a board to stabilize is independent of the order in which the earthquakes occur.
 
@@
 \begin{proof}
-Lemma~\ref{finiteextension} implies that there a finite board with $C$ total chips can only ever occupy a finite number $D$ of squares.
+Lemma~\ref{finiteextension} implies that a finite board with $C$ total chips can only ever occupy a finite number $D$ of squares.
 This implies a finite upper bound of $\binom{D+C-1}{D-1}$ possible boards (via an elementary combinatorics argument of how to distribute $C$ identical objects among $D$ distinguishable slots).
@@
 
-We first introduce conventional definitions of boolean variables and functions.
+We first introduce conventional definitions of Boolean variables and functions.
 \begin{definition} 

### 2014-10-13 finaldraft.tex
fixed some typos

@@
 We again specify certain nodes as  inputs and outputs, placing 4 chips on input bits that are to be set to 1.
-We also designate a specific node $r$, it can be any node including inputs and outputs.
-When $r$ is effected by an earthquake, we finish redistributing all of the chips and read the outputs, setting an output to be 1 if it has 4 chips on it and 0 otherwise.
+We also designate a specific node $r$ that can be any node including inputs and outputs.
+When $r$ is affected by an earthquake, we finish redistributing all of the chips and read the outputs, setting an output to be 1 if it has 4 chips on it and 0 otherwise.
 
-We show how to make a board that computes the identity, as well as AND and OR gates. They are essentially the same as for our simpler model of computation but they include now include an $r$ node.
+We show how to make a board that computes the identity, as well as AND and OR gates. They are essentially the same as in our simpler model of computation but they include now include an $r$ node.
 
@@
 
-Our goal in introducing this more complicated model of computation is to allow ourselves to build a board that would compute NOT. The following board does so:
+Our goal in introducing this more complicated model of computation is to allow ourselves to build a board that can compute NOT. The following board does so:
 \begin{figure}
@@
 \end{figure}
-We note that in Figure~\ref{fig:NOT}, the number of nodes between the 4 and $r$ is the as the wire from the 4 and the output node, but the number of nodes separating the input node and $r$ is shorter than the number of nodes separating the 4 and the output node.
+We note that in Figure~\ref{fig:NOT}, the number of nodes between the 4 and $r$ is same as the number of nodes between the 4 and the output node, but the number of nodes separating the input node and $r$ is shorter than the number of nodes separating the 4 and the output node.
 If $i$ is set to $1$, then $r$ will be activated while the output only has 3 chips on it, so the output will be read as 0.
@@
 This computes $f(x) = \neg x$.
-however, we have switched the standard positions of $r$ and the output node, so it is unclear how to compose this board with other ones.
+However, we have switched the standard positions of $r$ and the output node, so it is unclear how to compose this board with other ones.
 \end{document}

### 2014-10-13 finaldraft.tex
adding poem

--- 2014-10-13 finaldraft.tex
@@
 however, we have switched the standard positions of $r$ and the output node, so it is unclear how to compose this board with other ones.
+
+\section{Poem}
+
+\begin{quote}
+A board with chips.\\
+An earthquake redistibutes some chips.\\
+New board! never turn back. \\ \\
+
+Another earthquake hits.\\
+And another earthquakes hits again.\\
+Who cares which first! \\ \\
+
+One last earthquake\\
+So many it must be Japan.\\
+Computation.
+\end{quote}
+
 \end{document}

### 2014-10-13 finaldraft.tex

--- 2014-10-13 finaldraft.tex
@@
 \end{figure}
+
 We note that in Figure~\ref{fig:NOT}, the number of nodes between the 4 and $r$ is same as the number of nodes between the 4 and the output node, but the number of nodes separating the input node and $r$ is shorter than the number of nodes separating the 4 and the output node.
@@
 However, we have switched the standard positions of $r$ and the output node, so it is unclear how to compose this board with other ones.
+
+\section{Poem}
+
+\begin{quote}
+A board with chips.\\
+An earthquake redistibutes some chips.\\
+New board! never turn back. \\ \\
+
+Another earthquake hits.\\
+And another earthquakes hits again.\\
+Who cares which first! \\ \\
+
+One last earthquake\\
+So many it must be Japan.\\
+Computation.
+\end{quote}
 \end{document}

### 2014-10-13 finaldraft.tex
typos

--- 2014-10-13 finaldraft.tex
@@
 We claim that $A(0) \subset A(1)$.
-This is true as when we set the input square to 1 we only increase the locations that can chips redistributed from them by earthquakes.
-However the output square is not in $A(1)$ by definition of our function, which produces a contradiction.
+This is true as when we set the input square to 1 we only increase the locations that can have chips redistributed from them by earthquakes.
+However the output square is not in $A(1)$ by the definition of our function, which produces a contradiction.
 Therefore our current model of computation cannot have a board that computes NOT.

### 2014-10-13 finaldraft.tex
added haikus

@@
 
-\section{Poem}
+\section{Three Haikus}
 
-\begin{quote}
-A board with chips.\\
-An earthquake redistibutes some chips.\\
-New board! never turn back. \\ \\
+Poker board with chips\\
+Suddenly hit by earthquakes.\\
+Can never return. \\
 
-Another earthquake hits.\\
-And another earthquakes hits again.\\
-Who cares which first! \\ \\
+\noindent Another earthquake\\
+hits the board, and then one more\\
+Who cares which was first! \\
 
-One last earthquake\\
-So many it must be Japan.\\
-Computation.
-\end{quote}
+\noindent Now the aftershock\\
+is computing some functions,\\
+but just basic ones.
 \end{document}

### 2014-10-13 finaldraft.tex
fixed more typos

@@
 In fact we are able to show that not only is this final board the same but that the number of earthquakes that target a particular node in order to stabilize the board is invariant of the order in which the earthquakes occur.
-We show this by first describing conditions under which we can rearrange a sequence earthquakes without changing the resultant board.
+We show this by first describing conditions under which we can rearrange a sequence of earthquakes without changing the resultant board.
 We then define a process that rearranges a sequence of earthquakes without changing the final board state and for a given board will always produce the same sequence.\\
@@
 We say that an earthquake $E$ is valid if the node it targets has at least $4$ chips on it. 
-We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ to be a valid sequence of earthquakes if each earthquake is valid when applied one after another. 
+We say that a sequence of earthquakes $E_1, E_2, \ldots E_n$ is a valid sequence of earthquakes if each earthquake is valid when applied one after another. 
 \end{definition}
@@
 But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes.
-This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.
+This contradicts one of our hypotheses, and so $E_2, E_1$ is a valid sequence.
 

### 2014-10-13 finaldraft.tex
added commas and periods after equations
typo fixes

@@
 \begin{equation*}
-\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 
+\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2.
 \end{equation*}
@@
 \begin{equation*}
-\Psi(E_sb) > \Psi(b) 
+\Psi(E_sb) > \Psi(b),
 \end{equation*}
@@
 \begin{equation*}
- \Psi(E_sb) - \Psi(b) = 4
+ \Psi(E_sb) - \Psi(b) = 4,
  \end{equation*}
@@
 \begin{equation*}
-[[A]] = \begin{cases} 1 \text{ if A is true} \\ 0 \text{ if A is false} \end{cases}
+[[A]] = \begin{cases} 1 \text{ if A is true} \\ 0 \text{ if A is false} \end{cases}.
 \end{equation*}
@@
 \begin{equation*}
-E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
+E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]].
 \end{equation*}
@@
 \begin{equation*}
-E_2 E_1 b = E_1 E_2 b
+E_2E_1 b = E_1E_2 b.
 \end{equation*}
@@
 \begin{equation*}
-E_m, E_{m+1}, \ldots, E_{j-1}, E_j
+E_m, E_{m+1}, \ldots, E_{j-1}, E_j,
 \end{equation*}
@@
 We can do this for any other sequence that stabilizes the board and get the same sequence.
-This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them).
+This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes; it never adds or removes them).
 \end{proof}
@@
 \begin{equation*}
-G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
+G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}.
 \end{equation*}
@@
 \begin{equation*}
-b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
+b:  G_d \rightarrow \mathbb{Z}_{\geq 0}.
 \end{equation*}
@@
 \begin{equation*}
-\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
+\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2.
 \end{equation*}
@@
 \begin{equation*}
-\Psi( E_s b) > \Psi (b)
+\Psi( E_s b) > \Psi (b).
 \end{equation*}
@@
 \begin{align*}
-\Psi( E_s b) - \Psi(b) > d >0
+\Psi( E_s b) - \Psi(b) > d >0,
 \end{align*}
@@
 
-In this section, we show how we can compute introduce the ideas of Boolean functions and Boolean circuits and show how we can compute these use the poker chip board and the earthquake operator.
+In this section, we show how we can compute Boolean functions using the poker chip problem setup.
+In particular, we use the fact that Boolean functions can be represented as Boolean circuits, and we show how to construct elements of Boolean circuits using the poker chip board and earthquake operator.
 
@@
 We then apply global earthquakes to the board until it stabilizes.
-If an output node ever has chips redistributed from it by an earthquake, then we set the output bit to 1; otherwise, the output bit is set to 0.
+If an output node ever has chips redistributed to it by an earthquake, then we set the output bit to 1; otherwise, the output bit is set to 0.
 
@@
 \begin{equation*}
-f(x) = x
+f(x) = x.
 \end{equation*}
@@
 
-Using this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
+Using this simple model, there is a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).
 
@@
 \begin{theorem}
-There does not exists a board which computes $f(x) = \neg x$.
+There does not exist a board which computes $f(x) = \neg x$.
 \end{theorem}
@@
 We claim that $A(0) \subset A(1)$.
-This is true as when we set the input square to 1 we only increase the locations that can have chips redistributed from them by earthquakes.
+This is true as when we set the input square to 1, we only increase the locations that can have chips redistributed to them by earthquakes.
 However the output square is not in $A(1)$ by the definition of our function, which produces a contradiction.
@@
 \noindent Another earthquake\\
-hits the board, and then one more\\
+hits the board, and then one more.\\
 Who cares which was first! \\
@@
 \noindent Now the aftershock\\
-is computing some functions,\\
-but just basic ones.
+is computing some functions\\
+---but just basic ones!
 \end{document}
